---
title: "Moorea Coral Reef Productivity Analysis: Long-term Ecosystem Dynamics"
author: "Nyssa Silbiger"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  cache: true
---


# Introduction

This report analyzes long-term ecosystem dynamics at LTER Site 1 in Moorea, French Polynesia. We examine relationships between benthic community composition, environmental conditions, fish assemblages, and ecosystem metabolism over nearly two decades (2005-2025). The analysis integrates multiple datasets to understand how coral reef ecosystems are responding to environmental change.

## Setup

```{r setup}
#| label: setup
#| cache: false

# Load required libraries
# Data manipulation
library(tidyverse)
library(here)
library(lubridate)
library(zoo)
library(slider)
library(glue)

# Bayesian modelling
library(brms)
library(tidybayes)
library(bayesplot)

# Visualisation
library(scales)
library(patchwork)
library(cowplot)
library(viridis)
library(ggridges)
library(ggsci)
library(ggraph)
library(igraph)
library(showtext)
library(sysfonts)

# Statistics / tables
library(psych)
library(ggcorrplot2)
library(kableExtra)
library(httr2)

# Enable custom fonts
showtext_auto(enable = TRUE)

# Set global theme
theme_set(theme_bw())

# Set seed for reproducibility
set.seed(223)
```

# Data

## 1. Temperature Data

Extract yearly average and maximum daily temperature values from the backreef temperature time series.

```{r temperature-processing}
#| label: process-temperature

# Read and process temperature data
Temp_LTER1 <- read_csv(here("Data", "raw_data", "MCR_LTER02_BTM_Backreef_Forereef_20251114.csv")) |>
  filter(reef_type_code == "Backreef")

# Calculate daily means, then yearly statistics (optimized with single pass)
Mean_year_Temp <- Temp_LTER1 |>
  mutate(date = as.Date(time_local)) |>
  group_by(date) |>
  summarise(daily_temp = mean(temperature_c, na.rm = TRUE), .groups = "drop") |>
  mutate(Year = year(date)) |>
  group_by(Year) |>
  summarise(
    Mean_temp = mean(daily_temp, na.rm = TRUE),
    Max_temp = max(daily_temp, na.rm = TRUE),
    .groups = "drop"
  )

# Save processed data
write_csv(Mean_year_Temp, here("Data", "InSituTemp.csv"))

# Quick summary
summary(Mean_year_Temp)
```

## 2. Nutrient Data

Process macroalgae tissue nitrogen content and water column nutrients, using tissue N to impute missing water column data.

```{r nutrient-processing}
#| label: process-nutrients

# Turbinaria %N data (tissue nitrogen as nutrient proxy)
Turb <- read_csv(here("Data", "raw_data", "MCR_LTER_Macroalgal_CHN_2005_to_2024_20250616.csv")) |>
  filter(
    Habitat == "Backreef",
    Genus == "Turbinaria",
    Site == "LTER_1",
    !is.na(N)
  ) |>
  group_by(Year) |>
  summarise(
    N_percent = mean(N, na.rm = TRUE),
    C_percent = mean(C, na.rm = TRUE),
    CN = mean(CN_ratio, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(Turb, here("Data", "Turb_mean.csv"))

# Water column nutrients (bimonthly surveys 2005-2018)
water <- read_csv(here("Data", "raw_data", "WaterColumnN.csv")) |>
  mutate(
    Date = mdy(Date),
    Year = year(Date)
  ) |>
  group_by(Year) |>
  summarise(across(Phosphate:Nitrite_and_Nitrate,  \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  mutate(Site = "LTER 1")

write_csv(water, here("Data", "Water_N_mean.csv"))

# Impute water N+N from tissue N using linear model
AllNutrients <- Turb |>
  full_join(water, by = "Year") |>
  arrange(Year)

# Model and predict
modN <- lm(Nitrite_and_Nitrate ~ N_percent, data = AllNutrients)
AllNutrients$NN_impute <- predict(modN, newdata = AllNutrients)

# Model summary
cat("\nNutrient imputation model (Nitrite+Nitrate ~ Tissue %N):\n")
summary(modN)
```

## 3. Benthic Cover Data

Calculate percent cover for functional groups from benthic transect surveys.

```{r benthic-processing}
#| label: process-benthic

# Read and classify benthic data
BenthicCover_Algae <- read_csv(here("Data", "raw_data", "MCR_LTER_Annual_Survey_Benthic_Cover_20251009.csv")) |>
  filter(Habitat == "Backreef")

# Create functional group classifications (vectorized)
Benthic_summary_Algae <- BenthicCover_Algae |>
  rename(name = Taxonomy_Substrate_Functional_Group) |>
  mutate(
    name = case_when(
      name %in% c(
        "Amansia rhodantha", "Turbinaria ornata", "Dictyota sp.", "Halimeda sp.",
        "Galaxaura sp.", "Liagora ceranoides", "Cyanophyta", "Halimeda minima",
        "Amphiroa fragilissima", "Caulerpa serrulata", "Corallimorpharia",
        "Dictyota friabilis", "Galaxaura rugosa", "Cladophoropsis membranacea",
        "Galaxaura filamentosa", "Halimeda discoidea", "Peyssonnelia inamoena",
        "Caulerpa racemosa", "Valonia ventricosa", "Actinotrichia fragilis",
        "Dictyota bartayresiana", "Microdictyon umbilicatum", "Halimeda distorta",
        "Halimeda incrassata", "Halimeda macroloba", "Dictyota implexa",
        "Gelidiella acerosa", "Dictyosphaeria cavernosa", "Valonia aegagropila",
        "Microdictyon okamurae", "Halimeda opuntia", "Dichotomaria obtusata",
        "Chlorodesmis fastigiata", "Phyllodictyon anastomosans", "Phormidium sp.",
        "Cladophoropsis luxurians", "Sargassum pacificum", "Chnoospora implexa",
        "Halimeda taenicola", "Boodlea kaeneana", "Padina boryana",
        "Coelothrix irregularis", "Gelidiella sp.", "Hydroclathrus clathratus",
        "Dictyota divaricata", "Hypnea spinella", "Dichotomaria marginata",
        "Sporolithon sp.", "Chaetomorpha antennina", "Asparagopsis taxiformis"
      ) ~ "Fleshy Macroalgae",
      name %in% c(
        "Algal Turf", "Damselfish Turf", "Coral Rubble",
        "Lobophora variegata", "Shell Debris", "Bare Space"
      ) ~ "Turf/Cyanobacteria",
      name %in% c(
        "Peyssonnelia bornetii", "Peyssonnelia sp.",
        "Sponge", "Tridacna sp.", "No data"
      ) ~ "Sand",
      TRUE ~ name
    )
  ) |>
  group_by(Year, Site, name) |>
  summarise(
    total_cover = sum(Percent_Cover, na.rm = TRUE),
    mean_cover = 100 * total_cover / 5000,
    .groups = "drop"
  ) |>
  mutate(Site = str_replace_all(Site, "_", " "))

write_csv(Benthic_summary_Algae, here("Data", "Benthic_summary_algae.csv"))

# Calculate total calcifiers and living cover
Total_Calc <- Benthic_summary_Algae |>
  filter(name %in% c("Coral", "Crustose Corallines")) |>
  group_by(Year, Site) |>
  reframe(total_Calc = sum(mean_cover))

TotalLiving <- Benthic_summary_Algae |>
  filter(name %in% c("Coral", "Crustose Corallines", "Fleshy Macroalgae")) |>
  group_by(Year, Site) |>
  summarise(
    mean_alive = sum(mean_cover),
    mean_fleshy = sum(mean_cover[name == "Fleshy Macroalgae"], na.rm = TRUE),
    mean_coral = sum(mean_cover[name == "Coral"], na.rm = TRUE),
    .groups = "drop"
  )
```

## 4. Fish Biomass Data

Process fish survey data by trophic groups.

```{r fish-processing}
#| label: process-fish

# Read and process fish data
fish <- read_csv(here("Data", "raw_data", "MCR_LTER_Annual_Fish_Survey_20250324.csv"))

# Process by trophic groups (optimized filtering)
fish_summary <- fish |>
  filter(
    Site == "LTER_1",
    Habitat == "Backreef",
    Biomass < 8000, # Remove outlier sharks
    Biomass >= 0 # Remove negative (missing) values
  ) |>
  mutate(
    trophic_new = case_when(
      Fine_Trophic %in% c("Brusher", "Browser", "Excavator", "Scraper") ~ "Herbivores",
      Fine_Trophic == "Corallivore" ~ "Corallivore",
      TRUE ~ "Other"
    )
  ) |>
  group_by(Year, trophic_new) |>
  summarise(
    fish_g_m2 = sum(Biomass, na.rm = TRUE) / 1200, # 4 transects × 300 m²
    .groups = "drop"
  )

write_csv(fish_summary, here("Data", "fish_summary.csv"))
```

## 5. Primary Production Processing

Process raw production data and extract Pmax and Rd using Bayesian photosynthesis-irradiance curves.

```{r pp-processing}
#| label: process-pp
#| cache: true

# Read all QC PP files efficiently
filedir <- here("Data", "raw_data", "QC_PP")
files <- dir(path = filedir, pattern = ".csv", full.names = TRUE)

All_PP_data <- files |>
  set_names() |>
  map_df(read_csv, .id = "filename", show_col_types = FALSE) |>
  mutate(DateTime = mdy_hm(DateTime))

# Unit conversion and data cleaning (optimized)
All_PP_data <- All_PP_data |>
  mutate(
    # Convert 2007-2014 from g O2/m2/h to mmol O2/m2/h
    across(
      c(UP_Oxy, DN_Oxy, PP),
      ~ if_else(DateTime < ymd_hms("2014-04-01 00:00:00"), (.x / 32) * 1000, .x)
    ),
    Date = as_date(DateTime),
    DielDateTime = DateTime + hours(12),
    DielDate = as_date(DielDateTime),
    Year = year(DateTime)
  ) |>
  filter(!Date %in% mdy(c("5/27/2011", "5/28/2011", "1/21/2014", "05/25/2024")))

# Keep only complete 24-hour datasets
complete_dates <- All_PP_data |>
  count(DielDate) |>
  filter(n == 24)

All_PP_data <- complete_dates |>
  left_join(All_PP_data, by = "DielDate")

# Calculate GP and environmental averages
Daily_R <- All_PP_data |>
  group_by(Year, Season, DielDate) |>
  summarise(R_average = mean(PP[PAR == 0], na.rm = TRUE), .groups = "drop")

All_PP_data <- All_PP_data |>
  left_join(Daily_R, by = c("Year", "Season", "DielDate")) |>
  mutate(
    GP = if_else(PAR == 0, NA_real_, PP - R_average),
    GP = if_else(GP < 0, NA_real_, GP),
    Temperature_mean = (UP_Temp + DN_Temp) / 2,
    Flow_mean = (UP_Velocity_mps + DN_Velocity_mps) / 2
  )

# Calculate yearly averages
Year_Averages_PP <- All_PP_data |>
  mutate(
    NP = PP,
    R = if_else(PP < 0, PP, NA_real_)
  ) |>
  group_by(Year) |>
  summarise(
    NP_mean = mean(NP, na.rm = TRUE),
    NP_SE = sd(NP, na.rm = TRUE) / sqrt(n()),
    NP_max = max(NP, na.rm = TRUE),
    GP_mean = mean(GP, na.rm = TRUE),
    GP_SE = sd(GP, na.rm = TRUE) / sqrt(n()),
    R_mean = mean(R, na.rm = TRUE),
    R_SE = sd(R, na.rm = TRUE) / sqrt(n()),
    Temperature_mean = mean(Temperature_mean, na.rm = TRUE),
    Flow_mean = mean(Flow_mean, na.rm = TRUE),
    PAR_mean = mean(PAR[PAR > 0], na.rm = TRUE),
    .groups = "drop"
  )
```

## Sampling Coverage Summary

```{r}
#| label: sampling-coverage

# Calculate sampling metrics
metabolism_coverage <- All_PP_data |>
  group_by(Year) |>
  summarise(
    n_measurements = n() / 24,  # Convert hourly to daily measurements
    .groups = "drop"
  )

benthic_coverage <- Benthic_summary_Algae |>
  filter(Site == "LTER 1") |>
  group_by(Year) |>
  summarise(surveys = n_distinct(Year), .groups = "drop") |>
  nrow()

fish_coverage <- fish_summary |>
  group_by(Year) |>
  summarise(surveys = n_distinct(Year), .groups = "drop") |>
  nrow()

cat("=== Data Coverage Summary ===\n\n")
cat(sprintf("Ecosystem Metabolism:\n"))
cat(sprintf("  Total 24-hour cycles: %.0f over 20 years\n", sum(metabolism_coverage$n_measurements)))
cat(sprintf("  Annual coverage: %.1f ± %.1f measurements per year (range: %.0f-%.0f)\n",
            mean(metabolism_coverage$n_measurements),
            sd(metabolism_coverage$n_measurements),
            min(metabolism_coverage$n_measurements),
            max(metabolism_coverage$n_measurements)))
cat(sprintf("  Temporal span: %d years (%d-%d)\n\n", 
            length(unique(All_PP_data$Year)),
            min(All_PP_data$Year),
            max(All_PP_data$Year)))

cat(sprintf("Benthic Surveys:\n"))
cat(sprintf("  Total surveys: %d (annual coverage)\n", benthic_coverage))
cat(sprintf("  Years with data: %d of %d possible (100%%)\n\n", benthic_coverage, 21))

cat(sprintf("Fish Surveys:\n"))
cat(sprintf("  Total surveys: %d (annual coverage)\n", fish_coverage))
cat(sprintf("  Years with data: %d of %d possible (100%%)\n\n", fish_coverage, 21))

cat(sprintf("Temperature:\n"))
cat(sprintf("  Continuous 15-minute measurements: %d-%d (100%% coverage)\n\n", 
            min(lubridate::year(Temp_LTER1$time_local), na.rm = TRUE),
            max(lubridate::year(Temp_LTER1$time_local), na.rm = TRUE)))

cat(sprintf("Nutrients:\n"))
cat(sprintf("  Water column (direct): 2005-2018\n"))
cat(sprintf("  Tissue nitrogen: 2005-2024 (2019-2025 water column imputed)\n"))
```

## Bayesian Photosynthesis-Irradiance (PI) Curve Analysis

We fit a Bayesian non-linear model to extract photosynthetic parameters (alpha, Pmax, and Rd) for each year. The model uses a rectangular hyperbola:

$$PP = \frac{\alpha \cdot P_{max} \cdot PAR}{\alpha \cdot PAR + P_{max}} + R_d$$

where: - $\alpha$ = initial slope (photosynthetic efficiency) - $P_{max}$ = maximum photosynthetic capacity - $R_d$ = dark respiration (ecosystem respiration) - $PAR$ = photosynthetically active radiation

```{r bayesian-pi-curve}
# Bayesian PI curves to extract light-independent Pmax and Rd
LTER1_Pnet <- All_PP_data |>
  mutate(
    tempc = Temperature_mean,
    Flowmean = Flow_mean,
    flow_log = log(Flowmean),
    UPDN = as.factor(UPDN),
    Year = as.factor(Year)
  )

# Fit Bayesian nonlinear model
fit1_f <- brm(
  bf(
    PP ~ ((alpha * Pmax * PAR) / (alpha * PAR + Pmax)) + Rd,
    nl = TRUE,
    alpha ~ 0 + Year,
    Pmax ~ 0 + Year,
    Rd ~ 0 + Year
  ) + student(),
  data = LTER1_Pnet,
  set_rescor(FALSE),
  prior = c(
    prior(normal(0.1, 10), nlpar = "alpha", lb = 0),
    prior(normal(-100, 10), nlpar = "Rd", ub = 0),
    prior(normal(100, 10), nlpar = "Pmax", lb = 0)
  ),
  control = list(adapt_delta = 0.95, max_treedepth = 20),
  cores = 3,
  chains = 3,
  seed = 223,
  iter = 8000,
  warmup = 2000
)

# Extract Pmax and Rd coefficients
names <- rownames(fixef(fit1_f))
coefficients <- as_tibble(fixef(fit1_f)) |>
  mutate(params = names)

PRValues <- coefficients |>
  separate(params, sep = "_", into = c("param_type", "yearname"), remove = FALSE) |>
  filter(param_type %in% c("Pmax", "Rd")) |>
  separate(yearname, sep = "r", into = c("y", "year")) |>
  mutate(year = as.numeric(year)) |>
  select(Year = year, Estimate, param_type) |>
  pivot_wider(names_from = param_type, values_from = Estimate)

# Combine with raw averages
Year_Averages_PP <- Year_Averages_PP |>
  full_join(PRValues, by = "Year") |>
  mutate(Rd = -Rd) # Make Rd positive

# Calculate Bayesian R²
pi_curve_r2 <- median(bayes_R2(fit1_f)[, 1])
cat(sprintf("\nPhotosynthesis-Irradiance Curve Model:\n"))
cat(sprintf("  Bayesian R²: %.3f\n", pi_curve_r2))
cat(sprintf("  This indicates the model explains %.1f%% of variance in net production\n", pi_curve_r2 * 100))

write_csv(Year_Averages_PP, here("Data", "Year_Averages_PP.csv"))
```

## 6. Integrate All Datasets

```{r integrate-data}
#| label: integrate-datasets

# Read processed data files
Year_Averages_PP <- read_csv(here("Data", "Year_Averages_PP.csv"))
Turb <- read_csv(here("Data", "Turb_mean.csv"))
water <- read_csv(here("Data", "Water_N_mean.csv"))
insitu_temp <- read_csv(here("Data", "InSituTemp.csv"))

fish_trophic <- read_csv(here("Data", "fish_summary.csv")) |>
  pivot_wider(names_from = trophic_new, values_from = fish_g_m2) |>
  select(-Other)

Benthic_summary_Algae <- read_csv(here("Data", "Benthic_summary_algae.csv"))

# Create integrated dataset
Year_Averages <- Year_Averages_PP |>
  full_join(TotalLiving |> filter(Site == "LTER 1"), by = "Year") |>
  full_join(water) |>
  full_join(Turb, by = "Year") |>
  full_join(insitu_temp, by = "Year") |>
  full_join(fish_trophic, by = "Year") |>
  mutate(log_coral = log(mean_coral)) |>
  filter(Year < 2026) |>
  arrange(Year)

# Display structure
glimpse(Year_Averages)
```

# Methods

## Photosynthesis-Irradiance Curve Analysis

Annual values of maximum photosynthetic capacity (Pmax) and ecosystem respiration (Rd) were extracted from a Bayesian nonlinear model fit to 2,928 net O₂ flux measurements using the **brms** package (Bürkner 2017) and Stan (Carpenter et al. 2017). We used the rectangular hyperbola:

$$\mathrm{PP} = \frac{\alpha \cdot P_{\max} \cdot \mathrm{PAR}}{\alpha \cdot \mathrm{PAR} + P_{\max}} + R_d$$

where PP is net community production, PAR is photosynthetically active radiation, and α is photosynthetic efficiency. Year-specific parameters (α, Pmax, Rd) were estimated simultaneously from all deployments. A Student-*t* likelihood accommodated measurement outliers. Priors were weakly informative: α ~ Normal$_+$(0.1, 10); Pmax ~ Normal$_+$(100, 10); Rd ~ Normal$_-$(−100, 10). Three chains ran for 8,000 iterations (2,000 warmup) at adapt_delta = 0.95. Posterior Rd values were multiplied by −1 to express ecosystem respiration as positive rates.

## Temporal Trend Analysis

We estimated standardised temporal trends for 10 ecosystem variables (coral cover, fleshy macroalgae, corallivore biomass, herbivore biomass, algal nitrogen content, ecosystem respiration, Pmax, net ecosystem production, maximum temperature, nitrate+nitrite) using independent Bayesian linear regression models (Variable ~ Year, default **brms** priors) on z-scored data. The slope posterior (change per standardised year) was used to assess direction and magnitude of each trend. Effects were considered directional when the 95% credible interval excluded zero. Bayesian R² (Gelman et al. 2019) is reported as the posterior median. Absolute percent change was calculated between the first and last non-missing observations on the original scale.

## Correlation Analysis

Pearson correlations among all major ecosystem variables were computed and visualised as a mixed correlation matrix (upper triangle: ellipses scaled by correlation strength; lower triangle: numeric values; diagonal: variable names). Significance was assessed using two-tailed t-tests; variables were ordered by hierarchical clustering of the correlation matrix.

## Structural Equation Modeling

To evaluate simultaneous causal pathways among ecosystem components, we fit a Bayesian multivariate SEM in **brms** with residual correlations disabled (`set_rescor = FALSE`). The directed acyclic graph (DAG) was motivated by ecological theory and temporal precedence, comprising eight response equations:

| # | Equation |
|---|----------|
| 1 | Temperature ~ Year |
| 2 | log(coral cover) \| trunc(ub = 3.39) ~ temperature + yearresid |
| 3 | log(ecosystem respiration) ~ log(coral cover) + temperature |
| 4 | log(fleshy macroalgae) \| trunc(ub = 2.26) ~ log(coral cover) |
| 5 | log(Pmax) ~ log(coral cover) + temperature + log(ecosystem respiration) |
| 6 | corallivore biomass ~ log(coral cover) |
| 7 | herbivore biomass ~ log(fleshy macroalgae) |
| 8 | log(nitrogen content) ~ log(ecosystem respiration) |

All continuous responses were log-transformed and z-score standardised before modelling. The `yearresid` predictor is the residual of calendar year regressed on maximum annual temperature, capturing non-thermal secular coral decline. Cover responses were upper-truncated at the z-log scale equivalent of 100% cover to respect the biological ceiling; truncation values (3.39 for coral, 2.26 for fleshy algae) were verified to back-transform to exactly 100%.

Missing predictor values (ecosystem respiration and temperature as predictors of Pmax; ecosystem respiration as predictor of N content) were handled via **brms** integrated missing-data imputation (`mi()` syntax). Priors were weakly informative: regression coefficients ~ Normal(0, 2); intercepts ~ Normal(0, 2); residual SDs ~ Normal$_+$(0, 0.25) for the temperature sub-model and Normal$_+$(0, 2) otherwise. Three chains ran for 10,000 iterations (5,000 warmup) with adapt_delta = 0.99 and max_treedepth = 15. Convergence was assessed via Rhat (< 1.01 for all parameters), bulk and tail effective sample sizes (> 1,000), and posterior predictive checks.

### Parsimonious (reduced) SEM

A reduced model was fit by removing three pathways whose 95% credible intervals included zero in the full model: Temperature → Respiration (direct), Temperature → Pmax (direct), and Coral → Pmax (direct). All retained coefficients were stable between full and reduced models (within 0.02 SD). The reduced model is used for all ecosystem projections.

### Direct and indirect temperature effects

Total standardised effects of temperature on each ecosystem variable were decomposed into direct paths and indirect paths (products of coefficients along each chain through the DAG) using all 15,000 post-warmup draws. Indirect effects were computed as the product of relevant path coefficients; for example, the indirect effect on respiration was β(Temp→Coral) × β(Coral→Rd). Total effects are the sum across all contributing paths. An effect was considered credible if its 95% CI excluded zero.

## Ecosystem Projections

### CMIP6 climate scenario projections (to 2100)

To project ecosystem trajectories under realistic climate forcing, monthly sea surface temperature (*tos*, variable `sea_surface_temperature`) for the Moorea grid cell (~17.5°S, 149.8°W) was obtained from ten coupled CMIP6 ScenarioMIP models — ACCESS-CM2 (Bi et al. 2020), BCC-CSM2-MR (Wu et al. 2019), EC-Earth3 (Döscher et al. 2022), GFDL-ESM4 (Dunne et al. 2020), IPSL-CM6A-LR (Boucher et al. 2020), MIROC6 (Tatebe et al. 2019), MPI-ESM1-2-LR (Müller et al. 2018), MRI-ESM2-0 (Yukimoto et al. 2019), NorESM2-LM (Seland et al. 2020), and UKESM1-0-LL (Sellar et al. 2019) (Eyring et al. 2016) — downloaded from the Copernicus Climate Data Store (`projections-cmip6` dataset; Copernicus Climate Change Service 2023; https://cds.climate.copernicus.eu/datasets/projections-cmip6; DOI: 10.24381/cds.c866074c) via the **ecmwfr** R package (Hufkens et al. 2019; https://bluegreen-labs.github.io/ecmwfr/). Data were requested for 2015–2100 under three Shared Socioeconomic Pathways: SSP1-2.6 (strong mitigation, ~1.5–2°C global warming by 2100), SSP2-4.5 (intermediate stabilisation), and SSP5-8.5 (high emissions, ~4–5°C warming; O'Neill et al. 2016; Riahi et al. 2017). A historical baseline (1950–2014) was obtained from the same models and stitched to the SSP series at 2015. All data were spatially subsetted to a 3° bounding box centred on Moorea (~16°S to ~19°S, ~148°E to ~151°E) and the nearest grid cell to the LTER Site 1 backreef location was extracted. Monthly *tos* values were converted from Kelvin to Celsius where necessary (units check per NetCDF metadata), aggregated to annual maxima per model-year, and ensemble p10, median, and p90 quantiles were computed across models for each scenario-year.

Each model's annual maximum SST series was delta bias-corrected to the observed MCR 2005–2025 annual maximum temperature mean (additive shift per model, computed over the 2005–2025 overlap window combining the historical run for 2005–2014 and the SSP run for 2015–2025). Using location-specific coupled-model SST eliminates the need for an air-to-SST offset correction and provides warming rates directly relevant to the reef environment. All ten models were confirmed available on the CDS for the `projections-cmip6` dataset with the `sea_surface_temperature` variable.

Projections used a two-phase approach. For the historical period (2005–2025), observed annual z-scores from the fitted dataset were used directly for each ecosystem variable where data were available, replacing the sequential SEM chain prediction; the SEM chain was retained only for 2005 (pre-data) and for variable–year combinations with missing observations. This ensures the historical projection lines track the observed data rather than the amplified sequential predictions of the SEM chain. For the projection period (2026–2100), a delta method anchored the projection to the 2025 *observed* ecosystem state: only the *change* in temperature from the scenario's CMIP6 2025 value was propagated forward through the SEM path coefficients, starting from the actual 2025 coral cover (6.8%), ecosystem respiration (28.5 mmol O₂ m⁻² hr⁻¹), fleshy macroalgae (14.7%), and Pmax (113.3 mmol O₂ m⁻² hr⁻¹). Algal N content in 2025 was imputed from the observed Rd via the SEM regression (N_percent was not measured in 2025). This approach was used for all nine scenario × percentile combinations (3,000 posterior draws each, totalling 27,000 projection runs). The inner uncertainty ribbon on projection figures represents the SEM posterior 95% credible interval using the ensemble median temperature; the outer ribbon represents the CMIP6 multi-model p10–p90 spread at the SEM posterior median, thereby separating climate-forcing uncertainty from model-parameter uncertainty. For the historical period (2005–2014), the outer uncertainty ribbon shows the inter-model p10–p90 spread across the ten bias-corrected CMIP6 historical runs.

## Software and Reproducibility

Key packages: **tidyverse** 2.0 (Wickham et al. 2019), **brms** 2.21 (Bürkner 2017), **tidybayes** 3.0 (Kay 2023), **patchwork** 1.2 (Pedersen 2024), **ggraph** 2.1 and **igraph** 1.6 (Csárdi & Nepusz 2006), **httr2** (Wickham 2023), **zoo** (Zeileis & Grothendieck 2005), **cowplot** (Wilke 2024), **ecmwfr** (Hufkens et al. 2019; https://bluegreen-labs.github.io/ecmwfr/), **ncdf4** (Pierce 2024). Stan 2.32 served as the computational backend. CMIP6 SST data were accessed from the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/datasets/projections-cmip6; DOI: 10.24381/cds.c866074c). All analysis code and processed data are available at \[**INSERT REPOSITORY DOI**\]; MCR LTER raw data are publicly archived through the EDI Data Portal (https://edirepository.org).

---

# Results

## Benthic Community Composition

```{r benthic-plot}
#| label: fig-benthic
#| fig-cap: "**Reef benthic community composition at MCR LTER Site 1 backreef, 2006–2025.** Stacked area plot shows annual mean proportional cover of major benthic functional groups derived from point-intercept photoquadrat transects (6 transects × 30 points per year). Coral cover (green) declined from 31.0% in 2006 to 6.8% in 2025, a 78% reduction over 20 years, while fleshy macroalgae (yellow-green) expanded from 5.5% to 14.7% (+167%). Crustose coralline algae (pink), turf algae (tan), and other categories are included for completeness. Error ribbons show ±1 SE across transects."
#| fig-width: 10
#| fig-height: 6

# Define color palette
myPal <- c(
  Coral = "#c38370",
  `Fleshy Macroalgae` = "#99ddaa",
  `Turf/Cyanobacteria` = "#9da993",
  `Crustose Corallines` = "#523a28",
  Other = "#bdc3cb",
  Sand = "#d6ad60"
)

# Create stacked area plot
Benthic_summary_Algae |>
  mutate(name = factor(name, levels = c(
    "Coral", "Fleshy Macroalgae",
    "Crustose Corallines", "Turf/Cyanobacteria",
    "Sand", "Other"
  ))) |>
  filter(Site == "LTER 1") |>
  ggplot(aes(x = Year, y = mean_cover, fill = name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = myPal) +
  labs(
    fill = "",
    y = "% Cover",
    title = "Long-term Benthic Community Dynamics"
  ) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```

**Key Observation:** Coral cover declined from `r round(Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover), 1)`% in 2006 to `r round(Benthic_summary_Algae |> filter(Year == 2025, name == "Coral", Site == "LTER 1") |> pull(mean_cover), 1)`% in 2025, representing a `r round((1 - (Benthic_summary_Algae |> filter(Year == 2025, name == "Coral", Site == "LTER 1") |> pull(mean_cover)) / (Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover))) * 100, 1)`% reduction.



## Temporal Trends

Assess standardized change per year using Bayesian linear models.

```{r temporal-trends}
#| label: temporal-trends
#| cache: true

# Create standardized dataset
std_data <- Year_Averages |>
  select(
    mean_coral, mean_fleshy, Pmax, Rd, N_percent,
    Herbivores, Corallivore, NP_mean, Max_temp, GP_mean,
    Nitrite_and_Nitrate, Phosphate
  ) |>
  mutate(across(everything(), ~ as.numeric(scale(.x)))) |>
  bind_cols(Year_Averages |> select(Year))

# Fit Bayesian models for each variable and extract R²
fit_temporal_model <- function(response_var, param_name) {
  form <- as.formula(paste(response_var, "~ Year"))
  mod <- brm(form, data = std_data, silent = 2, refresh = 0, seed = 223)
  
  # Extract posterior samples
  post <- as_tibble(as.matrix(mod)) |>
    select(Year = b_Year) |>
    mutate(Parameter = param_name)
  
  # Calculate Bayesian R²
  r2 <- median(bayes_R2(mod)[, 1])
  
  # Store model for later R² extraction
  list(posterior = post, model = mod, r2 = r2, param = param_name)
}

# Fit all models
temporal_models <- list(
  fit_temporal_model("mean_coral", "% Coral Cover"),
  fit_temporal_model("mean_fleshy", "% Macroalgae Cover"),
  fit_temporal_model("Corallivore", "Corallivore"),
  fit_temporal_model("Herbivores", "Herbivore/bioeroding fish"),
  fit_temporal_model("N_percent", "%N Content"),
  fit_temporal_model("Rd", "Ecosystem Respiration"),
  fit_temporal_model("Pmax", "Maximum Photosynthetic Capacity"),
  fit_temporal_model("NP_mean", "Net Ecosystem Production"),
  fit_temporal_model("Max_temp", "Max Temperature"),
  fit_temporal_model("Nitrite_and_Nitrate", "Nitrate + Nitrite")
)

# Extract posteriors for plotting
temporal_results <- temporal_models |>
  map_df(~.x$posterior)

# Print R² values
cat("\n=== Temporal Trend Model Performance (Bayesian R²) ===\n\n")
temporal_models |>
  map_df(~tibble(Parameter = .x$param, R2 = .x$r2)) |>
  arrange(desc(R2)) |>
  print(n = Inf)
```

```{r temporal-plot}
#| label: fig-temporal
#| fig-cap: "**Directional change across all major reef ecosystem variables at MCR LTER Site 1 backreef, 2005–2025.** (a) Posterior distributions of the standardised temporal slope (change per SD yr⁻¹) from independent Bayesian linear regression models fit to z-scored annual data. The vertical dashed line marks zero; white tick marks indicate 95% posterior credible intervals. Red = increasing trend, blue = decreasing trend; intervals excluding zero indicate credible directional change. (b) Raw annual observations (points) overlaid with posterior median trend lines and 95% credible interval ribbons on original measurement scales. Variables shown (top to bottom): coral cover (%), fleshy macroalgae cover (%), crustose coralline algae cover (%), corallivore fish biomass (g m⁻²), herbivore fish biomass (g m⁻²), algal tissue nitrogen content (%), ecosystem respiration Rd (mmol O₂ m⁻² hr⁻¹), maximum photosynthetic capacity Pmax (mmol O₂ m⁻² hr⁻¹), net community production NP (mmol O₂ m⁻² hr⁻¹), and annual maximum sea surface temperature (°C). Bayesian R² values are reported for each trend model."
#| fig-width: 8.9
#| fig-height: 7.5



# Noto Sans has full Unicode coverage (superscripts, µ, °, subscripts)
# Falls back gracefully if offline — font is cached after first download
font_add_google("Noto Sans", family = "noto")

# ── Nature-ready base theme (7 pt, classic, no gridlines) ──────────────────
theme_nature <- function(base_size = 7) {
  theme_classic(base_size = base_size) +
    theme(
      text             = element_text(family = "noto"),
      axis.text        = element_text(size = base_size, color = "black"),
      axis.title       = element_text(size = base_size + 0.5),
      axis.line        = element_line(linewidth = 0.35, color = "black"),
      axis.ticks       = element_line(linewidth = 0.35, color = "black"),
      strip.background = element_blank(),
      strip.text       = element_text(size = 6.5, face = "plain", color = "grey20"),
      panel.spacing    = unit(0.5, "lines"),
      legend.position  = "none",
      plot.tag         = element_text(face = "bold", size = 9)
    )
}

col_inc <- "#E64B35"   # NPG red  — increasing
col_dec <- "#4DBBD5"   # NPG teal — decreasing

# ── Summarise posterior slopes (order + color mapping) ─────────────────────
slope_summary <- temporal_results |>
  group_by(Parameter) |>
  summarise(
    median_slope = median(Year),
    direction    = if_else(median_slope > 0, "Increasing", "Decreasing"),
    .groups = "drop"
  ) |>
  arrange(median_slope)

param_levels <- slope_summary$Parameter

# ── Scale parameters — defined here so they are available to both panels ────
short_label_map <- tibble(
  var = c("mean_coral", "mean_fleshy", "Corallivore", "Herbivores",
          "N_percent", "Rd", "Pmax", "NP_mean", "Max_temp", "Nitrite_and_Nitrate"),
  Parameter = c("% Coral Cover", "% Macroalgae Cover", "Corallivore",
                "Herbivore/bioeroding fish", "%N Content", "Ecosystem Respiration",
                "Maximum Photosynthetic Capacity", "Net Ecosystem Production",
                "Max Temperature", "Nitrate + Nitrite"),
  label = c(
    "Coral cover\n(%)",
    "Macroalgae cover\n(%)",
    "Corallivore\n(g m\u207b\u00b2)",
    "Herbivores\n(g m\u207b\u00b2)",
    "N content\n(%)",
    "Respiration\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "Pmax\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "NEP\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "Max temperature\n(\u00b0C)",
    "Nitrate+nitrite\n(\u00b5mol L\u207b\u00b9)"
  )
)

scale_params <- tibble(
  var = short_label_map$var,
  mu  = map_dbl(short_label_map$var, \(x) mean(Year_Averages[[x]], na.rm = TRUE)),
  sd  = map_dbl(short_label_map$var, \(x) sd(Year_Averages[[x]],   na.rm = TRUE))
) |>
  left_join(short_label_map, by = "var")

label_levels <- scale_params$label[match(param_levels, scale_params$Parameter)]

# Short names for panel a y-axis (strip units line from panel b labels)
short_name_map <- scale_params |>
  select(Parameter, label) |>
  mutate(short_name = str_remove(label, "\n.*"))

# ── Panel a: Posterior density ridges ──────────────────────────────────────
panel_a <- temporal_results |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(short_name_map, by = "Parameter") |>
  mutate(short_name = factor(short_name, levels = short_name_map$short_name[match(param_levels, short_name_map$Parameter)])) |>
  ggplot(aes(x = Year, y = short_name, fill = direction)) +
  geom_vline(
    xintercept = 0, linetype = "dashed",
    linewidth = 0.35, color = "grey55"
  ) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.025, 0.975),
    alpha = 0.75, linewidth = 0.25, color = "white",
    scale = 0.88, rel_min_height = 0.01
  ) +
  scale_fill_manual(values = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  annotate("text", x =  0.06, y = 0.6, label = "Increasing \u2192",
           hjust = 0, size = 1.9, color = col_inc, fontface = "italic", family = "noto") +
  annotate("text", x = -0.06, y = 0.6, label = "\u2190 Decreasing",
           hjust = 1, size = 1.9, color = col_dec, fontface = "italic", family = "noto") +
  labs(x = "Standardized change per year", y = NULL, tag = "a") +
  lims(x = c(-0.35, 0.35)) +
  theme_nature() +
  theme(axis.text.y = element_text(size = 6.5))

# ── Back-transform fitted values and raw data to original scales ────────────

# ── Extract fitted values + raw data from each Bayesian model ───────────────
get_time_series <- function(model_obj) {
  mod      <- model_obj$model
  param    <- model_obj$param
  resp_var <- as.character(formula(mod)$formula)[2]
  year_seq <- tibble(Year = seq(min(std_data$Year), max(std_data$Year), by = 0.5))
  fitted_df <- fitted(mod, newdata = year_seq) |>
    as_tibble() |>
    bind_cols(year_seq) |>
    mutate(Parameter = param)
  raw_df <- std_data |>
    select(Year, value = all_of(resp_var)) |>
    mutate(Parameter = param)
  list(fitted = fitted_df, raw = raw_df)
}

all_series <- map(temporal_models, get_time_series)

fitted_all <- map_df(all_series, ~.x$fitted) |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(scale_params |> select(Parameter, mu, sd, label), by = "Parameter") |>
  mutate(
    Estimate_orig = Estimate * sd + mu,
    Q2.5_orig     = Q2.5    * sd + mu,
    Q97.5_orig    = Q97.5   * sd + mu,
    label = factor(label, levels = label_levels)
  )

raw_all <- map_df(all_series, ~.x$raw) |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(scale_params |> select(Parameter, mu, sd, label), by = "Parameter") |>
  mutate(
    value_orig = value * sd + mu,
    label = factor(label, levels = label_levels)
  )

# ── Bayesian R² annotations ─────────────────────────────────────────────────
r2_df <- map_df(temporal_models, ~tibble(Parameter = .x$param, r2 = .x$r2)) |>
  left_join(scale_params |> select(Parameter, label), by = "Parameter") |>
  mutate(
    label    = factor(label, levels = label_levels),
    r2_label = paste0("R\u00b2 = ", round(r2, 2))
  )

# ── Panel b: Small-multiple time series on original scale ───────────────────
panel_b <- ggplot() +
  geom_ribbon(
    data = fitted_all,
    aes(x = Year, ymin = Q2.5_orig, ymax = Q97.5_orig, fill = direction),
    alpha = 0.18
  ) +
  geom_line(
    data = fitted_all,
    aes(x = Year, y = Estimate_orig, color = direction),
    linewidth = 0.55
  ) +
  geom_point(
    data = raw_all,
    aes(x = Year, y = value_orig),
    size = 0.85, alpha = 0.70, shape = 16, color = "grey25"
  ) +
  geom_text(
    data = r2_df,
    aes(label = r2_label),
    x = -Inf, y = Inf,
    hjust = -0.1, vjust = 1.4,
    size = 1.8, color = "grey45", fontface = "italic", family = "noto"
  ) +
  facet_wrap(~label, scales = "free_y", ncol = 5) +
  scale_fill_manual(values  = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  scale_color_manual(values = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  scale_x_continuous(breaks = c(2008, 2014, 2020)) +
  labs(x = "Year", y = NULL, tag = "b") +
  theme_nature() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 5.5),
    axis.text.y = element_text(size = 5.5)
  )

# ── Combine and render ───────────────────────────────────────────────────────
fig7 <- panel_a / panel_b +
  plot_layout(heights = c(1.05, 1.4)) &
  theme(plot.background = element_blank())

fig7

ggsave(
  here("Output", "fig7_temporal_trends.pdf"),
  plot   = fig7,
  width  = 183,
  height = 155,
  units  = "mm",
  device = cairo_pdf
)
```

**Model Performance Summary**: The Bayesian temporal trend models showed varying explanatory power across ecosystem variables. Temperature and coral cover trends had the highest R² values (>0.5), indicating highly predictable linear trajectories over time. Metabolic variables (Pmax, Rd, NP) showed moderate explanatory power (R² = 0.2-0.5), reflecting both directional trends and substantial year-to-year variability. Fish biomass trends had lower R² values, suggesting additional drivers beyond benthic habitat changes influence these populations.

## Correlation Analysis

```{r correlations}
#| label: fig-correlations
#| eval: false
#| include: false
#| fig-cap: "**Pairwise Pearson correlations among major reef ecosystem variables at MCR LTER Site 1 backreef, 2005–2025.** Upper triangle: ellipses scaled and oriented by Pearson correlation coefficient (r); narrow horizontal ellipses indicate strong positive correlation, narrow vertical ellipses indicate strong negative correlation. Lower triangle: numeric r values. Diagonal: variable names. Asterisks denote significance thresholds from two-tailed t-tests (* p < 0.05, ** p < 0.01, *** p < 0.001; no correction for multiple comparisons). Variables are ordered by hierarchical clustering of the correlation matrix. NP = annual mean net community production (mmol O₂ m⁻² hr⁻¹); Rd = ecosystem dark respiration; Pmax = maximum photosynthetic capacity; %N = algal tissue nitrogen content; N+N = water-column nitrite + nitrate concentration (µmol L⁻¹); % Coral = coral cover; % Algae = fleshy macroalgae cover; Herbs/eroder = herbivorous and bioeroding fish biomass (g m⁻²); Corallivore = corallivore fish biomass; Max Temp = annual maximum sea surface temperature (°C). n = 15–21 year-pairs depending on variable availability."
#| fig-width: 10
# Compute correlation matrix
ct <- corr.test(
  Year_Averages |>
    select(
      NP = NP_mean, Rd, Pmax, `%N` = N_percent,
      `N+N` = Nitrite_and_Nitrate, `% Coral` = mean_coral,
      `% Algae` = mean_fleshy, `Herbs/eroder` = Herbivores,
      `Corallivore` = Corallivore, `Max Temp` = Max_temp
    ),
  adjust = "none"
)

corr <- ct$r
p.mat <- ct$p

ggcorrplot.mixed(corr,
  upper = "ellipse",
  lower = "number",
  p.mat = p.mat,
  insig = "label_sig",
  sig.lvl = c(0.05, 0.01, 0.001)
) +
  labs(title = "Ecosystem Variable Correlations")
```

## Structural Equation Modeling

Fit a Bayesian SEM to understand causal pathways among ecosystem components.

```{r sem-prep}
#| label: sem-preparation

# Calculate year residuals (disturbances beyond temperature)
yearresid <- resid(lm(Year ~ Max_temp, data = Year_Averages, na.action = na.pass))

# Prepare scaled SEM data
semdata2 <- Year_Averages |>
  mutate(
    logfleshy = log(mean_fleshy),
    logrd = log(Rd),
    logpmax = log(Pmax),
    logproducers = log(mean_alive),
    N_percent = log(N_percent),
    yearresid = yearresid
  ) |>
  select(
    Year, meancoral = mean_coral, logproducers, meanfleshy = mean_fleshy,
    temperature = Max_temp, flow = Flow_mean, Npercent = N_percent,
    logrd, logpmax, logcoral = log_coral, logfleshy, Corallivore,
    herbs = Herbivores, yearresid
  ) |>
  drop_na(logcoral) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))
```

```{r sem-fit}
#| label: sem-model
#| cache: true

# Define SEM structure
bf_temp <- bf(temperature | mi() ~ 1 + Year)
bf_coraltemp <- bf(logcoral | trunc(ub = 3.39) ~ 1 + temperature + yearresid)
bf_logpmax <- bf(logpmax | mi() ~ 1 + logcoral + mi(temperature) + mi(logrd)) ####
bf_logrd <- bf(logrd | mi() ~ 1 + logcoral + mi(temperature))
bf_herbs <- bf(herbs | mi() ~ 1 + logfleshy)
bf_corallivore <- bf(Corallivore | mi() ~ 1 + logcoral)
bf_fleshy <- bf(logfleshy | trunc(ub = 2.26) ~ logcoral)
bf_nutrients <- bf(Npercent | mi() ~ 1 + mi(logrd))

# Fit full SEM
brms_sem_full <- brm(
  bf_temp + bf_coraltemp + bf_logrd + bf_herbs + bf_corallivore +
    bf_nutrients + bf_fleshy + bf_logpmax + set_rescor(FALSE),
  data = semdata2,
  chains = 3,
  iter = 10000,
  warmup = 5000,
  seed = 11,
  sample_prior = "no",
  prior = c(
    set_prior("normal(0, 2)", class = "b"),
    set_prior("normal(0, 0.25)", resp = "temperature", class = "sigma", lb = 0),
    set_prior("normal(0, 2)", class = "Intercept")
  ),
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  family = gaussian(),
  silent = 2,
  refresh = 0
)
```

```{r sem-summary}
#| label: sem-results

summary(brms_sem_full)
```

```{r}
#| label: sem-r2-calculation

# Calculate Bayesian R² for each response variable in the SEM
# Note: Some responses with missing data imputation may have NaN - filter these out
sem_r2 <- bayes_R2(brms_sem_full, summary = FALSE)
sem_r2_summary <- apply(sem_r2, 2, median, na.rm = TRUE)

# Remove NaN values
sem_r2_valid <- sem_r2_summary[!is.na(sem_r2_summary)]

cat("\n=== Structural Equation Model: Bayesian R² by Response ===\n\n")
sem_r2_df <- tibble(
  Response = names(sem_r2_valid),
  R2 = sem_r2_valid
) |>
  mutate(Response = case_when(
    Response == "R2temperature" ~ "Temperature",
    Response == "R2logcoral" ~ "Coral Cover",
    Response == "R2logpmax" ~ "Ecosystem Pmax",
    Response == "R2logrd" ~ "Ecosystem Rd",
    Response == "R2herbs" ~ "Herbivorous Fish",
    Response == "R2Corallivore" ~ "Corallivore Fish",
    Response == "R2logfleshy" ~ "Fleshy Macroalgae",
    Response == "R2Npercent" ~ "N Content",
    TRUE ~ Response
  )) |>
  arrange(desc(R2))

print(sem_r2_df, n = Inf)

cat("\n")
cat(sprintf("Mean R² across computable responses: %.3f\n", mean(sem_r2_valid)))
cat(sprintf("The SEM explains between %.1f%% and %.1f%% of variance across responses.\n",
            min(sem_r2_valid) * 100, max(sem_r2_valid) * 100))
cat("Note: R² not computed for responses with extensive missing data imputation.\n")
```

## SEM Conceptual Model

```{r sem-dag}
#| label: fig-sem-dag
#| fig-cap: "**Bayesian structural equation model reveals coral cover as the central mediator of climate-driven ecosystem change.** Directed acyclic graph shows the 11 hypothesised causal pathways among nine ecosystem components, fit simultaneously in a multivariate Bayesian SEM (n = 15 annual observations, 2008–2025). Arrow width is proportional to the absolute standardised path coefficient (|β|); blue = positive effect, red = negative effect; grey = 95% posterior credible interval crosses zero (inconclusive pathway). Standardised β values are shown on each arrow. Shaded background regions distinguish ecosystem function variables (top; Ecosystem Respiration and Ecosystem Pmax in a horizontal row, with Algal N Stock elevated above) from community composition variables (bottom; Fleshy Algae, Herbivore Fish, Corallivore Fish). All continuous variables were log-transformed where appropriate and z-score standardised (mean = 0, SD = 1) prior to analysis to permit direct comparison of effect magnitudes."
#| fig-width: 9
#| fig-height: 4.2

pos_col  <- "#1565C0"
neg_col  <- "#B71C1C"
nsig_col <- "#AAAAAA"

# N Stock elevated above the Rd–Pmax horizontal line (y=5.0 vs Rd/Pmax at y=4.0)
# Rd→Pmax is a straight horizontal beneath N; Rd→N is a short diagonal upward
# All edges are straight segments — no curves at all
# Rows compressed: ecosystem at y=4.0, community at y=2.0, gap = 2.0 units
dag_nodes <- tribble(
  ~id,           ~label,                    ~x,    ~y,    ~fill,       ~border,
  "Year",        "Year",                    0.4,   3.5,   "white",     "#888888",
  "Temp",        "Temperature",             2.0,   3.0,   "white",     "#888888",
  "Coral",       "Coral\nCover",            3.8,   3.0,   "#E8F5E9",   "#2E7D32",
  # Ecosystem function: Rd and Pmax in horizontal row; N elevated above
  "Rd",          "Ecosystem\nRespiration",  5.6,   4.0,   "#EDE7F6",   "#4527A0",
  "N",           "N\nStock",               7.4,   5.0,   "#FFF8E1",   "#E65100",
  "Pmax",        "Ecosystem\nPmax",         9.2,   4.0,   "#EDE7F6",   "#4527A0",
  # Community composition row
  "Algae",       "Fleshy\nAlgae",           5.6,   2.0,   "#E8F5E9",   "#2E7D32",
  "Herbs",       "Herbivore\nFish",         7.4,   2.0,   "#E3F2FD",   "#0D47A1",
  "Corallivore", "Corallivore\nFish",       9.2,   2.0,   "#E3F2FD",   "#0D47A1"
)

dag_edges <- tribble(
  ~from,   ~to,           ~est,   ~sig,  ~loff_x, ~loff_y,
  "Year",  "Temp",        0.465,  TRUE,   0.28,    0.18,
  "Temp",  "Coral",      -0.549,  TRUE,   0.00,    0.24,
  "Temp",  "Rd",          0.214,  FALSE, -0.35,    0.24,
  "Temp",  "Pmax",       -0.158,  FALSE, -0.75,   -0.22,
  "Coral", "Rd",          1.080,  TRUE,  -0.08,    0.24,
  "Coral", "Algae",      -0.914,  TRUE,  -0.08,   -0.24,
  "Coral", "Corallivore", 0.645,  TRUE,   0.38,   -0.20,
  "Coral", "Pmax",       -0.353,  FALSE,  0.38,    0.20,
  "Rd",    "N",           0.623,  TRUE,  -0.18,    0.22,   # short diagonal up-right
  "Rd",    "Pmax",        0.819,  TRUE,   0.00,    0.24,   # straight horizontal below N
  "Algae", "Herbs",       0.711,  TRUE,   0.00,    0.24    # straight horizontal
) |>
  mutate(direction = if_else(est >= 0, "pos", "neg"), abs_est = abs(est)) |>
  left_join(dag_nodes |> select(id, x, y) |> rename(from = id, x0 = x, y0 = y), by = "from") |>
  left_join(dag_nodes |> select(id, x, y) |> rename(to   = id, x1 = x, y1 = y), by = "to") |>
  mutate(mx = (x0 + x1) / 2, my = (y0 + y1) / 2, lbl = sprintf("%.2f", est))

# All edges are straight — no geom_curve needed
dag_arr <- arrow(length = unit(3.0, "mm"), type = "closed")

ggplot() +
  # Faint region backgrounds
  annotate("rect", xmin = 4.8, xmax = 10.0, ymin = 3.2, ymax = 5.65,
           fill = "#EDE7F6", alpha = 0.18, color = NA) +
  annotate("rect", xmin = 4.8, xmax = 10.0, ymin = 1.2, ymax = 2.75,
           fill = "#E3F2FD", alpha = 0.18, color = NA) +
  # Region labels
  annotate("text", x = 10.12, y = 4.5, label = "Ecosystem\nfunction",
           hjust = 0, size = 2.6, color = "#6A1B9A", fontface = "italic", lineheight = 0.9) +
  annotate("text", x = 10.12, y = 2.0, label = "Community\ncomposition",
           hjust = 0, size = 2.6, color = "#1565C0", fontface = "italic", lineheight = 0.9) +
  # Significant edges
  geom_segment(data = filter(dag_edges,  sig),
    aes(x = x0, y = y0, xend = x1, yend = y1, color = direction, linewidth = abs_est),
    alpha = 0.82, arrow = dag_arr, lineend = "butt") +
  # Non-significant edges — reduced alpha to de-emphasise
  geom_segment(data = filter(dag_edges, !sig),
    aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = abs_est),
    color = nsig_col, alpha = 0.45, arrow = dag_arr, lineend = "butt") +
  geom_label(data = filter(dag_edges,  sig),
    aes(x = mx + loff_x, y = my + loff_y, label = lbl, color = direction),
    size = 2.8, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.07, "lines"), show.legend = FALSE) +
  geom_label(data = filter(dag_edges, !sig),
    aes(x = mx + loff_x, y = my + loff_y, label = lbl),
    color = nsig_col, size = 2.8, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.07, "lines"), show.legend = FALSE) +
  # Node boxes
  {lapply(seq_len(nrow(dag_nodes)), function(i) {
    nd <- dag_nodes[i, ]
    geom_label(data = nd, aes(x = x, y = y, label = label),
               fill = nd$fill, color = nd$border, size = 3.0,
               fontface = "bold", lineheight = 0.9, label.size = 0.65,
               label.padding = unit(0.38, "lines"), label.r = unit(4, "pt"),
               show.legend = FALSE)
  })} +
  scale_color_manual(values = c("pos" = pos_col, "neg" = neg_col),
                     labels = c("pos" = "Positive", "neg" = "Negative"),
                     name = "Effect direction") +
  scale_linewidth(range = c(0.6, 2.8), guide = "none") +
  labs(caption = "Standardized β on arrows  ·  gray = 95% CI crosses zero") +
  theme_void(base_size = 12) +
  theme(
    plot.caption    = element_text(hjust = 0.5, color = "grey55", size = 9, margin = margin(t = 6)),
    legend.position = "bottom",
    legend.title    = element_text(size = 10, face = "bold"),
    legend.text     = element_text(size = 9),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin     = margin(8, 10, 6, 10)
  ) +
  guides(color = guide_legend(override.aes = list(linewidth = 1.5))) +
  coord_cartesian(xlim = c(-0.1, 11.8), ylim = c(1.35, 5.75))
```

## SEM Path Coefficients

```{r sem-paths}
#| label: fig-sem-paths
#| fig-cap: "**Standardised path coefficients from the Bayesian multivariate structural equation model.** Each panel corresponds to one response variable in the SEM; the x-axis shows the standardised effect size (β) and the y-axis shows the predictor(s) for that response. Points indicate posterior medians; horizontal bars span 95% credible intervals. Filled red symbols (full opacity) indicate paths whose 95% credible interval excludes zero; faded symbols indicate inconclusive paths (interval crosses zero). All variables were z-score standardised prior to modelling, so β values are directly comparable across panels as standard-deviation-per-standard-deviation effects. Coefficients are derived from 15,000 post-warmup MCMC draws of the full 12-pathway SEM (n = 15–21 annual observations, 2005–2025, depending on variable). Temperature and Ecosystem Rd Rate panels show imputed-predictor coefficients (mi() terms in brms) for years with missing observations."
#| fig-width: 16
#| fig-height: 6

# Extract and format path coefficients
gp_edges <- as_tibble(fixef(brms_sem_full)) |>
  mutate(coef = rownames(fixef(brms_sem_full))) |>
  separate(col = coef, into = c("resp", "pred")) |>
  filter(pred != "Intercept") |>
  rename(est = Estimate, lo = `Q2.5`, hi = `Q97.5`) |>
  mutate(sig = if_else(sign(lo) == sign(hi), 1, 0.5)) |>
  mutate(resp = case_when(
    resp == "temperature" ~ "Temperature",
    resp == "logfleshy" ~ "Fleshy Macroalgae",
    resp == "logpmax" ~ "Ecosystem Pmax",
    resp == "logrd" ~ "Ecosystem Rd",
    resp == "herbs" ~ "Herbivorous Fish",
    resp == "Corallivore" ~ "Corallivore Fish",
    resp == "logcoral" ~ "Coral",
    resp == "Npercent" ~ "N stock"
  )) |>
  mutate(resp = factor(resp, c(
    "Temperature", "Coral", "Fleshy Macroalgae", "Herbivorous Fish",
    "Corallivore Fish", "Ecosystem Rd", "Ecosystem Pmax", "N stock"
  ))) |>
  mutate(pred = case_when(
    pred == "temperature" ~ "Temperature",
    pred == "mitemperature" ~ "Temperature",
    pred == "milogrd" ~ "Ecosystem Rd Rate",
    pred == "logcoral" ~ "Coral Cover",
    pred == "logfleshy" ~ "Fleshy Macroalgae Cover",
    pred == "Year" ~ "Year",
    pred == "yearresid" ~ "Non-Temperature Yearly Variability"
  )) |>
  mutate(pred = factor(pred, c(
    "Ecosystem Rd Rate", "Fleshy Macroalgae Cover",
    "Coral Cover", "Non-Temperature Yearly Variability",
    "Temperature", "Year"
  )))

gp_edges |>
  ggplot(aes(x = est, y = pred)) +
  geom_point(size = 3, color = "firebrick", aes(alpha = sig)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi, alpha = sig), height = 0, color = "firebrick") +
  geom_vline(xintercept = 0) +
  scale_alpha(range = c(0.6, 1)) +
  labs(
    x = "Standardized Effect Size",
    y = "Predictor Variables",
    title = "SEM Path Coefficients"
  ) +
  theme_minimal() +
  facet_wrap(~resp, nrow = 1) +
  theme(
    text = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.position = "none"
  )
```

## Temperature Effects: Direct vs Indirect

```{r temp-effects-data}
#| label: temp-effects-data
#| echo: false
#| message: false
#| warning: false
#| cache: true

# ── Compute direct and indirect temperature effects from SEM posteriors ────────
# post_samples: 15,000 post-warmup draws from the full SEM
post_samples <- as_tibble(as.matrix(brms_sem_full))

# Individual path coefficient draws
b_tc  <- post_samples$b_logcoral_temperature       # Temp → Coral (direct)
b_cr  <- post_samples$b_logrd_logcoral             # Coral → Rd
b_cf  <- post_samples$b_logfleshy_logcoral         # Coral → Fleshy
b_cc  <- post_samples$b_Corallivore_logcoral       # Coral → Corallivore Fish
b_cp  <- post_samples$b_logpmax_logcoral           # Coral → Pmax
b_fh  <- post_samples$b_herbs_logfleshy            # Fleshy → Herbivore Fish
b_rn  <- post_samples$bsp_Npercent_milogrd         # Rd → N Content
b_tr  <- post_samples$bsp_logrd_mitemperature      # Temp → Rd (direct)
b_tp  <- post_samples$bsp_logpmax_mitemperature    # Temp → Pmax (direct)

n_draws <- nrow(post_samples)
zeros   <- rep(0, n_draws)

# Direct, indirect, and total effects for each response variable
effect_specs <- list(
  `Coral Cover`       = list(direct = b_tc,   indirect = zeros),
  `Ecosystem Rd`      = list(direct = b_tr,   indirect = b_tc * b_cr),
  `Fleshy Macroalgae` = list(direct = zeros,  indirect = b_tc * b_cf),
  `Corallivore Fish`  = list(direct = zeros,  indirect = b_tc * b_cc),
  `Ecosystem Pmax`    = list(direct = b_tp,   indirect = b_tc * b_cp),
  `Herbivore Fish`    = list(direct = zeros,  indirect = b_tc * b_cf * b_fh),
  `N Content`         = list(direct = zeros,  indirect = b_tc * b_cr * b_rn)
)

summarise_effect <- function(samps) {
  tibble(
    samples     = list(samps),
    median      = median(samps),
    lower       = quantile(samps, 0.025),
    upper       = quantile(samps, 0.975),
    significant = sign(quantile(samps, 0.025)) == sign(quantile(samps, 0.975))
  )
}

all_effects <- imap_dfr(effect_specs, function(spec, resp_name) {
  bind_rows(
    summarise_effect(spec$direct)               |> mutate(response = resp_name, effect_type = "Direct"),
    summarise_effect(spec$indirect)             |> mutate(response = resp_name, effect_type = "Indirect"),
    summarise_effect(spec$direct + spec$indirect) |> mutate(response = resp_name, effect_type = "Total")
  )
}) |>
  mutate(
    response    = factor(response,    levels = names(effect_specs)),
    effect_type = factor(effect_type, levels = c("Direct", "Indirect", "Total"))
  )
```

```{r temp-indirect-ranking}
#| label: fig-temp-indirect
#| fig-cap: "**Coral-mediated indirect effects of temperature warming on reef ecosystem components.** Horizontal bars show the posterior median standardised indirect effect of annual maximum sea surface temperature on each ecosystem variable, transmitted entirely through coral cover loss (β_indirect = β(Temperature→Coral) × β(Coral→Response)). Error bars span 95% credible intervals derived from 15,000 post-warmup MCMC draws of the full Bayesian SEM. Blue bars indicate positive indirect effects (temperature → coral decline → increase in response variable); red bars indicate negative indirect effects (temperature → coral decline → decrease in response variable). Opacity distinguishes credible effects (95% CI excludes zero; full opacity) from inconclusive effects (CI crosses zero; faded). Variables are ranked from most negative to most positive indirect effect. All variables were z-score standardised prior to modelling; effect sizes are in units of standard deviations of the response per standard deviation of temperature."
#| fig-width: 8
#| fig-height: 6

# Extract only indirect effects and rank them
indirect_effects <- all_effects |>
  filter(effect_type == "Indirect") |>
  # Remove coral (no indirect effect)
  filter(response != "Coral Cover") |>
  arrange(median) |>
  mutate(
    response = fct_reorder(response, median),
    direction = if_else(median < 0, "Negative", "Positive")
  )

# Create ranked bar plot
indirect_effects |>
  ggplot(aes(x = response, y = median, fill = direction, alpha = significant)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    linewidth = 0.8,
    color = "gray20",
    alpha = 0.6
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Negative" = "#E64B35", "Positive" = "#4DBBD5"),
    name = "Effect Direction"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 0.9, "FALSE" = 0.2),
    guide = "none"
  ) +
  labs(
    x = "",
    y = "Standardized Indirect Effect Size",
    title = "Indirect Effects of Temperature via Coral Decline",
  #  subtitle = "Ranked from negative to positive effects with 95% credible intervals"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.y = element_text(size = 16, hjust = 1),
    axis.text.x = element_text(size = 15),
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
   # legend.position = "bottom",
    legend.position = "none",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 19, face = "bold"),
    plot.subtitle = element_text(size = 15, color = "gray40")
  )

# Print ranked summary
cat("\n=== Ranked Indirect Effects (negative to positive) ===\n\n")
indirect_effects |>
  select(response, median, lower, upper, significant) |>
  mutate(across(median:upper, ~round(.x, 3))) |>
  arrange(median) |>
  print(n = Inf)
```

## SEM Conditional Effects

```{r sem-conditional}
#| label: fig-sem-conditional
#| fig-cap: "**Conditional effect plots for the eight significant causal relationships in the Bayesian SEM.** Each panel shows the marginal posterior-predicted relationship for a predictor–response pair, with all other model predictors held at their posterior means. Lines show posterior means; shaded ribbons show 95% credible intervals. Coloured points are annual observed values (back-transformed from z-score scale to original units). Panels are arranged in causal order from climate forcing through coral-mediated change to ecosystem function: (A) Year → Temperature, (B) Temperature → Coral Cover, (C) Coral Cover → Fleshy Algae, (D) Coral Cover → Ecosystem Respiration, (E) Coral Cover → Corallivore Fish, (F) Fleshy Algae → Herbivore Fish, (G) Ecosystem Respiration → Algal N Content, (H) Ecosystem Respiration → Ecosystem Pmax. Conditional effects were computed using the conditional_effects() function in **brms** and back-transformed to measurement units using study-wide z-score parameters."
#| fig-width: 10
#| fig-height: 16

# Unscaling parameters
Yearmodelvalues <- Year_Averages |>
  mutate(
    logfleshy    = log(mean_fleshy),
    logproducers = log(mean_alive),
    logrd        = log(Rd),
    logpmax      = log(Pmax),
    N_percent    = log(N_percent),
    yearresid    = yearresid
  ) |>
  select(
    Year, meancoral = mean_coral, meanfleshy = mean_fleshy,
    temperature = Max_temp, flow = Flow_mean, Npercent = N_percent,
    logrd, logpmax, logcoral = log_coral, logproducers,
    logfleshy, Corallivore, herbs = Herbivores, yearresid
  ) |>
  drop_na(logcoral)

modelmean <- Yearmodelvalues |> summarise(across(everything(), mean, na.rm = TRUE))
modelsd   <- Yearmodelvalues |> summarise(across(everything(), sd,   na.rm = TRUE))

# ── Row 1: Climate drivers ─────────────────────────────────────────────────

# Year → Temperature
temp_year <- conditional_effects(brms_sem_full, resp = "temperature", effects = "Year")
ty_plot <- as_tibble(temp_year$temperature.temperature_Year) |>
  mutate(
    Year       = Year       * modelsd$Year        + modelmean$Year,
    estimate__ = estimate__ * modelsd$temperature + modelmean$temperature,
    lower__    = lower__    * modelsd$temperature + modelmean$temperature,
    upper__    = upper__    * modelsd$temperature + modelmean$temperature
  ) |>
  ggplot(aes(x = Year, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = Year, y = Max_temp), alpha = 0.5) +
  labs(x = "Year", y = "Temperature (°C)", title = "Year → Temperature")

# Temperature → Coral Cover
coral_temp <- conditional_effects(brms_sem_full, resp = "logcoral", effects = "temperature")
ct_plot <- as_tibble(coral_temp$logcoral.logcoral_temperature) |>
  mutate(
    temperature = temperature * modelsd$temperature + modelmean$temperature,
    estimate__  = exp(estimate__ * modelsd$logcoral + modelmean$logcoral),
    lower__     = exp(lower__    * modelsd$logcoral + modelmean$logcoral),
    upper__     = exp(upper__    * modelsd$logcoral + modelmean$logcoral)
  ) |>
  ggplot(aes(x = temperature, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = Max_temp, y = exp(log_coral)), alpha = 0.5) +
  coord_trans(y = "log") +
  scale_y_continuous(breaks = c(1, 5, 10, 20, 40)) +
  labs(x = "Temperature (°C)", y = "Coral Cover (%)", title = "Temperature → Coral Cover")

# ── Row 2: Coral → benthos ────────────────────────────────────────────────

# Coral Cover → Fleshy Algae
fleshy_coral <- conditional_effects(brms_sem_full, resp = "logfleshy", effects = "logcoral")
fc_plot <- as_tibble(fleshy_coral$logfleshy.logfleshy_logcoral) |>
  mutate(
    logcoral   = exp(logcoral   * modelsd$logcoral   + modelmean$logcoral),
    estimate__ = exp(estimate__ * modelsd$logfleshy  + modelmean$logfleshy),
    lower__    = exp(lower__    * modelsd$logfleshy  + modelmean$logfleshy),
    upper__    = exp(upper__    * modelsd$logfleshy  + modelmean$logfleshy)
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = mean_fleshy), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(x = "Coral Cover (%)", y = "Fleshy Macroalgae Cover (%)",
       title = "Coral Cover → Fleshy Algae")

# Coral Cover → Ecosystem Respiration
rd_coral <- conditional_effects(brms_sem_full, resp = "logrd", effects = "logcoral")
rc_plot <- as_tibble(rd_coral$logrd.logrd_logcoral) |>
  mutate(
    logcoral   = exp(logcoral   * modelsd$logcoral + modelmean$logcoral),
    estimate__ = exp(estimate__ * modelsd$logrd    + modelmean$logrd),
    lower__    = exp(lower__    * modelsd$logrd    + modelmean$logrd),
    upper__    = exp(upper__    * modelsd$logrd    + modelmean$logrd)
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = Rd), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = "Coral Cover (%)",
    y = expression(atop("Ecosystem Respiration", "(mmol O"[2]*" m"^-2*" hr"^-1*")")),
    title = "Coral Cover → Respiration"
  )

# ── Row 3: Coral → fish ───────────────────────────────────────────────────

# Coral Cover → Corallivore Fish
corallivore_coral <- conditional_effects(brms_sem_full, resp = "Corallivore", effects = "logcoral")
cc_plot <- as_tibble(corallivore_coral$Corallivore.Corallivore_logcoral) |>
  mutate(
    logcoral   = exp(logcoral * modelsd$logcoral   + modelmean$logcoral),
    estimate__ = estimate__ * modelsd$Corallivore  + modelmean$Corallivore,
    lower__    = lower__    * modelsd$Corallivore  + modelmean$Corallivore,
    upper__    = upper__    * modelsd$Corallivore  + modelmean$Corallivore
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = Corallivore), alpha = 0.5) +
  coord_trans(x = "log") +
  labs(
    x = "Coral Cover (%)",
    y = expression(atop("Corallivore Fish Biomass", "(g m"^-2*")")),
    title = "Coral Cover → Corallivore Fish"
  )

# Fleshy Algae → Herbivore Fish
herbs_fleshy <- conditional_effects(brms_sem_full, resp = "herbs", effects = "logfleshy")
hf_plot <- as_tibble(herbs_fleshy$herbs.herbs_logfleshy) |>
  mutate(
    logfleshy  = exp(logfleshy  * modelsd$logfleshy + modelmean$logfleshy),
    estimate__ = estimate__ * modelsd$herbs         + modelmean$herbs,
    lower__    = lower__    * modelsd$herbs         + modelmean$herbs,
    upper__    = upper__    * modelsd$herbs         + modelmean$herbs
  ) |>
  ggplot(aes(x = logfleshy, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = mean_fleshy, y = Herbivores), alpha = 0.5) +
  coord_trans(x = "log") +
  labs(
    x = "Fleshy Macroalgae Cover (%)",
    y = expression(atop("Herbivore Fish Biomass", "(g m"^-2*")")),
    title = "Fleshy Algae → Herbivore Fish"
  )

# ── Row 4: Metabolism → ecosystem function ────────────────────────────────

# Respiration → N Content
N_Rd <- conditional_effects(brms_sem_full, resp = "Npercent", effects = "logrd")
nr_plot <- as_tibble(N_Rd$Npercent.Npercent_logrd) |>
  mutate(
    logrd      = exp(logrd      * modelsd$logrd    + modelmean$logrd),
    estimate__ = exp(estimate__ * modelsd$Npercent + modelmean$Npercent),
    lower__    = exp(lower__    * modelsd$Npercent + modelmean$Npercent),
    upper__    = exp(upper__    * modelsd$Npercent + modelmean$Npercent)
  ) |>
  ggplot(aes(x = logrd, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = Rd, y = N_percent), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = expression("Ecosystem Respiration (mmol O"[2]*" m"^-2*" hr"^-1*")"),
    y = "Nitrogen Content (%)",
    title = "Respiration → Algal N Content"
  )

# Respiration → Ecosystem Pmax
rd_pmax <- conditional_effects(brms_sem_full, resp = "logpmax", effects = "logrd")
rp_plot <- as_tibble(rd_pmax$logpmax.logpmax_logrd) |>
  mutate(
    logrd      = exp(logrd      * modelsd$logrd   + modelmean$logrd),
    estimate__ = exp(estimate__ * modelsd$logpmax + modelmean$logpmax),
    lower__    = exp(lower__    * modelsd$logpmax + modelmean$logpmax),
    upper__    = exp(upper__    * modelsd$logpmax + modelmean$logpmax)
  ) |>
  ggplot(aes(x = logrd, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_line(linewidth = 1) +
  geom_point(data = Year_Averages, aes(x = Rd, y = Pmax), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = expression("Ecosystem Respiration (mmol O"[2]*" m"^-2*" hr"^-1*")"),
    y = expression(atop("Ecosystem Pmax", "(mmol O"[2]*" m"^-2*" hr"^-1*")")),
    title = "Respiration → Ecosystem Pmax"
  )

# ── Assemble 4 × 2 grid ───────────────────────────────────────────────────
(ty_plot | ct_plot) /
  (fc_plot | rc_plot) /
  (cc_plot | hf_plot) /
  (nr_plot | rp_plot) +
  plot_annotation(tag_levels = "A") &
  theme_classic(base_size = 11) &
  theme(plot.title = element_text(size = 10, face = "bold"))
```

## Ecosystem Projections

```{r cover-ceilings}
# z-log scale ceilings corresponding to 100% benthic cover.
# Both cover equations were fitted with trunc(ub = z_ceil) to respect the
# 0–100% biological bound. The same ceiling is applied in all projection
# matrices so that manual propagation is consistent with the SEM likelihood.
# Values are hardcoded (verified: back-transform to exactly 100% cover).
z_ceil_coral  <- 3.39   # (log(100) - mean(log(coral%))) / sd(log(coral%))
z_ceil_fleshy <- 2.26   # (log(100) - mean(log(fleshy%))) / sd(log(fleshy%))
```

```{r full-sem-proj-data}
#| label: full-sem-proj-data
#| echo: false
#| message: false
#| warning: false
# Compute full-SEM projection matrices — needed by the reduced SEM comparison plot.

# ── Projection years (standardised to SEM scale) ─────────────────────────────
proj_years <- 2005:2055
year_std   <- (proj_years - modelmean$Year) / modelsd$Year

# 3 000 thinned posterior draws
set.seed(7341)
post <- as_draws_df(brms_sem_full) |>
  as_tibble() |>
  slice_sample(n = 3000)

n <- nrow(post)
p <- length(proj_years)

# Helper: multiply each row of an (n × p) matrix by a length-n coefficient vector
rs <- function(coef, mat) sweep(mat, 1, coef, "*")
yr_mat <- matrix(rep(year_std, each = n), nrow = n)  # n × p

# ── Propagate causal chain through the full SEM ───────────────────────────────
# 1. Temperature (z-scored)
temp_mat <- rs(post$b_temperature_Year, yr_mat) + post$b_temperature_Intercept

# ── yearresid matrix: observed z-scaled values for hindcast (2005–2025),
#    0 (standardised mean) for forecast years (2026–2055). yearresid captures
#    year-specific temperature anomalies beyond the linear trend; setting it to 0
#    for forecast years assumes future deviations are unknown and average out.
obs_yr_actual   <- round(semdata2$Year * modelsd$Year + modelmean$Year)
yr_resid_lookup <- tibble(Year = obs_yr_actual, yr = semdata2$yearresid)
yr_resid_vec    <- sapply(proj_years, function(y) {
  row <- yr_resid_lookup |> filter(Year == y)
  if (nrow(row) == 1 && !is.na(row$yr)) row$yr else 0
})
yr_resid_mat <- matrix(rep(yr_resid_vec, each = n), nrow = n)

# 2. Coral cover (log-standardised; temperature + yearresid predictors)
coral_mat <- rs(post$b_logcoral_temperature, temp_mat)    +
             rs(post$b_logcoral_yearresid,   yr_resid_mat) +
             post$b_logcoral_Intercept

# 3. Ecosystem respiration (coral + direct temp via bsp for mi() imputed predictor)
rd_mat <- rs(post$b_logrd_logcoral,          coral_mat) +
          rs(post$bsp_logrd_mitemperature,   temp_mat)  +
          post$b_logrd_Intercept

# 4. Fleshy macroalgae (log-standardised; coral predictor)
fleshy_mat <- pmin(rs(post$b_logfleshy_logcoral, coral_mat) + post$b_logfleshy_Intercept,
                   z_ceil_fleshy)

# 5. Corallivore fish (z-scored; coral predictor)
cov_mat <- rs(post$b_Corallivore_logcoral, coral_mat) + post$b_Corallivore_Intercept

# 6. Herbivore fish (z-scored; fleshy algae predictor)
herb_mat <- rs(post$b_herbs_logfleshy, fleshy_mat) + post$b_herbs_Intercept

# 7. Ecosystem Pmax (log-standardised; rd + coral + temp, all via bsp/b)
pmax_mat <- rs(post$bsp_logpmax_milogrd,      rd_mat)    +
            rs(post$b_logpmax_logcoral,        coral_mat) +
            rs(post$bsp_logpmax_mitemperature, temp_mat)  +
            post$b_logpmax_Intercept

# 8. Algal N content (log-standardised; rd via bsp imputed)
n_mat <- rs(post$bsp_Npercent_milogrd, rd_mat) + post$b_Npercent_Intercept

# ── Back-transform to original units ─────────────────────────────────────────
bt_log <- function(mat, mu, sig) exp(mat * sig + mu)
bt_lin <- function(mat, mu, sig) mat * sig + mu

proj_list <- list(
  Temperature    = bt_lin(temp_mat,   modelmean$temperature, modelsd$temperature),
  `Coral Cover`  = bt_log(coral_mat,  modelmean$logcoral,    modelsd$logcoral),
  Respiration    = bt_log(rd_mat,     modelmean$logrd,       modelsd$logrd),
  `Fleshy Algae` = bt_log(fleshy_mat, modelmean$logfleshy,   modelsd$logfleshy),
  Pmax           = bt_log(pmax_mat,   modelmean$logpmax,     modelsd$logpmax),
  # N_percent is log-transformed in semdata2; use explicit log-scale params
  `N Content`    = bt_log(n_mat,
                          mean(log(Year_Averages$N_percent), na.rm = TRUE),
                          sd(log(Year_Averages$N_percent),   na.rm = TRUE))
)

# Derive P:R ratio directly from back-transformed Pmax and Rd matrices
proj_list[["P:R Ratio"]] <- proj_list$Pmax / proj_list$Respiration

# ── Summarise: posterior median + 50% + 95% CI per variable per year ─────────
proj_tbl <- imap_dfr(proj_list, function(mat, vname) {
  tibble(variable = vname,
         Year     = rep(proj_years, each = n),
         value    = as.vector(mat)) |>
    group_by(variable, Year) |>
    summarise(median = median(value),
              lo95   = quantile(value, 0.025), hi95 = quantile(value, 0.975),
              lo50   = quantile(value, 0.25),  hi50 = quantile(value, 0.75),
              .groups = "drop")
})

# ── Observed data for overlay ─────────────────────────────────────────────────
obs_proj <- Year_Averages |>
  transmute(Year,
            Temperature    = Max_temp,
            `Coral Cover`  = mean_coral,
            Respiration    = Rd,
            `Fleshy Algae` = mean_fleshy,
            Pmax           = Pmax,
            `N Content`    = N_percent,
            `P:R Ratio`    = Pmax / Rd) |>
  pivot_longer(-Year, names_to = "variable", values_to = "observed")
```

## Reduced SEM: Parsimonious Model

Three of the twelve pathways in the full SEM had 95% credible intervals spanning zero — direct temperature effects on respiration (β = 0.22, 95% CI −0.22 to 0.62), direct temperature effects on Pmax (β = −0.16, −0.78 to 0.45), and a direct coral effect on Pmax (β = −0.35, −1.40 to 0.71) — indicating that these pathways are not reliably supported by the data. Retaining them in projection creates two problems: the positive temperature→Rd coefficient generates a spurious upward push on Rd when temperature is extrapolated beyond the fitted range, and the large uncertainty on the coral→Pmax and temperature→Pmax terms inflates Pmax credible intervals. A reduced 9-pathway SEM was therefore fitted with these three terms removed.

```{r sem-reduced-fit}
#| label: sem-reduced-model
#| cache: true

# Reduced SEM: remove 3 non-significant pathways
#   1. Temperature → Rd (direct):      remove mi(temperature) from bf_logrd
#   2. Temperature → Pmax (direct):    remove mi(temperature) from bf_logpmax
#   3. Coral → Pmax (direct):          remove logcoral from bf_logpmax

bf_temp_r        <- bf(temperature | mi()             ~ 1 + Year)
bf_coraltemp_r   <- bf(logcoral    | trunc(ub = 3.39) ~ 1 + temperature + yearresid)
bf_logrd_r       <- bf(logrd       | mi()             ~ 1 + logcoral)
bf_logpmax_r     <- bf(logpmax     | mi()             ~ 1 + mi(logrd))
bf_fleshy_r      <- bf(logfleshy   | trunc(ub = 2.26) ~ logcoral)
bf_herbs_r       <- bf(herbs       | mi()             ~ 1 + logfleshy)
bf_corallivore_r <- bf(Corallivore | mi()             ~ 1 + logcoral)
bf_nutrients_r   <- bf(Npercent    | mi()             ~ 1 + mi(logrd))

brms_sem_reduced <- brm(
  bf_temp_r + bf_coraltemp_r + bf_logrd_r + bf_herbs_r + bf_corallivore_r +
    bf_nutrients_r + bf_fleshy_r + bf_logpmax_r + set_rescor(FALSE),
  data    = semdata2,
  chains  = 3, iter = 10000, warmup = 5000, seed = 11,
  sample_prior = "no",
  prior   = c(
    set_prior("normal(0, 2)",    class = "b"),
    set_prior("normal(0, 0.25)", resp = "temperature", class = "sigma", lb = 0),
    set_prior("normal(0, 2)",    class = "Intercept")
  ),
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  family  = gaussian(),
  silent = 2, refresh = 0
)
```

```{r sem-reduced-summary}
#| label: sem-reduced-results

summary(brms_sem_reduced)
```

```{r sem-reduced-r2}
#| label: sem-reduced-r2

# Manual R² for mi()-imputed responses
for (resp in c("logrd", "logpmax")) {
  fit_f <- fitted(brms_sem_full,    resp = resp)[, "Estimate"]
  fit_r <- fitted(brms_sem_reduced, resp = resp)[, "Estimate"]
  obs_v <- semdata2[[resp]]
  ok    <- !is.na(obs_v)
  cat(sprintf("%-10s  Full R²: %.3f   Reduced R²: %.3f\n", resp,
              cor(fit_f[ok], obs_v[ok])^2, cor(fit_r[ok], obs_v[ok])^2))
}
```

```{r sem-reduced-projections}
#| label: reduced-sem-projection-data

set.seed(6214)
post_r <- as_draws_df(brms_sem_reduced) |> as_tibble() |> slice_sample(n = 3000)
n_r    <- nrow(post_r)

# yearresid matrix: observed for 2005-2025, zero for 2026-2055
yr_resid_vec_r <- sapply(proj_years, function(y) {
  row <- yr_resid_lookup |> filter(Year == y)
  if (nrow(row) == 1 && !is.na(row$yr)) row$yr else 0
})
yr_resid_mat_r <- matrix(rep(yr_resid_vec_r, each = n_r), nrow = n_r)
yr_mat_r       <- matrix(rep(year_std, each = n_r), nrow = n_r)

# ── Propagate reduced SEM causal chain ────────────────────────────────────────
temp_mat_r   <- rs(post_r$b_temperature_Year,     yr_mat_r)   + post_r$b_temperature_Intercept
coral_mat_r  <- rs(post_r$b_logcoral_temperature, temp_mat_r) +
                rs(post_r$b_logcoral_yearresid,   yr_resid_mat_r) +
                post_r$b_logcoral_Intercept
rd_mat_r     <- rs(post_r$b_logrd_logcoral, coral_mat_r)      + post_r$b_logrd_Intercept
fleshy_mat_r <- pmin(rs(post_r$b_logfleshy_logcoral, coral_mat_r) + post_r$b_logfleshy_Intercept,
                     z_ceil_fleshy)
cov_mat_r    <- rs(post_r$b_Corallivore_logcoral, coral_mat_r)+ post_r$b_Corallivore_Intercept
herb_mat_r   <- rs(post_r$b_herbs_logfleshy, fleshy_mat_r)    + post_r$b_herbs_Intercept
pmax_mat_r   <- rs(post_r$bsp_logpmax_milogrd, rd_mat_r)      + post_r$b_logpmax_Intercept
n_mat_r      <- rs(post_r$bsp_Npercent_milogrd, rd_mat_r)     + post_r$b_Npercent_Intercept

proj_list_r <- list(
  Temperature    = bt_lin(temp_mat_r,   modelmean$temperature, modelsd$temperature),
  `Coral Cover`  = bt_log(coral_mat_r,  modelmean$logcoral,    modelsd$logcoral),
  Respiration    = bt_log(rd_mat_r,     modelmean$logrd,       modelsd$logrd),
  `Fleshy Algae` = bt_log(fleshy_mat_r, modelmean$logfleshy,   modelsd$logfleshy),
  Pmax           = bt_log(pmax_mat_r,   modelmean$logpmax,     modelsd$logpmax),
  `N Content`    = bt_log(n_mat_r,
                          mean(log(Year_Averages$N_percent), na.rm = TRUE),
                          sd(log(Year_Averages$N_percent),   na.rm = TRUE))
)
proj_list_r[["P:R Ratio"]] <- proj_list_r$Pmax / proj_list_r$Respiration

proj_tbl_r <- imap_dfr(proj_list_r, function(mat, vname) {
  tibble(variable = vname, Year = rep(proj_years, each = n_r), value = as.vector(mat)) |>
    group_by(variable, Year) |>
    summarise(median = median(value),
              lo95 = quantile(value, .025), hi95 = quantile(value, .975),
              lo50 = quantile(value, .25),  hi50 = quantile(value, .75),
              .groups = "drop")
})

# ── Counterfactual: temperature stabilised at 2025 ────────────────────────────
# temp_2025_std: z-standardised 2025 annual maximum temperature
# = (observed 2025 max temp - SEM temperature mean) / SEM temperature SD
temp_2025_std <- (Year_Averages$Max_temp[Year_Averages$Year == 2025] -
                    modelmean$temperature) / modelsd$temperature
temp_mat_r_cf <- temp_mat_r
temp_mat_r_cf[, which(proj_years > 2025)] <- temp_2025_std

coral_mat_r_cf  <- rs(post_r$b_logcoral_temperature, temp_mat_r_cf)   +
                   rs(post_r$b_logcoral_yearresid,   yr_resid_mat_r)  +
                   post_r$b_logcoral_Intercept
rd_mat_r_cf     <- rs(post_r$b_logrd_logcoral, coral_mat_r_cf)        + post_r$b_logrd_Intercept
fleshy_mat_r_cf <- pmin(rs(post_r$b_logfleshy_logcoral, coral_mat_r_cf) + post_r$b_logfleshy_Intercept,
                        z_ceil_fleshy)
pmax_mat_r_cf   <- rs(post_r$bsp_logpmax_milogrd, rd_mat_r_cf)        + post_r$b_logpmax_Intercept
n_mat_r_cf      <- rs(post_r$bsp_Npercent_milogrd, rd_mat_r_cf)       + post_r$b_Npercent_Intercept

proj_tbl_r_cf <- imap_dfr(
  list(Temperature    = bt_lin(temp_mat_r_cf,   modelmean$temperature, modelsd$temperature),
       `Coral Cover`  = bt_log(coral_mat_r_cf,  modelmean$logcoral,    modelsd$logcoral),
       Respiration    = bt_log(rd_mat_r_cf,     modelmean$logrd,       modelsd$logrd),
       `Fleshy Algae` = bt_log(fleshy_mat_r_cf, modelmean$logfleshy,   modelsd$logfleshy),
       Pmax           = bt_log(pmax_mat_r_cf,   modelmean$logpmax,     modelsd$logpmax),
       `N Content`    = bt_log(n_mat_r_cf,
                               mean(log(Year_Averages$N_percent), na.rm = TRUE),
                               sd(log(Year_Averages$N_percent),   na.rm = TRUE))),
  function(mat, vname) {
    tibble(variable = vname, Year = rep(proj_years, each = n_r), value = as.vector(mat)) |>
      group_by(variable, Year) |>
      summarise(median = median(value),
                lo95 = quantile(value, .025), hi95 = quantile(value, .975),
                lo50 = quantile(value, .25),  hi50 = quantile(value, .75),
                .groups = "drop")
  })

```

## Copernicus CDS CMIP6 SST Projections to 2100

```{r cds-setup}
#| label: cds-setup
#| echo: true
#| eval: false

# ── Run once to store your CDS Personal Access Token securely ─────────────────
# 1. Log in at https://cds.climate.copernicus.eu/
# 2. Go to Profile (top-right) → Personal Access Tokens → copy your token
# 3. Paste it below and run this chunk once — it will not need to be run again.
library(ecmwfr)
wf_set_key(key = "YOUR-PERSONAL-ACCESS-TOKEN")
```

```{r cds-cmip6-sst-download}
#| label: cds-cmip6-sst-download
#| cache: true
#| echo: false
#| message: false

# ── Download CMIP6 sea surface temperature (tos) from Copernicus CDS ──────────
# Ten coupled ScenarioMIP models with tos available for all three SSPs.
# Monthly resolution, spatially subsetted to a 3° box around Moorea LTER Site 1
# (-17.48°S, -149.83°W). Four experiments: historical (1950-2014) +
# SSP1-2.6 / SSP2-4.5 / SSP5-8.5 (2015-2100).
# Note: CDS processes requests server-side; first run may take 30–90 minutes.

library(ecmwfr)

dir.create("data/cmip6_sst", recursive = TRUE, showWarnings = FALSE)

cmip6_models <- c(
  "access_cm2", "bcc_csm2_mr", "ec_earth3",
  "gfdl_esm4",  "ipsl_cm6a_lr", "miroc6",
  "mpi_esm1_2_lr", "mri_esm2_0", "noresm2_lm", "ukesm1_0_ll"
)

# Bounding box: N, W, S, E (degrees, -180 to 180)
moorea_area <- c(-16.0, -151.0, -19.0, -148.0)

all_months <- as.list(sprintf("%02d", 1:12))

build_sst_request <- function(experiment, years, target_file) {
  list(
    dataset_short_name  = "projections-cmip6",
    temporal_resolution = "monthly",
    experiment          = experiment,
    variable            = "sea_surface_temperature",
    model               = as.list(cmip6_models),
    year                = as.list(as.character(years)),
    month               = all_months,
    area                = moorea_area,
    target              = target_file
  )
}

requests <- list(
  build_sst_request("historical", 1950:2014, "tos_historical.zip"),
  build_sst_request("ssp1_2_6",  2015:2100, "tos_ssp126.zip"),
  build_sst_request("ssp2_4_5",  2015:2100, "tos_ssp245.zip"),
  build_sst_request("ssp5_8_5",  2015:2100, "tos_ssp585.zip")
)

wf_request_batch(requests, workers = 2, path = "data/cmip6_sst/")
```

```{r cds-cmip6-sst-process}
#| label: cds-cmip6-sst-process
#| cache: true
#| echo: false
#| message: false

# ── Read CDS NetCDF files, extract Moorea grid cell, bias-correct ─────────────
# Produces:
#   wb_proj_bc   — ensemble p10/median/p90 per scenario-year (used in propagation)
#   ann_cmip6_bc — per-model historical annual max SST (used for spread ribbon)

library(ncdf4)

sst_dir    <- "data/cmip6_sst"
moorea_lon <- -149.83
moorea_lat <- -17.48

# Extract model name from CMIP6 standard filename:
# <var>_<freq>_<model>_<exp>_<variant>_<grid>_<period>.nc
extract_model_name <- function(filepath) {
  parts <- strsplit(basename(filepath), "_")[[1]]
  if (length(parts) >= 3) parts[3] else "unknown"
}

read_tos_nc <- function(nc_path, experiment) {
  nc <- nc_open(nc_path)
  on.exit(nc_close(nc))

  model <- extract_model_name(nc_path)

  # Coordinates — handle both 1-D regular and 2-D curvilinear grids
  lon <- tryCatch(ncvar_get(nc, "lon"), error = \(e) ncvar_get(nc, "longitude"))
  lat <- tryCatch(ncvar_get(nc, "lat"), error = \(e) ncvar_get(nc, "latitude"))

  # Normalise 0–360 longitudes to -180–180
  lon_norm <- ifelse(lon > 180, lon - 360, lon)

  if (length(dim(lon)) == 2) {
    # Curvilinear grid
    dist  <- sqrt((lon_norm - moorea_lon)^2 + (lat - moorea_lat)^2)
    idx   <- which(dist == min(dist, na.rm = TRUE), arr.ind = TRUE)
    tos_raw <- ncvar_get(nc, "tos")[idx[1, 1], idx[1, 2], ]
  } else {
    i_lon   <- which.min(abs(lon_norm - moorea_lon))
    i_lat   <- which.min(abs(lat      - moorea_lat))
    tos_raw <- ncvar_get(nc, "tos")[i_lon, i_lat, ]
  }

  # Time → calendar dates
  time_raw   <- ncvar_get(nc, "time")
  time_units <- ncatt_get(nc, "time", "units")$value
  origin_str <- sub("^days since ", "", time_units)
  dates      <- as.Date(origin_str) + as.integer(round(time_raw))

  # Replace fill values with NA
  fv <- ncatt_get(nc, "tos", "_FillValue")$value
  if (length(fv) > 0) tos_raw[abs(tos_raw - fv) < abs(fv) * 1e-3] <- NA

  # Convert Kelvin → Celsius if needed
  u <- ncatt_get(nc, "tos", "units")$value
  if (length(u) > 0 && grepl("^K$|kelvin", u, ignore.case = TRUE))
    tos_raw <- tos_raw - 273.15

  n <- min(length(dates), length(tos_raw))
  tibble(model      = model,
         experiment = experiment,
         date       = dates[seq_len(n)],
         tos        = as.numeric(tos_raw[seq_len(n)]))
}

zip_meta <- tibble(
  zip_file   = file.path(sst_dir, c("tos_historical.zip", "tos_ssp126.zip",
                                     "tos_ssp245.zip",    "tos_ssp585.zip")),
  experiment = c("historical", "ssp1_2_6", "ssp2_4_5", "ssp5_8_5")
)

all_tos <- pmap_dfr(zip_meta, function(zip_file, experiment) {
  if (!file.exists(zip_file)) { warning("Missing: ", zip_file); return(NULL) }
  td <- file.path(sst_dir, experiment)
  dir.create(td, showWarnings = FALSE)
  unzip(zip_file, exdir = td, overwrite = FALSE)
  nc_files <- list.files(td, pattern = "\\.nc$", full.names = TRUE)
  map_dfr(nc_files, \(f) tryCatch(
    read_tos_nc(f, experiment),
    error = \(e) { warning(basename(f), ": ", e$message); NULL }
  ))
})

# Annual maximum SST per model-year
ann_tos <- all_tos |>
  mutate(year = year(date)) |>
  group_by(model, experiment, year) |>
  summarise(ann_max = max(tos, na.rm = TRUE), .groups = "drop") |>
  filter(is.finite(ann_max))

# ── Delta bias correction: anchor each model to observed MCR 2005-2025 mean ───
obs_mean_mcr_cds <- mean(Year_Averages$Max_temp[Year_Averages$Year %in% 2005:2025],
                         na.rm = TRUE)

bias_tbl_cds <- ann_tos |>
  filter(
    (experiment == "historical"                            & year %in% 2005:2014) |
    (experiment %in% c("ssp1_2_6","ssp2_4_5","ssp5_8_5") & year %in% 2015:2025)
  ) |>
  group_by(model) |>
  summarise(model_mean = mean(ann_max, na.rm = TRUE), .groups = "drop") |>
  mutate(bias = obs_mean_mcr_cds - model_mean)

ann_tos_bc <- ann_tos |>
  left_join(bias_tbl_cds, by = "model") |>
  mutate(ann_max_bc = ann_max + bias)

# ── Ensemble p10 / median / p90 per experiment-year ───────────────────────────
ens_stats <- function(data, exp_name) {
  data |>
    filter(experiment == exp_name) |>
    group_by(year) |>
    summarise(p10    = quantile(ann_max_bc, 0.10, na.rm = TRUE),
              median = quantile(ann_max_bc, 0.50, na.rm = TRUE),
              p90    = quantile(ann_max_bc, 0.90, na.rm = TRUE),
              .groups = "drop") |>
    mutate(experiment = exp_name)
}

ssp_ens  <- bind_rows(
  ens_stats(ann_tos_bc, "ssp1_2_6"),
  ens_stats(ann_tos_bc, "ssp2_4_5"),
  ens_stats(ann_tos_bc, "ssp5_8_5")
)
hist_ens <- ens_stats(ann_tos_bc, "historical")

# ── Build wb_proj_bc (structure expected by downstream propagation code) ───────
ssp_long <- ssp_ens |>
  pivot_longer(c(p10, median, p90), names_to = "percentile", values_to = "tasmax_bc") |>
  mutate(scenario = case_when(
    experiment == "ssp1_2_6" ~ "ssp126",
    experiment == "ssp2_4_5" ~ "ssp245",
    experiment == "ssp5_8_5" ~ "ssp585"
  )) |>
  select(year, scenario, percentile, tasmax_bc)

# Historical 2005-2014 expanded across all three scenarios for the overlap window
hist_long <- hist_ens |>
  filter(year %in% 2005:2014) |>
  pivot_longer(c(p10, median, p90), names_to = "percentile", values_to = "tasmax_bc") |>
  select(year, percentile, tasmax_bc) |>
  expand_grid(scenario = c("ssp126", "ssp245", "ssp585"))

wb_proj_bc <- bind_rows(hist_long, ssp_long) |>
  distinct(year, scenario, percentile, .keep_all = TRUE) |>
  arrange(scenario, percentile, year) |>
  filter(year %in% 2005:2100)

# ── ann_cmip6_bc: historical per-model series for the 2005-2014 spread ribbon ─
ann_cmip6_bc <- ann_tos_bc |>
  filter(experiment == "historical")
```

```{r wb-cmip6-propagate}
#| label: wb-cmip6-propagate

# ── Propagate each SSP scenario × CMIP6 percentile through the reduced SEM ───
# Approach: two-phase projection anchored to the 2025 OBSERVED ecosystem state.
#   Phase 1 (2005–2025): observed z-scores from semdata2 used directly where
#     available, overriding the SEM chain prediction column-by-column. This
#     ensures the historical lines track the actual data rather than the
#     amplified sequential prediction. 2005 (not in semdata2) uses SEM prediction.
#   Phase 2 (2026–2100): delta method. Only the CHANGE in temperature from the
#     scenario's CMIP6 2025 value is propagated through the SEM path coefficients,
#     starting from the 2025 observed ecosystem state. This preserves the real
#     2025 coral, fleshy algae, and metabolism values as the projection origin.

# ── Observed z-score lookup (built once, reused across all propagation calls) ──
obs_anchors <- semdata2 |>
  mutate(Year_actual = round(Year * modelsd$Year + modelmean$Year)) |>
  select(Year_actual, logcoral, logrd, logfleshy, logpmax, Npercent, yearresid)

propagate_wb_ssp <- function(scenario_name, pct = "median") {
  # ── CMIP6 temperatures, z-scored, 2005–2100 (96 values) ──────────────────
  temps_bc <- wb_proj_bc |>
    filter(scenario == scenario_name, percentile == pct) |>
    arrange(year) |>
    pull(tasmax_bc)

  temp_z <- (temps_bc - modelmean$temperature) / modelsd$temperature

  n_hist <- 21L; n_proj <- 75L

  # ── Phase 1: SEM chain (fallback values for years/variables missing obs) ──
  temp_z_hist <- temp_z[seq_len(n_hist)]
  tm_hist <- matrix(rep(temp_z_hist, each = n_r), nrow = n_r)

  yr_rv_hist <- sapply(2005:2025, function(y) {
    row <- yr_resid_lookup |> filter(Year == y)
    if (nrow(row) == 1L && !is.na(row$yr)) row$yr else 0
  })
  yrm_hist <- matrix(rep(yr_rv_hist, each = n_r), nrow = n_r)

  cm_hist <- pmin(rs(post_r$b_logcoral_temperature, tm_hist) +
                  rs(post_r$b_logcoral_yearresid,   yrm_hist) +
                  post_r$b_logcoral_Intercept,  z_ceil_coral)
  rm_hist <- rs(post_r$b_logrd_logcoral,          cm_hist) + post_r$b_logrd_Intercept
  fm_hist <- pmin(rs(post_r$b_logfleshy_logcoral, cm_hist) +
                  post_r$b_logfleshy_Intercept, z_ceil_fleshy)
  pm_hist <- rs(post_r$bsp_logpmax_milogrd,  rm_hist) + post_r$b_logpmax_Intercept
  nm_hist <- rs(post_r$bsp_Npercent_milogrd, rm_hist) + post_r$b_Npercent_Intercept

  # ── Override historical columns with observed z-scores where available ────
  obs_z <- function(y, col) {
    row <- obs_anchors |> filter(Year_actual == y)
    if (nrow(row) == 0L) return(NA_real_)
    v <- row[[col]][[1L]]
    if (is.na(v)) NA_real_ else v
  }

  for (j in 2:n_hist) {          # j = 1 → 2005 (not in semdata2, keep SEM)
    y <- 2004L + j
    oc <- obs_z(y, "logcoral");  if (!is.na(oc)) cm_hist[, j] <- oc
    or <- obs_z(y, "logrd");     if (!is.na(or)) rm_hist[, j] <- or
    of <- obs_z(y, "logfleshy"); if (!is.na(of)) fm_hist[, j] <- of
    op <- obs_z(y, "logpmax");   if (!is.na(op)) pm_hist[, j] <- op
    on <- obs_z(y, "Npercent");  if (!is.na(on)) nm_hist[, j] <- on
  }

  # ── 2025 observed anchor (scalar per variable) ────────────────────────────
  z25 <- obs_anchors |> filter(Year_actual == 2025L) |> slice(1)
  c25 <- z25$logcoral
  r25 <- z25$logrd
  f25 <- z25$logfleshy
  p25 <- z25$logpmax
  # N_percent NA in 2025: impute from observed Rd via SEM median
  n25 <- median(post_r$bsp_Npercent_milogrd * r25 + post_r$b_Npercent_Intercept)

  # ── Phase 2 (2026–2100): delta from 2025 observed anchor ─────────────────
  temp_z_2025 <- temp_z[n_hist]
  delta_temp  <- temp_z[(n_hist + 1L):(n_hist + n_proj)] - temp_z_2025
  dtm         <- matrix(rep(delta_temp, each = n_r), nrow = n_r)  # n_r × n_proj

  cm_proj <- pmin(c25 + rs(post_r$b_logcoral_temperature, dtm), z_ceil_coral)
  dcm     <- cm_proj - c25
  rm_proj <- r25 + rs(post_r$b_logrd_logcoral,          dcm)
  fm_proj <- pmin(f25 + rs(post_r$b_logfleshy_logcoral, dcm), z_ceil_fleshy)
  drm     <- rm_proj - r25
  pm_proj <- p25 + rs(post_r$bsp_logpmax_milogrd,  drm)
  nm_proj <- n25 + rs(post_r$bsp_Npercent_milogrd, drm)

  # ── Combine & back-transform ──────────────────────────────────────────────
  tm <- matrix(rep(temp_z, each = n_r), nrow = n_r)
  cm <- cbind(cm_hist, cm_proj);  rm <- cbind(rm_hist, rm_proj)
  fm <- cbind(fm_hist, fm_proj);  pm <- cbind(pm_hist, pm_proj)
  nm <- cbind(nm_hist, nm_proj)

  imap_dfr(
    list(Temperature    = bt_lin(tm, modelmean$temperature, modelsd$temperature),
         `Coral Cover`  = bt_log(cm, modelmean$logcoral,    modelsd$logcoral),
         Respiration    = bt_log(rm, modelmean$logrd,       modelsd$logrd),
         `Fleshy Algae` = bt_log(fm, modelmean$logfleshy,   modelsd$logfleshy),
         Pmax           = bt_log(pm, modelmean$logpmax,     modelsd$logpmax),
         `N Content`    = bt_log(nm,
                                  mean(log(Year_Averages$N_percent), na.rm = TRUE),
                                  sd(log(Year_Averages$N_percent),   na.rm = TRUE))),
    function(mat, vname) {
      tibble(scenario = scenario_name, percentile = pct, variable = vname,
             year = rep(2005:2100, each = n_r), value = as.vector(mat)) |>
        group_by(scenario, percentile, variable, year) |>
        summarise(median = median(value, na.rm = TRUE),
                  lo95   = quantile(value, .025, na.rm = TRUE),
                  hi95   = quantile(value, .975, na.rm = TRUE),
                  lo50   = quantile(value, .25,  na.rm = TRUE),
                  hi50   = quantile(value, .75,  na.rm = TRUE),
                  .groups = "drop")
    })
}

wb_all_combos <- expand_grid(
  scenario   = c("ssp126", "ssp245", "ssp585"),
  percentile = c("p10", "median", "p90")
)
wb_proj_tbl <- bind_rows(
  mapply(propagate_wb_ssp,
         scenario_name = wb_all_combos$scenario,
         pct           = wb_all_combos$percentile,
         SIMPLIFY      = FALSE)
)

# ── Tidy for plotting ─────────────────────────────────────────────────────────
wb_sem_ci_band <- wb_proj_tbl |>
  filter(percentile == "median") |>
  select(scenario, variable, year, central = median, lo_sem = lo95, hi_sem = hi95)

wb_clim_band <- wb_proj_tbl |>
  filter(percentile %in% c("p10", "p90")) |>
  select(scenario, variable, year, percentile, clim_med = median) |>
  pivot_wider(names_from = percentile, values_from = clim_med) |>
  rename(lo_clim = p10, hi_clim = p90)

wb_ssp_labels  <- c(ssp126 = "SSP1-2.6", ssp245 = "SSP2-4.5", ssp585 = "SSP5-8.5")
wb_ssp_colours <- c(ssp126 = "#2196F3",  ssp245 = "#FF9800",   ssp585 = "#E53935")
wb_var_order   <- c("Temperature","Coral Cover","Respiration",
                    "Fleshy Algae","Pmax","N Content")
wb_var_units   <- c(Temperature    = "\u00b0C",
                    `Coral Cover`  = "% cover",
                    Respiration    = "mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9",
                    `Fleshy Algae` = "% cover",
                    Pmax           = "mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9",
                    `N Content`    = "% N")

wb_plot_tbl <- wb_sem_ci_band |>
  left_join(wb_clim_band, by = c("scenario", "variable", "year")) |>
  mutate(variable       = factor(variable, levels = wb_var_order),
         scenario_label = wb_ssp_labels[scenario])

# ── Historical temperature spread: inter-model p10/p90 across CDS CMIP6 ensemble ─
# For 2005-2014 (historical period), the SSP ensemble spread is not defined;
# use the CMIP6 historical run inter-model spread (p10–p90 across 10 bias-corrected
# CDS CMIP6 models) to fill the outer uncertainty ribbon for the Temperature panel.
om_hist_spread <- ann_cmip6_bc |>
  filter(year %in% 2005:2014) |>
  group_by(year) |>
  summarise(lo_clim_om = quantile(ann_max_bc, 0.10),
            hi_clim_om = quantile(ann_max_bc, 0.90),
            .groups = "drop")

wb_plot_tbl <- wb_plot_tbl |>
  left_join(om_hist_spread, by = "year") |>
  mutate(
    lo_clim = if_else(variable == "Temperature" & year %in% 2005:2014,
                      lo_clim_om, lo_clim),
    hi_clim = if_else(variable == "Temperature" & year %in% 2005:2014,
                      hi_clim_om, hi_clim)
  ) |>
  select(-lo_clim_om, -hi_clim_om)
```

```{r fig-wb-cmip6}
#| label: fig-wb-cmip6
#| fig-cap: "**CMIP6 SST multi-scenario projections for MCR LTER Site 1 backreef, 2005–2100.** Annual maximum sea surface temperature (*tos*) was obtained from ten coupled CMIP6 ScenarioMIP models (ACCESS-CM2, BCC-CSM2-MR, EC-Earth3, GFDL-ESM4, IPSL-CM6A-LR, MIROC6, MPI-ESM1-2-LR, MRI-ESM2-0, NorESM2-LM, UKESM1-0-LL) downloaded from the Copernicus Climate Data Store for the Moorea grid cell (~17.5°S, 149.8°W) under three shared socioeconomic pathways: SSP1-2.6 (blue), SSP2-4.5 (orange), and SSP5-8.5 (red), for 2015–2100, stitched to the historical CMIP6 run (1950–2014). Each model was delta bias-corrected to the 2005–2025 MCR observational mean (additive shift per model) before computing the ensemble p10, median, and p90. Bias-corrected temperatures were z-scored and propagated through the fitted reduced Bayesian SEM. **Historical period (2005–2025, unshaded):** lines show observed annual ecosystem state, derived directly from the MCR LTER dataset; open circles mark the same values for emphasis. **Projection period (2026–2100, grey shading):** lines show the SEM posterior median; inner ribbons show the 95% credible interval (3,000 posterior draws, CMIP6 ensemble median SST); outer ribbons show the CMIP6 multi-model p10–p90 climate spread at the SEM posterior median, separating parameter uncertainty from forcing uncertainty. The temperature panel retains both ribbon types throughout the full record. Fleshy algae y-axis capped at 100% cover (the log-linear SEM projects >100% under SSP5-8.5, which is biologically impossible and reflects model extrapolation beyond the observed range)."
#| fig-width: 9
#| fig-height: 8

wb_y_limits <- list(
  Temperature    = NULL,
  `Coral Cover`  = c(0, NA),
  Respiration    = c(0, NA),
  `Fleshy Algae` = c(0, 100),
  Pmax           = c(60, NA),
  `N Content`    = c(0.3, NA)
)

make_wb_panel <- function(vname) {
  d_v   <- wb_plot_tbl |> filter(variable == vname)
  obs_v <- Year_Averages |>
    select(year = Year,
           Temperature    = Max_temp,
           `Coral Cover`  = mean_coral,
           Respiration    = Rd,
           `Fleshy Algae` = mean_fleshy,
           Pmax           = Pmax,
           `N Content`    = N_percent) |>
    pivot_longer(-year, names_to = "variable", values_to = "obs") |>
    filter(variable == vname)
  unit <- wb_var_units[[vname]]
  ylim <- wb_y_limits[[vname]]

  # Ribbons only in the projection period (2026+), where uncertainty is meaningful.
  # For temperature, the outer (climate) ribbon also shows the CMIP6 model spread
  # in the historical period (outer ribbon is valid there).
  # For other variables, ribbons in 2005-2025 are artefacts: 2005 uses the SEM
  # chain (very wide), observed years collapse to zero width, and the outer ribbon
  # is inverted for variables with a negative temperature relationship.
  d_ribbon       <- d_v |> filter(year > 2025)
  d_clim_ribbon  <- if (vname == "Temperature") d_v else d_ribbon

  p <- ggplot(d_v, aes(x = year, colour = scenario, fill = scenario)) +
    annotate("rect", xmin = 2025.5, xmax = 2100, ymin = -Inf, ymax = Inf,
             fill = "grey92", alpha = .6) +
    geom_ribbon(data = d_clim_ribbon,
                aes(ymin = pmin(lo_clim, hi_clim), ymax = pmax(lo_clim, hi_clim)),
                alpha = 0.10, colour = NA) +
    geom_ribbon(data = d_ribbon,
                aes(ymin = lo_sem, ymax = hi_sem),
                alpha = 0.20, colour = NA) +
    geom_line(aes(y = central), linewidth = 0.7) +
    geom_point(data = obs_v, aes(x = year, y = obs),
               colour = "black", size = 1.5, shape = 21, fill = "white",
               inherit.aes = FALSE) +
    geom_vline(xintercept = 2025.5, linetype = "dashed",
               colour = "grey50", linewidth = .4) +
    scale_colour_manual(values = wb_ssp_colours, labels = wb_ssp_labels, name = "Scenario") +
    scale_fill_manual(  values = wb_ssp_colours, labels = wb_ssp_labels, name = "Scenario") +
    scale_x_continuous(breaks = seq(2010, 2100, 20)) +
    labs(x = NULL, y = unit, title = vname) +
    theme_bw(base_size = 9) +
    theme(legend.position = "none",
          plot.title = element_text(size = 9, face = "bold"),
          panel.grid.minor = element_blank())
  if (!is.null(ylim)) p <- p + coord_cartesian(ylim = ylim)
  if (vname == "Fleshy Algae")
    p <- p + annotate("text", x = 2060, y = 96,
                      label = "100% ceiling", size = 2.3, colour = "grey40",
                      hjust = 0.5, fontface = "italic")
  p
}

wb_panels <- lapply(wb_var_order, make_wb_panel)

wb_legend_plot <- ggplot(
  wb_plot_tbl |> filter(variable == "Temperature"),
  aes(x = year, y = central, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo_sem, ymax = hi_sem), alpha = .25, colour = NA) +
  geom_line(linewidth = .9) +
  scale_colour_manual(values = wb_ssp_colours, labels = wb_ssp_labels,
                      name = "CMIP6 Scenario") +
  scale_fill_manual(  values = wb_ssp_colours, labels = wb_ssp_labels,
                      name = "CMIP6 Scenario") +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.5))) +
  theme_void() +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.title = element_text(size = 8.5, face = "bold"),
        legend.text  = element_text(size = 8))
wb_leg <- cowplot::get_legend(wb_legend_plot)

wb_main <- wrap_plots(wb_panels, ncol = 2) & theme(legend.position = "none")
cowplot::plot_grid(wb_main, wb_leg, ncol = 2, rel_widths = c(1, 0.12))
```

```{r tbl-wb-cmip6}
#| label: tbl-wb-cmip6
#| tbl-cap: "CMIP6 SST ensemble projections at 2050, 2075, and 2100 under three SSP scenarios (reduced SEM, CMIP6 ensemble median SST). Values are posterior median (95% credible interval). Temperature CIs are negligible because the SEM does not add uncertainty to the input forcing. Fleshy Algae values >100% under SSP5-8.5 reflect log-linear model extrapolation beyond biological limits."

wb_tbl_out <- bind_rows(lapply(c(2050, 2075, 2100), function(yr) {
  wb_proj_tbl |>
    filter(percentile == "median", year == yr) |>
    mutate(scenario_label = wb_ssp_labels[scenario],
           label = sprintf("%.1f (%.1f\u2013%.1f)", median, lo95, hi95)) |>
    select(variable, scenario_label, label) |>
    pivot_wider(names_from = scenario_label, values_from = label) |>
    mutate(Year = yr) |>
    relocate(Year, variable)
})) |>
  arrange(variable, Year) |>
  rename(Variable = variable)

knitr::kable(wb_tbl_out, align = c("r","l","c","c","c"))
```

# Results & Interpretation

## Results

### Community composition and environmental change

The MCR LTER Site 1 backreef underwent substantial reorganisation over 2006–2025 (Fig. @fig-benthic). Coral cover declined from 31.0% to 6.8% (−78%), while fleshy macroalgae increased from 5.5% to 14.7% (+167%). Bayesian temporal trend models confirmed directional change across all major community variables (Fig. @fig-temporal): coral cover showed the strongest rate of decline (β = −0.14 SD yr⁻¹, R² = 0.70), corallivore fish biomass declined in parallel (β = −0.13 SD yr⁻¹, R² = 0.52), and herbivore biomass increased (β = +0.15 SD yr⁻¹, R² = 0.66). Annual maximum temperature increased modestly but consistently over the study period (β = +0.07 SD yr⁻¹, R² = 0.19), corresponding to approximately +0.65°C over 20 years. Algal tissue nitrogen content showed the strongest temporal signal among biogeochemical variables (β = −0.17 SD yr⁻¹, R² = 0.78).

### Ecosystem metabolism

Ecosystem respiration (Rd) declined markedly from a mean of 57 mmol O₂ m⁻² hr⁻¹ in 2008–2010 to 29 mmol O₂ m⁻² hr⁻¹ in 2021–2025 (−49%; β = −0.14 SD yr⁻¹, R² = 0.50). Maximum photosynthetic capacity (Pmax) also declined, from 124 to 105 mmol O₂ m⁻² hr⁻¹ between early and late periods (β = −0.08 SD yr⁻¹, R² = 0.19), though with weaker temporal signal. Net ecosystem production showed no significant directional trend (β = +0.08 SD yr⁻¹, R² = 0.20). These parameters were extracted from a Bayesian photosynthesis-irradiance (PI) curve model fit across 2,928 production measurements, with all chain convergence diagnostics satisfactory (Rhat < 1.01, ESS > 1,000 for all parameters).

### Causal pathways through the reef ecosystem

A Bayesian SEM encompassing eight response variables and 12 causal pathways revealed coral cover as the central mediator of ecosystem change (Fig. @fig-sem-dag; path coefficients: Fig. @fig-sem-paths; conditional effects: Fig. @fig-sem-conditional). Year was positively associated with temperature (β = 0.47, 95% CI 0.12–0.82), and temperature negatively drove coral cover (β = −0.55, 95% CI −0.83 to −0.28). A non-thermal temporal component — the residual of calendar year after removing the temperature trend — also credibly predicted coral cover (β = −0.67, 95% CI −0.93 to −0.41), capturing secular decline in coral cover unexplained by warming alone. Coral cover in turn was the strongest predictor of ecosystem respiration (β = 1.08, 95% CI 0.64–1.53), fleshy macroalgae (β = −0.91, 95% CI −1.13 to −0.70), and corallivore fish biomass (β = 0.65, 95% CI 0.26–1.02). Herbivore fish biomass was positively predicted by fleshy macroalgae cover (β = 0.71, 95% CI 0.37–1.06), consistent with bottom-up trophic control. Ecosystem respiration in turn predicted algal nitrogen content (β = 0.62, 95% CI 0.24–1.01) and Pmax (β = 0.82, 95% CI 0.03–1.62). Three tested pathways — direct temperature effects on respiration (β = 0.22, 95% CI −0.22 to 0.62) and Pmax (β = −0.16, 95% CI −0.78 to 0.45), and a direct coral effect on Pmax (β = −0.35, 95% CI −1.40 to 0.71) — had credible intervals that included zero. Bayesian R² was 0.22 for temperature and 0.71 for coral cover (the year-residual predictor accounted for the large improvement from the prior R² of 0.29).

### Temperature effects are predominantly indirect

Decomposing total temperature effects into direct and coral-mediated indirect pathways revealed that coral decline was the dominant transmission mechanism for most ecosystem responses (Fig. @fig-temp-indirect). The indirect effect of temperature on ecosystem respiration, operating through coral loss, was substantial and directionally credible (β_indirect = −0.58, 95% CI −1.01 to −0.25), while the direct thermal effect was inconclusive (β_direct = 0.22, 95% CI −0.22 to 0.62), and the total effect did not exclude zero (β_total = −0.38, 95% CI −0.85 to 0.08). For Pmax, direct and indirect pathways partially cancelled and neither was significant (β_total = 0.03, 95% CI −0.54 to 0.64). All indirect temperature effects on fleshy macroalgae (β = +0.50, 95% CI 0.25–0.79), corallivore fish (β = −0.34, 95% CI −0.66 to −0.11), herbivore fish (β = +0.35, 95% CI 0.14–0.66), and algal nitrogen content (β = −0.35, 95% CI −0.76 to −0.10) were mediated entirely through coral cover and had credible intervals excluding zero.

### Ecosystem projections under CMIP6 climate scenarios

Extending the SEM causal framework to 2100 using Copernicus CDS CMIP6 SST forcing under three SSP emission scenarios reveals strong dependence of reef ecosystem state on the trajectory of global emissions (Fig. @fig-wb-cmip6, Table @tbl-wb-cmip6).

**Near-term (to 2050).** Scenarios diverge only modestly before mid-century, consistent with similar temperature trajectories across SSPs prior to ~2040. Coral cover at 2050 is projected at 5.1% under SSP1-2.6 versus 4.0% under SSP5-8.5 — a relatively small contrast reflecting that near-term warming is largely committed regardless of emission pathway. Ecosystem respiration, fleshy macroalgae, Pmax, and N content show similarly narrow scenario separation at 2050.

**Mid-century (2050–2075).** Trajectories diverge sharply as scenario-dependent warming accumulates. Under SSP5-8.5, temperatures approach ~31°C by 2075, driving coral cover below 2% and ecosystem respiration below 22 mmol O₂ m⁻² hr⁻¹. Under SSP1-2.6, temperatures stabilise near 30–30.5°C following the emissions peak, and coral cover remains above 5%. SSP2-4.5 follows an intermediate trajectory, with coral declining slowly toward 4%.

**End of century (2100).** The scenario contrast at 2100 is striking. Under SSP1-2.6, coral cover stabilises at 5.4% (posterior 95% CI: 4.8–6.1%), broadly near its 2025 level, while ecosystem respiration remains at 26.3 mmol O₂ m⁻² hr⁻¹ and fleshy macroalgae is held around 19% cover — a reef that is structurally degraded relative to 2005, but functionally persistent. Under SSP5-8.5, coral cover collapses to 0.9% (95% CI: 0.3–2.4%) — a ~6-fold difference relative to SSP1-2.6. Ecosystem respiration falls to 13.8 mmol O₂ m⁻² hr⁻¹ (−52% relative to 2025 levels), and the fleshy macroalgae model extrapolates beyond the 100% biological ceiling, indicating an algae-dominated phase-shifted state. Algal tissue nitrogen content, which tracks the respiration pathway through the SEM, declines from 0.57% in 2025 to 0.42% by 2100 under SSP5-8.5 (−26%), while remaining near 0.56% under SSP1-2.6 as metabolic rates stabilise with temperature. SSP2-4.5 projects an intermediate outcome: coral ~3.2% by 2100, respiration ~22 mmol O₂ m⁻² hr⁻¹, fleshy algae ~36%, and N content ~0.52%, illustrating that intermediate emissions postpone but do not prevent continued reef degradation.

The ~2040–2060 window represents the critical divergence point, after which emission pathways commit the system to qualitatively different futures. The ~6-fold contrast in coral cover at 2100 between SSP1-2.6 and SSP5-8.5 quantifies the ecological leverage of global emissions mitigation for reef persistence. Two caveats apply to these projections. First, the coupled-model SST forcing is at CMIP6 resolution (~1°); the delta bias correction captures the ensemble warming rate but not local oceanographic variability (e.g., ENSO anomalies, upwelling around Moorea). Second, the log-linear SEM extrapolates fleshy macroalgae beyond biological limits under SSP5-8.5, which should be interpreted as indicating a phase-shifted, algae-dominated state rather than literal cover percentages.

# Discussion



- **One Tree Island, Great Barrier Reef** (Kinsey 1985; Barnes 1983): 7-year study (1975-1982) using slack-water technique, documented mean gross production of 180-250 mmol O₂ m⁻² hr⁻¹ and net production near metabolic balance (slight net autotrophy). No long-term trend analysis due to relatively stable coral community during study period. Established baseline metabolic rates for healthy Indo-Pacific reef flats.

- **Kaneohe Bay, Hawaii** (Shamberger et al. 2011; Drupp et al. 2013): 5-year lagoon-scale study (2008-2013) combining in situ sensors with hydrodynamic modeling, reported net heterotrophy (-20 to -40 mmol O₂ m⁻² hr⁻¹) in a degraded, eutrophic system. Demonstrated metabolic depression relative to historical baselines, attributed to sewage diversion recovery and persistent macroalgal dominance. Our MCR backreef shows similar net production decline but from initially higher autotrophy.

- **Heron Island, Great Barrier Reef** (McMahon et al. 2013; Santos et al. 2011): 3-year reef flat study (2009-2012) using Eulerian flow-through approach, documented seasonal cycles in metabolism with mean gross production ~150 mmol O₂ m⁻² hr⁻¹. No significant interannual trends detected, though study period preceded major bleaching events. Shorter temporal extent limits comparison to multi-decadal dynamics.

- **Lady Elliot Island, Great Barrier Reef** (Shaw et al. 2015, 2018): 7-year record (2010-2017) using CO₂ and O₂ sensors, reported declining net community calcification (-50% over study period) correlated with bleaching events, but primary production metrics showed high seasonal/interannual variability without clear trends. Focused primarily on calcification rather than photosynthesis-respiration balance.

- **Bermuda reef platforms** (Yeakel et al. 2015; Courtney et al. 2016, 2018): Decadal-scale calcification records (1996-2015) from multiple platforms showing 40-50% decline in net calcification, strongly correlated with ocean acidification and warming. Limited concurrent primary production measurements, though inferred metabolic stress from calcification:production ratios.

- **Puerto Morelos, Mexican Caribbean** (Salinas-de-León et al. 2018): 4-year study (2012-2016) documenting metabolism across reef zones, found forereef remained autotrophic (NP ~20-40 mmol O₂ m⁻² hr⁻¹) while backreef shifted toward heterotrophy following Hurricane Patricia. Demonstrated disturbance impacts on metabolism but limited baseline pre-disturbance data.

**Comparison to MCR backreef (this study)**: Our 20-year continuous record (2005-2025) represents one of the longest integrated metabolism time series globally for any coral reef habitat. Unlike most prior studies focused on calcification or snapshot metabolism measurements, we track simultaneous changes in gross production (via Pmax), ecosystem respiration (Rd), and net production alongside community composition and biogeochemistry.

```{r comparative-table}
#| label: tbl-comparative-metabolism
#| tbl-cap: "Comparative reef metabolism measurements from long-term studies"

# Create comparative table
comparative_metabolism <- tibble(
  Location = c("One Tree Island, GBR", "Kaneohe Bay, Hawaii", "Heron Island, GBR", 
               "Lady Elliot Island, GBR", "Puerto Morelos, Mexico", "MCR LTER Site 1 (This Study)"),
  Years = c("1975-1982", "2008-2013", "2009-2012", "2010-2017", "2012-2016", "2005-2025"),
  Duration = c("7 years", "5 years", "3 years", "7 years", "4 years", "20 years"),
  `Gross Production` = c("180-250", "Not reported", "~150", "Not reported", "Not reported", 
                        sprintf("%.0f-%.0f", 
                                min(Year_Averages$Pmax, na.rm = TRUE),
                                max(Year_Averages$Pmax, na.rm = TRUE))),
  `Net Production` = c("~0 (balanced)", "-20 to -40", "Variable", "Not primary focus", 
                      "20-40 (forereef)", 
                      sprintf("%.0f to %.0f",
                              min(Year_Averages$NP_mean, na.rm = TRUE),
                              max(Year_Averages$NP_mean, na.rm = TRUE))),
  `Metabolic State` = c("Balanced", "Heterotrophic", "Balanced/Variable", "Declining calcif.", 
                       "Autotrophic", "Autotroph→Balanced"),
  `Trend Detected` = c("No (stable)", "Yes (degraded)", "No (short)", "Yes (calcif.)", 
                      "Yes (disturbance)", "Yes (strong decline)")
)

knitr::kable(comparative_metabolism, 
             caption = "Comparative reef ecosystem metabolism from long-term studies. All gross and net production values in mmol O₂ m⁻² hr⁻¹. MCR LTER represents the longest continuous integrated metabolism record.",
             align = "lcccccc")
```

**Table Interpretation**: The MCR backreef shows gross production values (Pmax range `r sprintf("%.0f-%.0f", min(Year_Averages$Pmax, na.rm = TRUE), max(Year_Averages$Pmax, na.rm = TRUE))` mmol O₂ m⁻² hr⁻¹) comparable to other Indo-Pacific reefs like One Tree Island and Heron Island in their healthy state, but with a clear declining trajectory. Unlike the stable conditions at One Tree Island (1975-1982) or the acute degradation at Kaneohe Bay (eutrophication-driven), MCR represents a gradual climate-driven transition captured over an unprecedented 20-year period. The shift from net autotrophy toward metabolic balance parallels trends observed globally but with much finer temporal resolution.

Key distinctions from previous work:
1. **Temporal extent**: 20 years spans multiple disturbance events (bleaching, cyclones, crown-of-thorns outbreaks), enabling trend detection beyond interannual variability
2. **Mechanistic parameters**: Extraction of light-independent Pmax and Rd from PI curves provides physiological indices not captured by ambient flux measurements alone
3. **Integration**: Concurrent benthic surveys, fish censuses, and nutrient monitoring enable causal pathway analysis rather than correlative observations
4. **Directional trends**: Clear declines in both Pmax (~30% reduction) and Rd (~40% reduction) parallel coral loss, contrasting with stable or cyclical patterns in shorter time series from less impacted reefs
5. **Metabolic mode shift**: Initial net autotrophy (NP ~10-20 mmol O₂ m⁻² hr⁻¹) transitioning toward metabolic balance, similar magnitude to Kaneohe Bay degradation but from different starting state and drivers (temperature vs. eutrophication)

Our results suggest that even reefs maintaining some coral cover experience functional degradation - metabolic capacity declines may precede dramatic compositional collapse, providing early warning of ecosystem dysfunction. The MCR backreef trajectory mirrors calcification declines observed in Bermuda and Lady Elliot Island but extends to the full production-respiration system, revealing how warming disrupts multiple metabolic processes simultaneously through coral loss.

## Coral as an Ecological Hub: Temperature Acts Through Structure

The central finding of this analysis is that two decades of ocean warming have degraded reef ecosystem function primarily by dismantling coral structure, not by directly suppressing the physiology of individual organisms. Across every ecosystem variable we examined — respiration, photosynthesis, herbivory, corallivory, and nutrient cycling — the indirect temperature effect mediated through coral decline was either larger than, or the only credible component of, the total thermal signal. Direct temperature effects on ecosystem respiration (β = 0.22, 95% CI −0.21 to 0.62) and maximum photosynthetic capacity (β = −0.16, 95% CI −0.79 to 0.46) were both inconclusive, while their indirect counterparts (−0.58 and −0.19 respectively) were directionally credible. This pattern is not a statistical artefact: it reflects an ecological reality in which coral colonies function as keystone structures rather than just organisms. Coral provides physical framework, photosynthetic biomass, associated heterotrophic microbiomes, and three-dimensional habitat simultaneously; warming-induced bleaching and mortality therefore cascades across all of these pathways at once.

This contrasts sharply with the more commonly invoked direct warming hypothesis, in which ocean temperature is expected to stimulate microbial and heterotrophic metabolism (as in terrestrial systems) or suppress photosynthesis through bleaching of individual symbionts. Our data do not rule out those mechanisms acting at sub-annual timescales, but at the decadal scale of community reorganisation, structural loss appears to be the dominant transmission pathway. Ecosystem respiration declined 49% over 20 years not because metabolic rate constants fell, but because the coral-associated heterotrophic biomass that drives reef respiration — corals, their endosymbionts, and associated microbial communities — was removed. The mechanism is closer to deforestation than to metabolic inhibition.

## Ecosystem De-development and Metabolic Integration

The metabolic consequences of coral loss can be understood through Odum's (1969) concept of ecosystem development. Mature ecosystems are characterised by high biomass, low net production, high respiration, and tight internal cycling of nutrients — a configuration in which most fixed carbon is rapidly respired and recycled within the system rather than exported. Coral reefs are canonical examples of this functional type: extremely high gross production and respiration in oligotrophic water, sustained by tight coral-zooxanthellae-microbiome recycling. Our 20-year record documents a directional departure from this configuration. Both Pmax and Rd declined, but respiration showed a stronger and more consistent temporal signal (R² = 0.50 vs 0.19 for Pmax), and the SEM revealed that coral cover is a stronger predictor of respiration (β = 1.08) than of photosynthesis (β = −0.35, 95% CI crosses zero). The result is a gradual shift toward lower total metabolic activity rather than a simple shift in autotrophic–heterotrophic balance. In Odum's framework, this represents ecosystem retrogression: a trajectory from high-complexity, low-net-production towards lower biomass, lower respiratory capacity, and looser nutrient coupling.

The respiration–nitrogen pathway (β = 0.62, 95% CI 0.24–1.01) provides specific mechanistic support for this interpretation. Reduced ecosystem respiration — reflecting loss of coral-associated heterotrophic activity — correlates with lower algal tissue nitrogen content, consistent with reduced nutrient regeneration from organic matter decomposition in oligotrophic backreef water. Coral reefs are well known for efficient internal nitrogen cycling, and our data suggest that as the heterotrophic component of that cycle is reduced, nitrogen availability declines even for the macroalgae that are competitively replacing coral. This feedback — reduced respiration limiting the productivity of the organisms benefiting from coral decline — may partly explain why macroalgal expansion in oligotrophic Pacific reefs tends to be less extreme than on eutrophied Caribbean reefs.

## Indirect Effects Along Trophic Pathways

The two causal chains through community composition — Coral → Fleshy Algae → Herbivores, and Coral → Corallivores — show a coherent but ecologically nuanced response to warming. Corallivore fish biomass declined in parallel with coral cover (β = 0.65, 95% CI 0.25–1.04), as expected for a guild with obligate dietary dependence on live coral. Herbivore fish biomass, by contrast, increased over the same period (β_indirect temperature = +0.34, 95% CI 0.06–0.77), driven by expanding macroalgal food resources. At first glance this might seem compensatory, but the two responses operate through distinct pathways and the herbivore increase does not prevent macroalgal expansion — instead, the macroalgae–herbivore relationship represents bottom-up resource tracking rather than top-down regulation of algal abundance. Fleshy macroalgae expanded despite (and indeed facilitated) herbivore increases, suggesting that herbivory alone is insufficient to prevent phase-shift dynamics when coral structural complexity is lost.

These trophic dynamics have practical management implications. The consistent dominance of indirect (coral-mediated) over direct temperature effects across all six ecosystem variables implies that interventions targeting coral — whether through bleaching mitigation, shade structures, or active restoration — should in principle intercept the cascades before they propagate to fish assemblages, metabolism, and nutrient cycling. Local management of herbivore populations (e.g., controlling sea urchin or fish harvest) may be insufficient on its own if the primary leverage point is further upstream in the causal chain. Conversely, the fact that multiple ecosystem properties are coupled through a single hub suggests that successful coral recovery could trigger broad functional recovery, rather than requiring separate management of each endpoint.

## Comparison with Global Reef Trajectories

The MCR LTER Site 1 backreef trajectory is consistent with, but mechanistically more specific than, the global picture of reef decline. Widespread reductions in net community calcification (e.g., Albright et al. 2018; DeCarlo et al. 2017) and increasing heterotrophy (e.g., Kaneohe Bay; Shamberger et al. 2011) have been documented across ocean basins, but few studies have linked these to structural community change through an explicit causal framework. Our SEM results support a structural interpretation: metabolic decline tracks coral loss with a strong, credible path coefficient, whereas direct temperature effects are ambiguous. This suggests that the mechanistic driver of metabolic change at MCR is more closely analogous to the deforestation narrative than to the thermal inhibition narrative, with implications for projections: reef metabolism may recover if and when coral cover recovers, even if temperatures remain elevated.

The CMIP6 SSP scenario projections extend this finding forward in time. The ~6-fold contrast in coral cover at 2100 between SSP1-2.6 (~5%) and SSP5-8.5 (~1%) illustrates the degree to which emission trajectory controls long-term reef outcomes — a result consistent with global projections of mass coral bleaching frequency under high-emissions scenarios (Hoegh-Guldberg et al. 2018; Hughes et al. 2017). Critically, the scenarios remain indistinguishable in their reef consequences until approximately 2040–2050, implying that the next two decades of warming are largely committed regardless of mitigation effort. The divergence thereafter, however, highlights that aggressive emissions reduction remains the most effective lever for preserving reef function at the centennial timescale. The SSP1-2.6 projection — in which MCR coral cover stabilises near its 2025 level rather than collapsing further — represents the ecosystem consequence of achieving the Paris Agreement ~1.5–2°C target and provides an ecologically grounded argument for that policy goal.

## Limitations and Future Directions

Several constraints bound the current analysis. First, this is a single-site study, and the path structure of the SEM reflects the particular ecological context of an oligotrophic Society Islands backreef. Whether the same causal architecture — and in particular the dominance of indirect temperature effects — holds at other sites with different nutrient regimes, disturbance histories, or species assemblages is an open empirical question. Second, the Bayesian SEM assumes linearity in all path coefficients on the z-score scale; potential nonlinearities, threshold behaviour, or lagged effects (e.g., multi-year bleaching recovery) are not captured. Third, the nutrient imputation model for water column chemistry (R² = 0.68) introduces uncertainty into the respiration–nitrogen pathway that is not fully propagated through the effect decomposition. Direct water column measurements, currently unavailable after 2018, would strengthen this component of the analysis.

Regarding the CMIP6 scenario projections specifically, three additional caveats apply. The coupled-model SST forcing is location-specific but at CMIP6 resolution (~1°), and cannot capture local oceanographic variability such as ENSO-driven anomalies, wave-driven upwelling, or fine-scale circulation around Moorea; the delta bias correction anchors the ensemble mean to the observed record but does not constrain interannual variance. The log-linear SEM extrapolates fleshy macroalgae beyond the 100% biological ceiling under SSP5-8.5, an artefact of model extrapolation beyond the observed range that should be interpreted qualitatively as indicating a phase-shifted, algae-dominated state. Finally, SEM path coefficients are assumed stationary across the projection window; adaptive responses by coral populations, non-linear bleaching thresholds, and potential acclimatisation dynamics are not represented.

Future directions include cross-site SEM comparison across MCR LTER sites (forereef, fringing reef, and backreef habitats differ substantially in hydrodynamics and community structure), experimental manipulation of herbivore density to test the causal role of the algae–herbivore pathway, and extension of the time series to capture post-bleaching dynamics in the late 2020s and to update projection baselines as the observational record grows.

## Box 1: Coral Reefs as Metabolic Paradigms — From Odum to Decline

**From healthy reefs to ecosystem dysfunction**: In 1954, Howard T. Odum conducted the first whole-ecosystem metabolism measurements at Eniwetok Atoll in the Marshall Islands, revolutionizing our understanding of coral reefs and ecosystem ecology broadly. Using innovative upstream-downstream oxygen flux techniques, Odum and colleagues demonstrated that coral reefs maintain exceptionally high gross production (200-400 mmol O₂ m⁻² hr⁻¹), rivaling the most productive terrestrial ecosystems despite occurring in nutrient-poor tropical waters. Crucially, Odum showed that reef production and respiration were nearly balanced (P/R ≈ 1), indicating "mature" ecosystems maintaining high biomass with minimal net production—energy was tightly recycled rather than exported.

**Conceptual contributions**: Odum's reef studies established foundational concepts in ecosystem ecology: (1) **P/R ratios as maturity indices** - mature systems approach metabolic balance; (2) **Nutrient spiraling** - efficient nutrient retention and recycling in oligotrophic environments; (3) **Trophic efficiency** - high biomass per unit energy flow through tight food web linkages; (4) **Ecosystem development** - progression from r-selected, high-net-production systems to K-selected, high-biomass systems. Coral reefs became the archetype of "climax" ecosystems optimized for stability rather than growth.

**Contemporary decline**: Our MCR data reveal the reversal of Odum's trajectory. Where Eniwetok showed P/R ≈ 1 with high biomass and efficient nutrient cycling, MCR exhibits increasing P/R (from ~`r round(mean(Year_Averages$Pmax[Year_Averages$Year %in% 2008:2010] / Year_Averages$Rd[Year_Averages$Year %in% 2008:2010], na.rm = TRUE), 1)` to ~`r round(mean(Year_Averages$Pmax[Year_Averages$Year %in% 2023:2025] / Year_Averages$Rd[Year_Averages$Year %in% 2023:2025], na.rm = TRUE), 1)`) driven by collapsing biomass (70% coral loss) and disproportionate loss of respiratory capacity, alongside disrupted nutrient dynamics (respiration-N coupling weakened). This represents **ecosystem retrogression**—the undoing of ecological development. The shift away from P/R ≈ 1 reflects loss of the tight metabolic balance that characterized mature coral reefs, replaced by metabolic imbalance where production persists but respiratory biomass is lost. Climate change is dismantling the metabolic paradigm Odum described, with coral reefs losing the very characteristics that made them foundational to ecosystem science. The MCR time series documents this transition in unprecedented detail, providing a metabolic obituary for a disappearing ecosystem type.

## Box 2: Structural Equation Modelling for Causal Inference in Long-Term Ecology

**The challenge**: Long-term ecological monitoring generates rich observational data across multiple variables, but distinguishing correlation from causation remains difficult. Traditional regression approaches test pairwise relationships but cannot simultaneously evaluate complex networks of direct and indirect effects. This is critical for understanding how climate impacts cascade through ecosystems—does temperature affect fish biomass directly through physiological stress, or indirectly through coral habitat loss?

**Bayesian SEM solution**: Structural Equation Modeling (SEM) extends regression to multivariate systems with explicit causal structure. We specify hypothesized relationships among variables using a Directed Acyclic Graph (DAG), where arrows represent causal pathways (e.g., Temperature → Coral → Respiration). The Bayesian implementation offers key advantages: (1) **Uncertainty quantification** - full posterior distributions for all parameters, not just point estimates; (2) **Missing data handling** - imputation integrated into model estimation; (3) **Prior information** - incorporates ecological knowledge through weakly informative priors; (4) **Effect decomposition** - separates direct effects (Temperature → Respiration) from indirect effects (Temperature → Coral → Respiration) using posterior sampling.

**Application to MCR**: Our DAG encoded eight causal relationships based on ecological theory (e.g., coral loss causes algal expansion; algae support herbivores) and temporal precedence (temperature drives coral loss, not vice versa). The Bayesian SEM revealed that **indirect effects dominate** for metabolic variables—temperature impacts respiration 2-3× more strongly through coral decline than through direct thermal effects. This finding was invisible to pairwise correlations but emerged from the full causal network.

**Broader implications**: The approach is applicable to any long-term monitoring program with multivariate data and hypothesized causal structure. LTER sites, marine observatories, and climate monitoring networks all collect data suitable for SEM, yet most analyses remain correlative. Our workflow (DAG specification → Bayesian estimation → effect decomposition → projection) provides a template for extracting causal insights from observational time series, advancing LTER from description to mechanistic understanding. The key is being explicit about causal assumptions (DAG) rather than pretending they don't exist in correlation analysis.

# Supplementary Material

## Supplementary Figure 1: Photosynthesis-Irradiance Curves

```{r fig-pi-curves}
#| label: fig-pi-curves
#| fig-cap: "**Supplementary Figure 1. Annual photosynthesis-irradiance (PI) curves for MCR LTER Site 1 backreef, 2008–2025.** Points show individual hourly net community production measurements (mmol O₂ m⁻² hr⁻¹) plotted against photosynthetically active radiation (PAR; µmol photons m⁻² s⁻¹). Lines show the fitted rectangular hyperbola for each year, with parameters (α, Pmax, Rd) estimated simultaneously from all 2,928 observations using the Bayesian non-linear model described in Methods. Inset text in each panel shows the posterior median Pmax and Rd (rounded to the nearest integer; units: mmol O₂ m⁻² hr⁻¹). The progressive decline in both Pmax and Rd across years reflects the loss of photosynthetic capacity and ecosystem respiration documented in the main text."
#| fig-width: 9
#| fig-height: 7

par_seq <- seq(0, max(LTER1_Pnet$PAR, na.rm = TRUE), length.out = 200)

pi_params_plot <- fixef(fit1_f) |>
  as.data.frame() |>
  rownames_to_column("param") |>
  as_tibble() |>
  separate(param, into = c("type", "year_str"), sep = "_Year", remove = FALSE) |>
  mutate(year = as.integer(year_str)) |>
  filter(!is.na(year)) |>
  select(type, year, Estimate) |>
  pivot_wider(names_from = type, values_from = Estimate) |>
  mutate(Rd = -Rd)

fitted_curves <- pi_params_plot |>
  rowwise() |>
  reframe(
    year   = year,
    PAR    = par_seq,
    PP_fit = (alpha * Pmax * par_seq) / (alpha * par_seq + Pmax) - Rd
  )

raw_data_plot <- LTER1_Pnet |>
  mutate(year = as.integer(as.character(Year)))

param_labels <- pi_params_plot |>
  mutate(
    label = sprintf("Pmax = %.0f\nRd = %.0f", Pmax, Rd),
    PAR   = max(LTER1_Pnet$PAR, na.rm = TRUE) * 0.97,
    PP    = min(LTER1_Pnet$PP,  na.rm = TRUE) * 0.85
  )

ggplot() +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_point(data = raw_data_plot,
             aes(x = PAR, y = PP),
             colour = "grey50", size = 0.5, alpha = 0.4, shape = 16) +
  geom_line(data = fitted_curves,
            aes(x = PAR, y = PP_fit),
            colour = "#1A237E", linewidth = 0.7) +
  geom_text(data = param_labels,
            aes(x = PAR, y = PP, label = label),
            hjust = 1, vjust = 0, size = 2.2, colour = "#1A237E",
            lineheight = 1.2) +
  facet_wrap(~year, ncol = 6) +
  scale_x_continuous(breaks = c(0, 1000, 2000)) +
  labs(
    x = expression(PAR~(mu*mol~photons~m^{-2}~s^{-1})),
    y = expression(Net~production~(mmol~O[2]~m^{-2}~hr^{-1}))
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(size = 7.5, face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text        = element_text(size = 6)
  )
```

## Supplementary Table 1: SEM Path Coefficients

```{r tbl-sem-paths}
#| label: tbl-sem-paths
#| tbl-cap: "**Supplementary Table 1.** Standardised path coefficients from the Bayesian multivariate structural equation model (12 pathways, n = 20 annual observations, 2005–2025). All continuous variables were log-transformed where appropriate and z-score standardised (mean = 0, SD = 1) prior to modelling. Year residual is the residual of calendar year regressed on maximum annual temperature, capturing secular coral decline beyond the thermal trend. β = posterior median standardised path coefficient; 95% CI = posterior credible interval; Rhat = potential scale reduction factor (values < 1.01 indicate convergence); Bulk ESS = bulk effective sample size; Tail ESS = tail effective sample size. Pathways for which the 95% CI excludes zero are considered credible (marked †)."

# Build tidy table from brms fixed effects summary
sem_coef_table <- as.data.frame(summary(brms_sem_full)$fixed) |>
  tibble::rownames_to_column("term") |>
  # Keep only regression coefficients (exclude intercepts)
  filter(!grepl("Intercept", term)) |>
  mutate(
    # Parse response and predictor from "response_predictor" format
    Pathway = case_when(
      term == "temperature_Year"         ~ "Year → Temperature",
      term == "logcoral_temperature"     ~ "Temperature → Coral cover",
      term == "logcoral_yearresid"       ~ "Year residual → Coral cover",
      term == "logrd_logcoral"           ~ "Coral cover → Ecosystem respiration",
      term == "logrd_mitemperature"      ~ "Temperature → Ecosystem respiration (direct)",
      term == "logfleshy_logcoral"       ~ "Coral cover → Fleshy macroalgae",
      term == "Corallivore_logcoral"     ~ "Coral cover → Corallivore biomass",
      term == "herbs_logfleshy"          ~ "Fleshy macroalgae → Herbivore biomass",
      term == "logpmax_milogrd"          ~ "Ecosystem respiration → Ecosystem Pmax",
      term == "logpmax_logcoral"         ~ "Coral cover → Ecosystem Pmax (direct)",
      term == "logpmax_mitemperature"    ~ "Temperature → Ecosystem Pmax (direct)",
      term == "Npercent_milogrd"         ~ "Ecosystem respiration → Algal N content",
      TRUE ~ term
    ),
    Credible = if_else(sign(`l-95% CI`) == sign(`u-95% CI`), "†", ""),
    beta  = sprintf("%.3f", Estimate),
    CI    = sprintf("[%.3f, %.3f]", `l-95% CI`, `u-95% CI`),
    Rhat  = sprintf("%.4f", Rhat),
    BESS  = formatC(round(Bulk_ESS), format = "d", big.mark = ","),
    TESS  = formatC(round(Tail_ESS), format = "d", big.mark = ",")
  ) |>
  # Order by causal chain
  mutate(order = case_when(
    grepl("Year →",              Pathway) ~ 1,
    grepl("Temperature →",       Pathway) ~ 2,
    grepl("Year residual →",     Pathway) ~ 2,
    grepl("Coral cover →",       Pathway) ~ 3,
    grepl("Fleshy macroalgae →", Pathway) ~ 4,
    grepl("Ecosystem respiration →", Pathway) ~ 5,
    TRUE ~ 6
  )) |>
  arrange(order, Pathway) |>
  select(Pathway, Credible, beta, CI, Rhat, `Bulk ESS` = BESS, `Tail ESS` = TESS)

kable(sem_coef_table, align = c("l", "c", "r", "c", "r", "r", "r"),
      col.names = c("Pathway", "", "β", "95% CI", "R̂", "Bulk ESS", "Tail ESS")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE,
                font_size = 12) |>
  column_spec(1, width = "14em") |>
  column_spec(2, width = "1em") |>
  footnote(symbol = "95% credible interval excludes zero.")
```

# Conclusions

This long-term analysis of Moorea's backreef ecosystem demonstrates the value of integrated, multi-decadal monitoring for understanding reef dynamics. The data reveal:

1.  **Clear directional change** in community composition driven primarily by warming
2.  **Complex feedbacks** among benthic communities, fish assemblages, and biogeochemical processes
3.  **Declining ecosystem function** as measured by net productivity and respiratory capacity
4.  **Predictable trajectories** that can inform management and restoration planning

Future research should focus on:

-   Identifying potential thresholds and tipping points in the transition process
-   Understanding resilience mechanisms and recovery potential
-   Evaluating management interventions to alter current trajectories
-   Exploring fine-scale spatial heterogeneity in ecosystem responses

The structural equation modeling approach employed here provides a framework for testing causal hypotheses and making projections, but predictions should be interpreted cautiously given the complexity of reef systems and potential for non-linear dynamics.

# References

## Climate Data and CMIP6 Models

**Copernicus Climate Data Store**

Copernicus Climate Change Service (C3S) (2023). CMIP6 climate projections. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). https://cds.climate.copernicus.eu/datasets/projections-cmip6. DOI: 10.24381/cds.c866074c

**CMIP6 Overview**

Eyring, V., Bony, S., Meehl, G. A., Senior, C. A., Stevens, B., Stouffer, R. J., & Taylor, K. E. (2016). Overview of the Coupled Model Intercomparison Project Phase 6 (CMIP6) experimental design and organization. *Geoscientific Model Development*, 9(5), 1937–1958. https://doi.org/10.5194/gmd-9-1937-2016

**SSP Scenarios**

O'Neill, B. C., Tebaldi, C., van Vuuren, D. P., Eyring, V., Friedlingstein, P., Hurtt, G., Knutti, R., Kriegler, E., Lamarque, J.-F., Lowe, J., Meehl, G. A., Moss, R., Riahi, K., & Sanderson, B. M. (2016). The Scenario Model Intercomparison Project (ScenarioMIP) for CMIP6. *Geoscientific Model Development*, 9(9), 3461–3482. https://doi.org/10.5194/gmd-9-3461-2016

Riahi, K., van Vuuren, D. P., Kriegler, E., Edmonds, J., O'Neill, B. C., Fujimori, S., Bauer, N., Calvin, K., Dellink, R., Fricko, O., Lutz, W., Popp, A., Cuaresma, J. C., Kc, S., Leimbach, M., Jiang, L., Kram, T., Rao, S., Emmerling, J., … Tavoni, M. (2017). The Shared Socioeconomic Pathways and their energy, land use, and greenhouse gas emissions implications: An overview. *Global Environmental Change*, 42, 153–168. https://doi.org/10.1016/j.gloenvch.2016.05.009

**Individual CMIP6 Models**

Bi, D., Dix, M., Marsland, S., O'Farrell, S., Sullivan, A., Bodman, R., Law, R., Harman, I., Srbinovsky, J., Rashid, H. A., Dobrohotoff, P., Mackallah, C., Yan, H., Hirst, A., Savita, A., Dias, F. B., Woodhouse, M., Fiedler, R., & Heerdegen, A. (2020). Configuration and spin-up of ACCESS-CM2, the new generation Australian Community Climate and Earth System Simulator Coupled Model. *Journal of Southern Hemisphere Earth Systems Science*, 70(1), 225–251. https://doi.org/10.1071/ES19040

Boucher, O., Servonnat, J., Albright, A. L., Aumont, O., Balkanski, Y., Bastrikov, V., Bekki, S., Bonnet, R., Bony, S., Bopp, L., Braconnot, P., Brockmann, P., Cadule, P., Caubel, A., Cheruy, F., Codron, F., Cozic, A., Cugnet, D., D'Andrea, F., … Vuichard, N. (2020). Presentation and evaluation of the IPSL‐CM6A‐LR climate model. *Journal of Advances in Modeling Earth Systems*, 12(7), e2019MS002010. https://doi.org/10.1029/2019MS002010

Döscher, R., Acosta, M., Alessandri, A., Anthoni, P., Arsouze, T., Bergman, T., Bernardello, R., Boussetta, S., Caron, L.-P., Carver, G., Castrillo, M., Catalano, F., Cvijanovic, I., Davini, P., Dekker, E., Doblas-Reyes, F. J., Docquier, D., Echevarria, P., Fladrich, U., … Zhang, Q. (2022). The EC-Earth3 Earth system model for the Coupled Model Intercomparison Project 6. *Geoscientific Model Development*, 15(7), 2973–3020. https://doi.org/10.5194/gmd-15-2973-2022

Dunne, J. P., Horowitz, L. W., Adcroft, A. J., Ginoux, P., Held, I. M., John, J. G., Krasting, J. P., Malyshev, S., Naik, V., Paulot, F., Shevliakova, E., Stock, C. A., Zadeh, N., Balaji, V., Blanton, C., Dunne, K. A., Dupuis, C., Durachta, J., Dussin, R., … Zhao, M. (2020). The GFDL Earth System Model version 4.1 (GFDL‐ESM 4.1): Overall coupled model description and simulation characteristics. *Journal of Advances in Modeling Earth Systems*, 12(11), e2019MS002015. https://doi.org/10.1029/2019MS002015

Müller, W. A., Jungclaus, J. H., Mauritsen, T., Baehr, J., Bittner, M., Budich, R., Bunzel, F., Esch, M., Ghosh, R., Haak, H., Ilyina, T., Kleine, T., Kornblueh, L., Li, H., Modali, K., Notz, D., Pohlmann, H., Segschneider, J., Tian, F., … Marotzke, J. (2018). A higher-resolution version of the Max Planck Institute Earth System Model (MPI‐ESM1.2‐HR). *Journal of Advances in Modeling Earth Systems*, 10(7), 1383–1413. https://doi.org/10.1029/2017MS001217

Seland, Ø., Bentsen, M., Olivié, D., Toniazzo, T., Gjermundsen, A., Graff, L. S., Debernard, J. B., Gupta, A. K., He, Y.-C., Kirkevåg, A., Schwinger, J., Tjiputra, J., Aas, K. S., Bethke, I., Fan, Y., Griesfeller, J., Grini, A., Guo, C., Ilicak, M., … Schulz, M. (2020). Overview of the Norwegian Earth System Model (NorESM2) and key climate response of CMIP6 DECK and historical simulations. *Geoscientific Model Development*, 13(12), 6165–6200. https://doi.org/10.5194/gmd-13-6165-2020

Sellar, A. A., Jones, C. G., Mulcahy, J. P., Tang, Y., Yool, A., Wiltshire, A., O'Connor, F. M., Stringer, M., Hill, R., Palmieri, J., Woodward, S., de Mora, L., Kuhlbrodt, T., Rumbold, S. T., Kelley, D. I., Ellis, R., Johnson, C. E., Walton, J., Abraham, N. L., … Zerroukat, M. (2019). UKESM1: Description and evaluation of the UK Earth System Model. *Journal of Advances in Modeling Earth Systems*, 11(12), 4513–4558. https://doi.org/10.1029/2019MS001739

Tatebe, H., Ogura, T., Nitta, T., Komuro, Y., Ogochi, K., Takemura, T., Sudo, K., Sekiguchi, M., Abe, M., Saito, F., Chikira, M., Watanabe, S., Mori, M., Hirota, N., Kawatani, Y., Mochizuki, T., Yoshimura, K., Takata, K., O'ishi, R., … Kimoto, M. (2019). Description and basic evaluation of simulated mean state, internal variability, and climate sensitivity in MIROC6. *Geoscientific Model Development*, 12(7), 2727–2765. https://doi.org/10.5194/gmd-12-2727-2019

Wu, T., Lu, Y., Fang, Y., Xin, X., Li, L., Li, W., Jie, W., Zhang, J., Liu, Y., Zhang, L., Zhang, F., Zhang, Y., Wu, F., Li, J., Chu, M., Wang, Z., Shi, X., Liu, X., Wei, M., … Zhou, W. (2019). The Beijing Climate Center Climate System Model (BCC‐CSM2): Main progress from CMIP5 to CMIP6. *Geoscientific Model Development*, 12(4), 1573–1600. https://doi.org/10.5194/gmd-12-1573-2019

Yukimoto, S., Kawai, H., Koshiro, T., Oshima, N., Yoshida, K., Urakawa, S., Tsujino, H., Deushi, M., Tanaka, T., Hosaka, M., Yabu, S., Yoshimura, H., Shindo, E., Mizuta, R., Obata, A., Adachi, Y., & Ishii, M. (2019). The Meteorological Research Institute Earth System Model version 2.0, MRI‐ESM2.0: Description and basic evaluation of the physical component. *Journal of the Meteorological Society of Japan*, 97(5), 931–965. https://doi.org/10.2151/jmsj.2019-051

---

## Statistical Methods and Software

Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using Stan. *Journal of Statistical Software*, 80(1), 1–28. https://doi.org/10.18637/jss.v080.i01

Bürkner, P.-C. (2018). Advanced Bayesian multilevel modeling with the R package brms. *The R Journal*, 10(1), 395–411. https://doi.org/10.32614/RJ-2018-017

Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M., Guo, J., Li, P., & Riddell, A. (2017). Stan: A probabilistic programming language. *Journal of Statistical Software*, 76(1), 1–32. https://doi.org/10.18637/jss.v076.i01

Csárdi, G., & Nepusz, T. (2006). The igraph software package for complex network research. *InterJournal, Complex Systems*, 1695. https://igraph.org

Gelman, A., Goodrich, B., Gabry, J., & Vehtari, A. (2019). R-squared for Bayesian regression models. *The American Statistician*, 73(3), 307–309. https://doi.org/10.1080/00031305.2018.1549100

Hufkens, K., Stauffer, R., & Campitelli, E. (2019). ecmwfr: Programmatic interface to the two main European Centre for Medium-Range Weather Forecasts APIs. *Journal of Open Source Software*, 4(43), 1867. https://doi.org/10.21105/joss.01867

Kay, M. (2023). tidybayes: Tidy data and 'Geoms' for Bayesian models. R package version 3.0.6. https://mjskay.github.io/tidybayes/

Pedersen, T. L. (2024). patchwork: The composer of plots. R package version 1.2.0. https://patchwork.data-imaginist.com

Pierce, D. (2024). ncdf4: Interface to Unidata netCDF (Version 4 or Earlier) Format Data Files. R package version 1.22. https://CRAN.R-project.org/package=ncdf4

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, 4(43), 1686. https://doi.org/10.21105/joss.01686

Wilke, C. O. (2024). cowplot: Streamlined Plot Theme and Plot Annotations for 'ggplot2'. R package version 1.1.3. https://wilkelab.org/cowplot/

Zeileis, A., & Grothendieck, G. (2005). zoo: S3 infrastructure for regular and irregular time series. *Journal of Statistical Software*, 14(6), 1–27. https://doi.org/10.18637/jss.v014.i06

---

## Reef Ecology and Ecosystem Metabolism

Barnes, D. J. (1983). Profiling coral reef productivity and calcification using pH and oxygen electrodes. *Journal of Experimental Marine Biology and Ecology*, 66(2), 149–161. https://doi.org/10.1016/0022-0981(83)90036-0

Courtney, T. A., Andersson, A. J., Bates, N. R., Collins, A., Doherty, S., Cooper, T. F., & Cyronak, T. (2016). Comparing chemistry and census-based estimates of net ecosystem calcification on a rim reef in Bermuda. *Frontiers in Marine Science*, 3, 181. https://doi.org/10.3389/fmars.2016.00181

DeCarlo, T. M., Cohen, A. L., Barkley, H. C., Cobban, Q., Young, C., Shamberger, K. E., Brainard, R. E., & Golbuu, Y. (2017). Coral macrobioerosion is accelerated by ocean acidification and nutrients. *Geology*, 45(5), 399–402. https://doi.org/10.1130/G38737.1

Hoegh-Guldberg, O., Poloczanska, E. S., Skirving, W., & Dove, S. (2017). Coral reef ecosystems under climate change and ocean acidification. *Frontiers in Marine Science*, 4, 158. https://doi.org/10.3389/fmars.2017.00158

Hughes, T. P., Barnes, M. L., Bellwood, D. R., Cinner, J. E., Cumming, G. S., Jackson, J. B. C., Kleypas, J., van de Leemput, I. A., Lough, J. M., Morrison, T. H., Palumbi, S. R., van Nes, E. H., & Scheffer, M. (2017). Coral reefs in the Anthropocene. *Nature*, 546(7656), 82–90. https://doi.org/10.1038/nature22901

Kinsey, D. W. (1985). Metabolism, calcification and carbon production. I. Systems level studies. In *Proceedings of the 5th International Coral Reef Congress* (Vol. 4, pp. 505–526). Antenne Museum-EPHE, Moorea.

McMahon, A., Santos, I. R., Cyronak, T., & Eyre, B. D. (2013). Hysteresis between coral reef calcification and the seawater aragonite saturation state. *Geophysical Research Letters*, 40(15), 4090–4094. https://doi.org/10.1002/grl.50802

Odum, H. T. (1956). Primary production in flowing waters. *Limnology and Oceanography*, 1(2), 102–117. https://doi.org/10.4319/lo.1956.1.2.0102

Odum, E. P. (1969). The strategy of ecosystem development. *Science*, 164(3877), 262–270. https://doi.org/10.1126/science.164.3877.262

Salinas-de-León, P., Dryden, C., & Costello, M. J. (2018). Characterisation of net ecosystem production in shallow tropical coral reef and seagrass ecosystems. *Marine Ecology Progress Series*, 590, 37–49. https://doi.org/10.3354/meps12471

Santos, I. R., Glud, R. N., Maher, D., Erler, D., & Eyre, B. D. (2011). Diel coral reef acidification driven by porewater advection in permeable carbonate sands, Heron Island, Great Barrier Reef. *Geophysical Research Letters*, 38(3), L03604. https://doi.org/10.1029/2010GL046053

Shamberger, K. E. F., Feely, R. A., Sabine, C. L., Atkinson, M. J., DeCarlo, E. H., Mackenzie, F. T., Drupp, P. S., & Butterfield, D. A. (2011). Calcification and organic production on a Hawaiian coral reef. *Marine Chemistry*, 127(1–4), 64–75. https://doi.org/10.1016/j.marchem.2011.08.003

Shaw, E. C., McNeil, B. I., & Tilbrook, B. (2015). Chemistry and dynamics of the inorganic carbon species in an Australian coral reef. *Marine Chemistry*, 177(2), 194–207. https://doi.org/10.1016/j.marchem.2015.06.008

Yeakel, K. L., Andersson, A. J., Bates, N. R., Courtney, T. A., & Houck, J. (2015). Shifts in coral reef biogeochemistry and resulting acidification linked to offshore productivity. *Proceedings of the National Academy of Sciences*, 112(47), 14512–14517. https://doi.org/10.1073/pnas.1507021112

---

# Session Information

```{r session-info}
#| label: session-info

sessionInfo()
```
