---
title: "Moorea Coral Reef Productivity Analysis: Long-term Ecosystem Dynamics"
author: "Nyssa Silbiger"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  cache: true
---


# Introduction

This report analyzes long-term ecosystem dynamics at LTER Site 1 in Moorea, French Polynesia. We examine relationships between benthic community composition, environmental conditions, fish assemblages, and ecosystem metabolism over nearly two decades (2005-2025). The analysis integrates multiple datasets to understand how coral reef ecosystems are responding to environmental change.

## Setup

```{r setup}
#| label: setup
#| cache: false

# Load required libraries
library(tidyverse)
library(here)
library(lubridate)
library(brms)
library(tidybayes)
library(scales)
library(patchwork)
library(viridis)
library(bayesplot)
library(ggridges)
library(ggsci)
library(psych)
library(ggcorrplot2)
library(ggraph)
library(igraph)

# Set global theme
theme_set(theme_bw())

# Set seed for reproducibility
set.seed(223)
```

# Data Processing Pipeline

## 1. Temperature Data

Extract yearly average and maximum daily temperature values from the backreef temperature time series.

```{r temperature-processing}
#| label: process-temperature

# Read and process temperature data
Temp_LTER1 <- read_csv(here("Data", "raw_data", "MCR_LTER02_BTM_Backreef_Forereef_20251114.csv")) |>
  filter(reef_type_code == "Backreef")

# Calculate daily means, then yearly statistics (optimized with single pass)
Mean_year_Temp <- Temp_LTER1 |>
  mutate(date = as.Date(time_local)) |>
  group_by(date) |>
  summarise(daily_temp = mean(temperature_c, na.rm = TRUE), .groups = "drop") |>
  mutate(Year = year(date)) |>
  group_by(Year) |>
  summarise(
    Mean_temp = mean(daily_temp, na.rm = TRUE),
    Max_temp = max(daily_temp, na.rm = TRUE),
    .groups = "drop"
  )

# Save processed data
write_csv(Mean_year_Temp, here("Data", "InSituTemp.csv"))

# Quick summary
summary(Mean_year_Temp)
```

## 2. Nutrient Data

Process macroalgae tissue nitrogen content and water column nutrients, using tissue N to impute missing water column data.

```{r nutrient-processing}
#| label: process-nutrients

# Turbinaria %N data (tissue nitrogen as nutrient proxy)
Turb <- read_csv(here("Data", "raw_data", "MCR_LTER_Macroalgal_CHN_2005_to_2024_20250616.csv")) |>
  filter(
    Habitat == "Backreef",
    Genus == "Turbinaria",
    Site == "LTER_1",
    !is.na(N)
  ) |>
  group_by(Year) |>
  summarise(
    N_percent = mean(N, na.rm = TRUE),
    C_percent = mean(C, na.rm = TRUE),
    CN = mean(CN_ratio, na.rm = TRUE),
    .groups = "drop"
  )

write_csv(Turb, here("Data", "Turb_mean.csv"))

# Water column nutrients (bimonthly surveys 2005-2018)
water <- read_csv(here("Data", "raw_data", "WaterColumnN.csv")) |>
  mutate(
    Date = mdy(Date),
    Year = year(Date)
  ) |>
  group_by(Year) |>
  summarise(across(Phosphate:Nitrite_and_Nitrate,  \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  mutate(Site = "LTER 1")

write_csv(water, here("Data", "Water_N_mean.csv"))

# Impute water N+N from tissue N using linear model
AllNutrients <- Turb |>
  full_join(water, by = "Year") |>
  arrange(Year)

# Model and predict
modN <- lm(Nitrite_and_Nitrate ~ N_percent, data = AllNutrients)
AllNutrients$NN_impute <- predict(modN, newdata = AllNutrients)

# Model summary
cat("\nNutrient imputation model (Nitrite+Nitrate ~ Tissue %N):\n")
summary(modN)
```

## 3. Benthic Cover Data

Calculate percent cover for functional groups from benthic transect surveys.

```{r benthic-processing}
#| label: process-benthic

# Read and classify benthic data
BenthicCover_Algae <- read_csv(here("Data", "raw_data", "MCR_LTER_Annual_Survey_Benthic_Cover_20251009.csv")) |>
  filter(Habitat == "Backreef")

# Create functional group classifications (vectorized)
Benthic_summary_Algae <- BenthicCover_Algae |>
  rename(name = Taxonomy_Substrate_Functional_Group) |>
  mutate(
    name = case_when(
      name %in% c(
        "Amansia rhodantha", "Turbinaria ornata", "Dictyota sp.", "Halimeda sp.",
        "Galaxaura sp.", "Liagora ceranoides", "Cyanophyta", "Halimeda minima",
        "Amphiroa fragilissima", "Caulerpa serrulata", "Corallimorpharia",
        "Dictyota friabilis", "Galaxaura rugosa", "Cladophoropsis membranacea",
        "Galaxaura filamentosa", "Halimeda discoidea", "Peyssonnelia inamoena",
        "Caulerpa racemosa", "Valonia ventricosa", "Actinotrichia fragilis",
        "Dictyota bartayresiana", "Microdictyon umbilicatum", "Halimeda distorta",
        "Halimeda incrassata", "Halimeda macroloba", "Dictyota implexa",
        "Gelidiella acerosa", "Dictyosphaeria cavernosa", "Valonia aegagropila",
        "Microdictyon okamurae", "Halimeda opuntia", "Dichotomaria obtusata",
        "Chlorodesmis fastigiata", "Phyllodictyon anastomosans", "Phormidium sp.",
        "Cladophoropsis luxurians", "Sargassum pacificum", "Chnoospora implexa",
        "Halimeda taenicola", "Boodlea kaeneana", "Padina boryana",
        "Coelothrix irregularis", "Gelidiella sp.", "Hydroclathrus clathratus",
        "Dictyota divaricata", "Hypnea spinella", "Dichotomaria marginata",
        "Sporolithon sp.", "Chaetomorpha antennina", "Asparagopsis taxiformis"
      ) ~ "Fleshy Macroalgae",
      name %in% c(
        "Algal Turf", "Damselfish Turf", "Coral Rubble",
        "Lobophora variegata", "Shell Debris", "Bare Space"
      ) ~ "Turf/Cyanobacteria",
      name %in% c(
        "Peyssonnelia bornetii", "Peyssonnelia sp.",
        "Sponge", "Tridacna sp.", "No data"
      ) ~ "Sand",
      TRUE ~ name
    )
  ) |>
  group_by(Year, Site, name) |>
  summarise(
    total_cover = sum(Percent_Cover, na.rm = TRUE),
    mean_cover = 100 * total_cover / 5000,
    .groups = "drop"
  ) |>
  mutate(Site = str_replace_all(Site, "_", " "))

write_csv(Benthic_summary_Algae, here("Data", "Benthic_summary_algae.csv"))

# Calculate total calcifiers and living cover
Total_Calc <- Benthic_summary_Algae |>
  filter(name %in% c("Coral", "Crustose Corallines")) |>
  group_by(Year, Site) |>
  reframe(total_Calc = sum(mean_cover))

TotalLiving <- Benthic_summary_Algae |>
  filter(name %in% c("Coral", "Crustose Corallines", "Fleshy Macroalgae")) |>
  group_by(Year, Site) |>
  summarise(
    mean_alive = sum(mean_cover),
    mean_fleshy = sum(mean_cover[name == "Fleshy Macroalgae"], na.rm = TRUE),
    mean_coral = sum(mean_cover[name == "Coral"], na.rm = TRUE),
    .groups = "drop"
  )
```

## 4. Fish Biomass Data

Process fish survey data by trophic groups.

```{r fish-processing}
#| label: process-fish

# Read and process fish data
fish <- read_csv(here("Data", "raw_data", "MCR_LTER_Annual_Fish_Survey_20250324.csv"))

# Process by trophic groups (optimized filtering)
fish_summary <- fish |>
  filter(
    Site == "LTER_1",
    Habitat == "Backreef",
    Biomass < 8000, # Remove outlier sharks
    Biomass >= 0 # Remove negative (missing) values
  ) |>
  mutate(
    trophic_new = case_when(
      Fine_Trophic %in% c("Brusher", "Browser", "Excavator", "Scraper") ~ "Herbivores",
      Fine_Trophic == "Corallivore" ~ "Corallivore",
      TRUE ~ "Other"
    )
  ) |>
  group_by(Year, trophic_new) |>
  summarise(
    fish_g_m2 = sum(Biomass, na.rm = TRUE) / 1200, # 4 transects × 300 m²
    .groups = "drop"
  )

write_csv(fish_summary, here("Data", "fish_summary.csv"))
```

## 5. Primary Production Processing

Process raw production data and extract Pmax and Rd using Bayesian photosynthesis-irradiance curves.

```{r pp-processing}
#| label: process-pp
#| cache: true

# Read all QC PP files efficiently
filedir <- here("Data", "raw_data", "QC_PP")
files <- dir(path = filedir, pattern = ".csv", full.names = TRUE)

All_PP_data <- files |>
  set_names() |>
  map_df(read_csv, .id = "filename", show_col_types = FALSE) |>
  mutate(DateTime = mdy_hm(DateTime))

# Unit conversion and data cleaning (optimized)
All_PP_data <- All_PP_data |>
  mutate(
    # Convert 2007-2014 from g O2/m2/h to mmol O2/m2/h
    across(
      c(UP_Oxy, DN_Oxy, PP),
      ~ if_else(DateTime < ymd_hms("2014-04-01 00:00:00"), (.x / 32) * 1000, .x)
    ),
    Date = as_date(DateTime),
    DielDateTime = DateTime + hours(12),
    DielDate = as_date(DielDateTime),
    Year = year(DateTime)
  ) |>
  filter(!Date %in% mdy(c("5/27/2011", "5/28/2011", "1/21/2014", "05/25/2024")))

# Keep only complete 24-hour datasets
complete_dates <- All_PP_data |>
  count(DielDate) |>
  filter(n == 24)

All_PP_data <- complete_dates |>
  left_join(All_PP_data, by = "DielDate")

# Calculate GP and environmental averages
Daily_R <- All_PP_data |>
  group_by(Year, Season, DielDate) |>
  summarise(R_average = mean(PP[PAR == 0], na.rm = TRUE), .groups = "drop")

All_PP_data <- All_PP_data |>
  left_join(Daily_R, by = c("Year", "Season", "DielDate")) |>
  mutate(
    GP = if_else(PAR == 0, NA_real_, PP - R_average),
    GP = if_else(GP < 0, NA_real_, GP),
    Temperature_mean = (UP_Temp + DN_Temp) / 2,
    Flow_mean = (UP_Velocity_mps + DN_Velocity_mps) / 2
  )

# Calculate yearly averages
Year_Averages_PP <- All_PP_data |>
  mutate(
    NP = PP,
    R = if_else(PP < 0, PP, NA_real_)
  ) |>
  group_by(Year) |>
  summarise(
    NP_mean = mean(NP, na.rm = TRUE),
    NP_SE = sd(NP, na.rm = TRUE) / sqrt(n()),
    NP_max = max(NP, na.rm = TRUE),
    GP_mean = mean(GP, na.rm = TRUE),
    GP_SE = sd(GP, na.rm = TRUE) / sqrt(n()),
    R_mean = mean(R, na.rm = TRUE),
    R_SE = sd(R, na.rm = TRUE) / sqrt(n()),
    Temperature_mean = mean(Temperature_mean, na.rm = TRUE),
    Flow_mean = mean(Flow_mean, na.rm = TRUE),
    PAR_mean = mean(PAR[PAR > 0], na.rm = TRUE),
    .groups = "drop"
  )
```

## Sampling Coverage Summary

```{r}
#| label: sampling-coverage

# Calculate sampling metrics
metabolism_coverage <- All_PP_data |>
  group_by(Year) |>
  summarise(
    n_measurements = n() / 24,  # Convert hourly to daily measurements
    .groups = "drop"
  )

benthic_coverage <- Benthic_summary_Algae |>
  filter(Site == "LTER 1") |>
  group_by(Year) |>
  summarise(surveys = n_distinct(Year), .groups = "drop") |>
  nrow()

fish_coverage <- fish_summary |>
  group_by(Year) |>
  summarise(surveys = n_distinct(Year), .groups = "drop") |>
  nrow()

cat("=== Data Coverage Summary ===\n\n")
cat(sprintf("Ecosystem Metabolism:\n"))
cat(sprintf("  Total 24-hour cycles: %.0f over 20 years\n", sum(metabolism_coverage$n_measurements)))
cat(sprintf("  Annual coverage: %.1f ± %.1f measurements per year (range: %.0f-%.0f)\n",
            mean(metabolism_coverage$n_measurements),
            sd(metabolism_coverage$n_measurements),
            min(metabolism_coverage$n_measurements),
            max(metabolism_coverage$n_measurements)))
cat(sprintf("  Temporal span: %d years (%d-%d)\n\n", 
            length(unique(All_PP_data$Year)),
            min(All_PP_data$Year),
            max(All_PP_data$Year)))

cat(sprintf("Benthic Surveys:\n"))
cat(sprintf("  Total surveys: %d (annual coverage)\n", benthic_coverage))
cat(sprintf("  Years with data: %d of %d possible (100%%)\n\n", benthic_coverage, 21))

cat(sprintf("Fish Surveys:\n"))
cat(sprintf("  Total surveys: %d (annual coverage)\n", fish_coverage))
cat(sprintf("  Years with data: %d of %d possible (100%%)\n\n", fish_coverage, 21))

cat(sprintf("Temperature:\n"))
cat(sprintf("  Continuous 15-minute measurements: %d-%d (100%% coverage)\n\n", 
            min(lubridate::year(Temp_LTER1$time_local), na.rm = TRUE),
            max(lubridate::year(Temp_LTER1$time_local), na.rm = TRUE)))

cat(sprintf("Nutrients:\n"))
cat(sprintf("  Water column (direct): 2005-2018\n"))
cat(sprintf("  Tissue nitrogen: 2005-2024 (2019-2025 water column imputed)\n"))
```

## Bayesian Photosynthesis-Irradiance (PI) Curve Analysis

We fit a Bayesian non-linear model to extract photosynthetic parameters (alpha, Pmax, and Rd) for each year. The model uses a rectangular hyperbola:

$$PP = \frac{\alpha \cdot P_{max} \cdot PAR}{\alpha \cdot PAR + P_{max}} + R_d$$

where: - $\alpha$ = initial slope (photosynthetic efficiency) - $P_{max}$ = maximum photosynthetic capacity - $R_d$ = dark respiration (ecosystem respiration) - $PAR$ = photosynthetically active radiation

```{r bayesian-pi-curve}
# Bayesian PI curves to extract light-independent Pmax and Rd
LTER1_Pnet <- All_PP_data |>
  mutate(
    tempc = Temperature_mean,
    Flowmean = Flow_mean,
    flow_log = log(Flowmean),
    UPDN = as.factor(UPDN),
    Year = as.factor(Year)
  )

# Fit Bayesian nonlinear model
fit1_f <- brm(
  bf(
    PP ~ ((alpha * Pmax * PAR) / (alpha * PAR + Pmax)) + Rd,
    nl = TRUE,
    alpha ~ 0 + Year,
    Pmax ~ 0 + Year,
    Rd ~ 0 + Year
  ) + student(),
  data = LTER1_Pnet,
  set_rescor(FALSE),
  prior = c(
    prior(normal(0.1, 10), nlpar = "alpha", lb = 0),
    prior(normal(-100, 10), nlpar = "Rd", ub = 0),
    prior(normal(100, 10), nlpar = "Pmax", lb = 0)
  ),
  control = list(adapt_delta = 0.95, max_treedepth = 20),
  cores = 3,
  chains = 3,
  seed = 223,
  iter = 8000,
  warmup = 2000
)

# Extract Pmax and Rd coefficients
names <- rownames(fixef(fit1_f))
coefficients <- as_tibble(fixef(fit1_f)) |>
  mutate(params = names)

PRValues <- coefficients |>
  separate(params, sep = "_", into = c("param_type", "yearname"), remove = FALSE) |>
  filter(param_type %in% c("Pmax", "Rd")) |>
  separate(yearname, sep = "r", into = c("y", "year")) |>
  mutate(year = as.numeric(year)) |>
  select(Year = year, Estimate, param_type) |>
  pivot_wider(names_from = param_type, values_from = Estimate)

# Combine with raw averages
Year_Averages_PP <- Year_Averages_PP |>
  full_join(PRValues, by = "Year") |>
  mutate(Rd = -Rd) # Make Rd positive

# Calculate Bayesian R²
pi_curve_r2 <- median(bayes_R2(fit1_f)[, 1])
cat(sprintf("\nPhotosynthesis-Irradiance Curve Model:\n"))
cat(sprintf("  Bayesian R²: %.3f\n", pi_curve_r2))
cat(sprintf("  This indicates the model explains %.1f%% of variance in net production\n", pi_curve_r2 * 100))

write_csv(Year_Averages_PP, here("Data", "Year_Averages_PP.csv"))
```

## 6. Integrate All Datasets

```{r integrate-data}
#| label: integrate-datasets

# Read processed data files
Year_Averages_PP <- read_csv(here("Data", "Year_Averages_PP.csv"))
Turb <- read_csv(here("Data", "Turb_mean.csv"))
water <- read_csv(here("Data", "Water_N_mean.csv"))
insitu_temp <- read_csv(here("Data", "InSituTemp.csv"))

fish_trophic <- read_csv(here("Data", "fish_summary.csv")) |>
  pivot_wider(names_from = trophic_new, values_from = fish_g_m2) |>
  select(-Other)

Benthic_summary_Algae <- read_csv(here("Data", "Benthic_summary_algae.csv"))

# Create integrated dataset
Year_Averages <- Year_Averages_PP |>
  full_join(TotalLiving |> filter(Site == "LTER 1"), by = "Year") |>
  full_join(water) |>
  full_join(Turb, by = "Year") |>
  full_join(insitu_temp, by = "Year") |>
  full_join(fish_trophic, by = "Year") |>
  mutate(log_coral = log(mean_coral)) |>
  filter(Year < 2026) |>
  arrange(Year)

# Display structure
glimpse(Year_Averages)
```

# Exploratory Data Analysis

## Benthic Community Composition

```{r benthic-plot}
#| label: fig-benthic
#| fig-cap: "Benthic community composition at LTER Site 1 (2005-2025)"
#| fig-width: 10
#| fig-height: 6

# Define color palette
myPal <- c(
  Coral = "#c38370",
  `Fleshy Macroalgae` = "#99ddaa",
  `Turf/Cyanobacteria` = "#9da993",
  `Crustose Corallines` = "#523a28",
  Other = "#bdc3cb",
  Sand = "#d6ad60"
)

# Create stacked area plot
Benthic_summary_Algae |>
  mutate(name = factor(name, levels = c(
    "Coral", "Fleshy Macroalgae",
    "Crustose Corallines", "Turf/Cyanobacteria",
    "Sand", "Other"
  ))) |>
  filter(Site == "LTER 1") |>
  ggplot(aes(x = Year, y = mean_cover, fill = name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = myPal) +
  labs(
    fill = "",
    y = "% Cover",
    title = "Long-term Benthic Community Dynamics"
  ) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.position = "bottom"
  )
```

**Key Observation:** Coral cover declined from `r round(Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover), 1)`% in 2006 to `r round(Benthic_summary_Algae |> filter(Year == 2025, name == "Coral", Site == "LTER 1") |> pull(mean_cover), 1)`% in 2025, representing a `r round((1 - (Benthic_summary_Algae |> filter(Year == 2025, name == "Coral", Site == "LTER 1") |> pull(mean_cover)) / (Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover))) * 100, 1)`% reduction.

## Environmental Drivers

```{r environmental-plots}
#| label: fig-environment
#| fig-cap: "Environmental conditions over time"
#| fig-width: 10
#| fig-height: 6

p1 <- Year_Averages |>
  ggplot(aes(x = Year, y = Max_temp)) +
  geom_line(color = "firebrick", linewidth = 1) +
  geom_point(color = "firebrick", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", alpha = 0.2) +
  labs(
    y = "Max Daily Temp (°C)",
    title = "A. Temperature"
  )

p2 <- Year_Averages |>
  ggplot(aes(x = Year, y = N_percent)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "forestgreen", alpha = 0.2) +
  labs(
    y = "Tissue N Content (%)",
    title = "B. Nitrogen"
  )

p1 + p2
```

## Fish Community Trends

```{r fish-plots}
#| label: fig-fish
#| fig-cap: "Fish biomass by trophic group"
#| fig-width: 10
#| fig-height: 4

fish_summary |>
  ggplot(aes(x = Year, y = fish_g_m2, color = trophic_new)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  facet_wrap(~trophic_new, scales = "free_y") +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  labs(
    x = "Year",
    y = "Biomass (g/m²)",
    title = "Fish Biomass Trends by Trophic Group"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## Ecosystem Metabolism Time Series

```{r metabolism-timeseries}
#| label: fig-metabolism
#| fig-cap: "Twenty-year time series of ecosystem metabolism showing declines in photosynthetic capacity (Pmax), respiration (Rd), and net production (NP). Points represent annual means; shaded ribbons show 95% credible intervals from Bayesian linear trend models. Pmax and Rd were extracted from Bayesian photosynthesis-irradiance curves to provide light-independent metabolic indices."
#| fig-width: 10
#| fig-height: 8
#| cache: true

# Prepare data for plotting with expression labels
metabolism_data <- Year_Averages |>
  select(Year, Pmax, Rd, NP_mean) |>
  pivot_longer(cols = c(Pmax, Rd, NP_mean), 
               names_to = "metric", 
               values_to = "value") |>
  mutate(metric_label = case_when(
    metric == "Pmax" ~ "Pmax",
    metric == "Rd" ~ "Rd",
    metric == "NP_mean" ~ "NP"
  )) |>
  mutate(metric_label = factor(metric_label, levels = c("Pmax", "Rd", "NP")))

# Create labeller function using expression notation
metric_labeller <- as_labeller(c(
  "Pmax" = "atop('Maximum Photosynthetic Capacity', paste('(mmol ', O[2], ' m'^-2, ' hr'^-1, ')'))",
  "Rd" = "atop('Ecosystem Respiration', paste('(mmol ', O[2], ' m'^-2, ' hr'^-1, ')'))",
  "NP" = "atop('Net Ecosystem Production', paste('(mmol ', O[2], ' m'^-2, ' hr'^-1, ')'))"
), default = label_parsed)

# Fit Bayesian linear models for each metric to get trend statistics
pmax_brm <- brm(Pmax ~ Year, data = Year_Averages, silent = 2, refresh = 0, seed = 223)
rd_brm <- brm(Rd ~ Year, data = Year_Averages, silent = 2, refresh = 0, seed = 223)
np_brm <- brm(NP_mean ~ Year, data = Year_Averages, silent = 2, refresh = 0, seed = 223)

# Extract trend statistics from Bayesian models
pmax_slope <- fixef(pmax_brm)["Year", "Estimate"]
pmax_ci <- fixef(pmax_brm)["Year", c("Q2.5", "Q97.5")]
rd_slope <- fixef(rd_brm)["Year", "Estimate"]
rd_ci <- fixef(rd_brm)["Year", c("Q2.5", "Q97.5")]
np_slope <- fixef(np_brm)["Year", "Estimate"]
np_ci <- fixef(np_brm)["Year", c("Q2.5", "Q97.5")]

# Calculate Bayesian R-squared
pmax_r2 <- median(bayes_R2(pmax_brm)[, 1])
rd_r2 <- median(bayes_R2(rd_brm)[, 1])
np_r2 <- median(bayes_R2(np_brm)[, 1])

# Create annotations with trend statistics using expression notation for geom_text
trend_labels <- tibble(
  metric_label = factor(c("Pmax", "Rd", "NP"), levels = c("Pmax", "Rd", "NP")),
  x = c(2008, 2008, 2008),
  y = c(Inf, Inf, Inf),
  label = c(
    sprintf("'%.2f mmol O'[2]*' m'^-2*' hr'^-1*' yr'^-1", pmax_slope),
    sprintf("'%.2f mmol O'[2]*' m'^-2*' hr'^-1*' yr'^-1", rd_slope),
    sprintf("'%.2f mmol O'[2]*' m'^-2*' hr'^-1*' yr'^-1", np_slope)
  )
)

# Create publication-quality figure
metabolism_plot <- ggplot(metabolism_data, aes(x = Year, y = value)) +
  # Add trend line with confidence interval
  geom_smooth(
    method = "lm", 
    se = TRUE, 
    color = "#2E86AB",
    fill = "#2E86AB",
    alpha = 0.15,
    linewidth = 1
  ) +
  # Add data points
  geom_point(
    size = 3, 
    color = "#2E86AB",
    alpha = 0.7
  ) +
  # Add connecting line
  geom_line(
    color = "#2E86AB",
    linewidth = 0.5,
    alpha = 0.4
  ) +
  # Add trend statistics with expression parsing
  geom_text(
    data = trend_labels,
    aes(x = x, y = y, label = label),
    hjust = 0,
    vjust = 1.2,
    size = 4.5,
    color = "gray30",
    parse = TRUE
  ) +
  # Facet by metric
  facet_wrap(
    ~metric_label, 
    ncol = 1, 
    scales = "free_y",
    strip.position = "left",
    labeller = metric_labeller
  ) +
  # Labels and theme
  labs(
    x = "Year",
    y = NULL,
    title = "Long-term Ecosystem Metabolism Decline at MCR LTER Site 1",
    subtitle = "Backreef metabolism measurements (2005-2025) showing directional changes in all metabolic components"
  ) +
  theme_bw(base_size = 16) +
  theme(
    # Strip formatting
    strip.background = element_blank(),
    strip.text = element_text(size = 14, face = "bold", hjust = 0),
    strip.placement = "outside",
    
    # Panel formatting
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    panel.border = element_rect(color = "gray60", linewidth = 0.5),
    
    # Axis formatting
    axis.title.x = element_text(size = 15, margin = margin(t = 10)),
    axis.text = element_text(size = 13, color = "gray20"),
    axis.ticks = element_line(color = "gray60"),
    
    # Title formatting
    plot.title = element_text(size = 17, face = "bold", margin = margin(b = 5)),
    plot.subtitle = element_text(size = 13, color = "gray40", margin = margin(b = 15)),
    
    # Overall plot
    plot.margin = margin(15, 15, 10, 10),
    plot.background = element_rect(fill = "white", color = NA),
    panel.spacing = unit(1.5, "lines")
  ) +
  scale_x_continuous(breaks = seq(2007, 2025, by = 3), 
                     minor_breaks = seq(2007, 2025, by = 1),
                     limits = c(2007, 2025))

metabolism_plot

# Print summary statistics
cat("\n=== Metabolism Time Series Trends (Bayesian Analysis) ===\n\n")
cat("Pmax (Maximum Photosynthetic Capacity):\n")
cat(sprintf("  Slope: %.3f mmol O₂ m⁻² hr⁻¹ yr⁻¹ [95%% CI: %.3f, %.3f]\n", 
    pmax_slope, pmax_ci[1], pmax_ci[2]))
cat(sprintf("  Bayesian R²: %.3f\n", pmax_r2))
cat(sprintf("  Credible interval excludes zero: %s\n", 
    ifelse(sign(pmax_ci[1]) == sign(pmax_ci[2]), "Yes", "No")))
cat(sprintf("  Total change (20 yr): %.1f mmol O₂ m⁻² hr⁻¹ (%.1f%%)\n", 
    pmax_slope * 20, 
    (pmax_slope * 20 / mean(Year_Averages$Pmax, na.rm = TRUE)) * 100))

cat("\nRd (Ecosystem Respiration):\n")
cat(sprintf("  Slope: %.3f mmol O₂ m⁻² hr⁻¹ yr⁻¹ [95%% CI: %.3f, %.3f]\n", 
    rd_slope, rd_ci[1], rd_ci[2]))
cat(sprintf("  Bayesian R²: %.3f\n", rd_r2))
cat(sprintf("  Credible interval excludes zero: %s\n", 
    ifelse(sign(rd_ci[1]) == sign(rd_ci[2]), "Yes", "No")))
cat(sprintf("  Total change (20 yr): %.1f mmol O₂ m⁻² hr⁻¹ (%.1f%%)\n", 
    rd_slope * 20, 
    (rd_slope * 20 / mean(Year_Averages$Rd, na.rm = TRUE)) * 100))

cat("\nNP (Net Ecosystem Production):\n")
cat(sprintf("  Slope: %.3f mmol O₂ m⁻² hr⁻¹ yr⁻¹ [95%% CI: %.3f, %.3f]\n", 
    np_slope, np_ci[1], np_ci[2]))
cat(sprintf("  Bayesian R²: %.3f\n", np_r2))
cat(sprintf("  Credible interval excludes zero: %s\n", 
    ifelse(sign(np_ci[1]) == sign(np_ci[2]), "Yes", "No")))
cat(sprintf("  Total change (20 yr): %.1f mmol O₂ m⁻² hr⁻¹\n", np_slope * 20))
```

**Key Findings:** Over the 20-year study period, all three metabolic components showed significant declines. Maximum photosynthetic capacity (Pmax) declined by `r round(pmax_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹ (`r round((pmax_slope * 20 / mean(Year_Averages$Pmax, na.rm = TRUE)) * 100, 1)`%), ecosystem respiration (Rd) decreased by `r round(rd_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹ (`r round((rd_slope * 20 / mean(Year_Averages$Rd, na.rm = TRUE)) * 100, 1)`%), and net ecosystem production (NP) declined by `r round(np_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹. These parallel declines in both autotrophic and heterotrophic processes indicate fundamental metabolic dysfunction accompanying the coral-to-algae phase shift.

## Production/Respiration Ratio: Ecosystem Maturity Indicator

```{r pr-ratio}
#| label: fig-pr-ratio
#| fig-cap: "Production to Respiration (P/R) ratio over time. Values >1 indicate net autotrophy (production exceeds respiration), while values <1 indicate net heterotrophy. Odum (1969) proposed P/R as an indicator of ecosystem maturity, with mature systems approaching P/R = 1. The increasing trend away from 1 reflects disproportionate loss of respiratory capacity (coral biomass) relative to production."
#| fig-width: 5
#| fig-height: 3

# Calculate P/R ratio
Year_Averages <- Year_Averages |>
  mutate(P_R_ratio = Pmax / Rd)

# Fit Bayesian model for P/R ratio trend
pr_brm <- brm(P_R_ratio ~ Year, data = Year_Averages, silent = 2, refresh = 0, seed = 223)

# Extract coefficients
pr_slope <- fixef(pr_brm)["Year", "Estimate"]
pr_ci <- fixef(pr_brm)["Year", c("Q2.5", "Q97.5")]
pr_r2 <- median(bayes_R2(pr_brm)[, 1])

# Create plot
pr_plot <- ggplot(Year_Averages, aes(x = Year, y = P_R_ratio)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "#7570B3",
    fill = "#7570B3",
    alpha = 0.15,
    linewidth = 1
  ) +
  geom_point(size = 3, color = "#7570B3", alpha = 0.7) +
  geom_line(color = "#7570B3", linewidth = 0.5, alpha = 0.4) +
  annotate(
    "text",
    x = 2008, y = Inf,
    label = sprintf("%.3f yr⁻¹ [95%% CI: %.3f, %.3f]\nBayesian R² = %.2f",
                    pr_slope, pr_ci[1], pr_ci[2], pr_r2),
    hjust = 0, vjust = 1.2,
    size = 4.5, color = "gray30"
  ) +
  annotate("text", x = 2024, y = 1.05, label = "Net Autotrophy (P > R)", 
           hjust = 1, size = 4, color = "gray40") +
  annotate("text", x = 2024, y = 0.95, label = "Net Heterotrophy (P < R)", 
           hjust = 1, size = 4, color = "gray40") +
  labs(
    x = "Year",
    y = "P/R Ratio (Pmax / Rd)",
    title = "Ecosystem Maturity Decline: Production/Respiration Ratio",
    subtitle = "Increasing P/R (moving away from 1) indicates loss of respiratory biomass"
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13),
    plot.title = element_text(size = 17, face = "bold", margin = margin(b = 5)),
    plot.subtitle = element_text(size = 13, color = "gray40", margin = margin(b = 15))
  ) +
  scale_x_continuous(breaks = seq(2007, 2025, by = 3),
                     minor_breaks = seq(2007, 2025, by = 1),
                     limits = c(2007, 2025))

pr_plot

# Print summary
cat("\n=== P/R Ratio Analysis ===\n\n")
cat(sprintf("Initial P/R (2008-2010): %.2f ± %.2f\n",
            mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE),
            sd(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE)))
cat(sprintf("Recent P/R (2023-2025): %.2f ± %.2f\n",
            mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE),
            sd(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE)))
cat(sprintf("Slope: %.4f yr⁻¹ [95%% CI: %.4f, %.4f]\n", pr_slope, pr_ci[1], pr_ci[2]))
cat(sprintf("Total change: %.2f over 20 years\n", pr_slope * 20))
cat(sprintf("Bayesian R²: %.3f\n", pr_r2))
cat("\nInterpretation: P/R ratio increasing from ~%.1f to ~%.1f indicates shift away\n",
    mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE),
    mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE))
cat("from mature ecosystem balance. The higher P/R reflects disproportionate loss of\n")
cat("respiratory biomass (coral holobiont) while production remains—ecosystem retrogression\n")
cat("through loss of high-biomass, high-metabolism structure (sensu Odum 1969).\n")
```

**Ecosystem Maturity Decline**: The P/R ratio provides an integrative measure of ecosystem development (Odum 1969). Mature ecosystems approach P/R ≈ 1, balancing production and respiration with high biomass and efficient nutrient cycling. The observed increase in P/R from `r round(mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE), 2)` to `r round(mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE), 2)` represents ecosystem retrogression through disproportionate loss of respiratory capacity. While mature coral reefs maintain P/R ≈ 1 through high biomass and tight metabolic coupling, the increasing P/R indicates loss of the high-respiration coral holobiont while production (largely from algae) persists, creating metabolic imbalance characteristic of degraded systems.

```{r gp-r-ratio}
#| label: fig-gp-r-ratio
#| fig-cap: "Gross Production to Respiration (GP/R) ratio over time, using measured gross production (GP_mean) from daily metabolism rather than light-saturated Pmax. Values >1 indicate net autotrophy. The GP/R ratio represents actual in situ metabolic balance under ambient light conditions, while Pmax/R represents potential production capacity under saturating light."
#| fig-width: 5
#| fig-height: 3

# Calculate GP/R ratio using absolute value of R_mean (respiration is negative)
Year_Averages <- Year_Averages |>
  mutate(GP_R_ratio = GP_mean / abs(R_mean))

# Fit Bayesian model for GP/R ratio trend
gpr_brm <- brm(GP_R_ratio ~ Year, data = Year_Averages, silent = 2, refresh = 0, seed = 223)

# Extract coefficients
gpr_slope <- fixef(gpr_brm)["Year", "Estimate"]
gpr_ci <- fixef(gpr_brm)["Year", c("Q2.5", "Q97.5")]
gpr_r2 <- median(bayes_R2(gpr_brm)[, 1])

# Create plot
gpr_plot <- ggplot(Year_Averages, aes(x = Year, y = GP_R_ratio)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray30", linewidth = 0.8) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "#E7298A",
    fill = "#E7298A",
    alpha = 0.15,
    linewidth = 1
  ) +
  geom_point(size = 3, color = "#E7298A", alpha = 0.7) +
  geom_line(color = "#E7298A", linewidth = 0.5, alpha = 0.4) +
  annotate(
    "text",
    x = 2008, y = Inf,
    label = sprintf("%.3f yr⁻¹ [95%% CI: %.3f, %.3f]\nBayesian R² = %.2f",
                    gpr_slope, gpr_ci[1], gpr_ci[2], gpr_r2),
    hjust = 0, vjust = 1.2,
    size = 4.5, color = "gray30"
  ) +
  annotate("text", x = 2024, y = 1.05, label = "Net Autotrophy (GP > R)", 
           hjust = 1, size = 4, color = "gray40") +
  annotate("text", x = 2024, y = 0.95, label = "Net Heterotrophy (GP < R)", 
           hjust = 1, size = 4, color = "gray40") +
  labs(
    x = "Year",
    y = "GP/R Ratio (Gross Production / |Respiration|)",
    title = "In Situ Metabolic Balance: Gross Production/Respiration Ratio",
    subtitle = "GP/R based on daily integrated metabolism under ambient light conditions"
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13),
    plot.title = element_text(size = 17, face = "bold", margin = margin(b = 5)),
    plot.subtitle = element_text(size = 13, color = "gray40", margin = margin(b = 15))
  ) +
  scale_x_continuous(breaks = seq(2007, 2025, by = 3),
                     minor_breaks = seq(2007, 2025, by = 1),
                     limits = c(2007, 2025))

gpr_plot

# Print summary
cat("\n=== GP/R Ratio Analysis (In Situ Metabolism) ===\n\n")
cat(sprintf("Initial GP/R (2008-2010): %.2f ± %.2f\n",
            mean(Year_Averages$GP_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE),
            sd(Year_Averages$GP_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE)))
cat(sprintf("Recent GP/R (2023-2025): %.2f ± %.2f\n",
            mean(Year_Averages$GP_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE),
            sd(Year_Averages$GP_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE)))
cat(sprintf("Slope: %.4f yr⁻¹ [95%% CI: %.4f, %.4f]\n", gpr_slope, gpr_ci[1], gpr_ci[2]))
cat(sprintf("Total change: %.2f over 20 years\n", gpr_slope * 20))
cat(sprintf("Bayesian R²: %.3f\n", gpr_r2))

cat("\nInterpretation: GP/R represents the actual metabolic balance under ambient light\n")
cat("conditions, while Pmax/R represents potential production capacity. Both ratios show\n")
cat("similar patterns, though GP/R is consistently lower than Pmax/R because GP reflects\n")
cat("daily integrated production (including low-light periods) rather than light-saturated\n")
cat("maximum capacity. The ecosystem consistently operates below its photosynthetic potential.\n")

# Compare Pmax/R vs GP/R
cat("\n=== Comparison: Pmax/R vs GP/R ===\n\n")
cat(sprintf("Mean Pmax/R: %.2f ± %.2f\n",
            mean(Year_Averages$P_R_ratio, na.rm = TRUE),
            sd(Year_Averages$P_R_ratio, na.rm = TRUE)))
cat(sprintf("Mean GP/R: %.2f ± %.2f\n",
            mean(Year_Averages$GP_R_ratio, na.rm = TRUE),
            sd(Year_Averages$GP_R_ratio, na.rm = TRUE)))
cat(sprintf("Ratio of Pmax/R to GP/R: %.2f\n",
            mean(Year_Averages$P_R_ratio, na.rm = TRUE) / mean(Year_Averages$GP_R_ratio, na.rm = TRUE)))
cat("\nThis indicates the ecosystem operates at approximately %.0f%% of its light-saturated\n",
    (mean(Year_Averages$GP_R_ratio, na.rm = TRUE) / mean(Year_Averages$P_R_ratio, na.rm = TRUE)) * 100)
cat("photosynthetic capacity under daily ambient light conditions.\n")
```

**GP/R vs Pmax/R Comparison**: The GP/R ratio (mean: `r round(mean(Year_Averages$GP_R_ratio, na.rm = TRUE), 2)`) is consistently lower than Pmax/R (mean: `r round(mean(Year_Averages$P_R_ratio, na.rm = TRUE), 2)`) because gross production integrates over full daily light cycles including dawn, dusk, and night periods, while Pmax represents maximum photosynthetic capacity under light saturation. The ratio of these two metrics (`r round(mean(Year_Averages$P_R_ratio, na.rm = TRUE) / mean(Year_Averages$GP_R_ratio, na.rm = TRUE), 1)`:1) indicates the degree to which the ecosystem operates below its photosynthetic potential under natural light regimes. Both metrics show similar temporal trends, confirming that metabolic imbalance is occurring at both potential and realized levels of ecosystem function.

# Statistical Analysis

## Temporal Trends

Assess standardized change per year using Bayesian linear models.

```{r temporal-trends}
#| label: temporal-trends
#| cache: true

# Create standardized dataset
std_data <- Year_Averages |>
  select(
    mean_coral, mean_fleshy, Pmax, Rd, N_percent,
    Herbivores, Corallivore, NP_mean, Max_temp, GP_mean,
    Nitrite_and_Nitrate, Phosphate
  ) |>
  mutate(across(everything(), ~ as.numeric(scale(.x)))) |>
  bind_cols(Year_Averages |> select(Year))

# Fit Bayesian models for each variable and extract R²
fit_temporal_model <- function(response_var, param_name) {
  form <- as.formula(paste(response_var, "~ Year"))
  mod <- brm(form, data = std_data, silent = 2, refresh = 0, seed = 223)
  
  # Extract posterior samples
  post <- as_tibble(as.matrix(mod)) |>
    select(Year = b_Year) |>
    mutate(Parameter = param_name)
  
  # Calculate Bayesian R²
  r2 <- median(bayes_R2(mod)[, 1])
  
  # Store model for later R² extraction
  list(posterior = post, model = mod, r2 = r2, param = param_name)
}

# Fit all models
temporal_models <- list(
  fit_temporal_model("mean_coral", "% Coral Cover"),
  fit_temporal_model("mean_fleshy", "% Macroalgae Cover"),
  fit_temporal_model("Corallivore", "Corallivore"),
  fit_temporal_model("Herbivores", "Herbivore/bioeroding fish"),
  fit_temporal_model("N_percent", "%N Content"),
  fit_temporal_model("Rd", "Ecosystem Respiration"),
  fit_temporal_model("Pmax", "Maximum Photosynthetic Capacity"),
  fit_temporal_model("NP_mean", "Net Ecosystem Production"),
  fit_temporal_model("Max_temp", "Max Temperature"),
  fit_temporal_model("Nitrite_and_Nitrate", "Nitrate + Nitrite")
)

# Extract posteriors for plotting
temporal_results <- temporal_models |>
  map_df(~.x$posterior)

# Print R² values
cat("\n=== Temporal Trend Model Performance (Bayesian R²) ===\n\n")
temporal_models |>
  map_df(~tibble(Parameter = .x$param, R2 = .x$r2)) |>
  arrange(desc(R2)) |>
  print(n = Inf)
```

```{r temporal-plot}
#| label: fig-temporal
#| fig-cap: "Standardized temporal trends and raw time-series for 10 reef ecosystem
#|   variables. (a) Posterior distributions of the standardized slope (change per year)
#|   from Bayesian linear models; vertical dashed line at zero; white tick marks indicate
#|   95% credible intervals. (b) Raw annual observations (points) with posterior median
#|   trendline and 95% credible interval ribbon on original measurement scales. Units:
#|   coral and macroalgae cover (%); fish biomass (g m\u207b\u00b2); N content (%);
#|   ecosystem metabolism (mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9); temperature (\u00b0C);
#|   nitrate+nitrite (\u00b5mol L\u207b\u00b9). Colour encodes direction: red = increasing,
#|   blue = decreasing."
#| fig-width: 8.9
#| fig-height: 7.5

library(patchwork)
library(showtext)
library(sysfonts)

# Noto Sans has full Unicode coverage (superscripts, µ, °, subscripts)
# Falls back gracefully if offline — font is cached after first download
font_add_google("Noto Sans", family = "noto")
showtext_auto(enable = TRUE)

# ── Nature-ready base theme (7 pt, classic, no gridlines) ──────────────────
theme_nature <- function(base_size = 7) {
  theme_classic(base_size = base_size) +
    theme(
      text             = element_text(family = "noto"),
      axis.text        = element_text(size = base_size, color = "black"),
      axis.title       = element_text(size = base_size + 0.5),
      axis.line        = element_line(linewidth = 0.35, color = "black"),
      axis.ticks       = element_line(linewidth = 0.35, color = "black"),
      strip.background = element_blank(),
      strip.text       = element_text(size = 6.5, face = "plain", color = "grey20"),
      panel.spacing    = unit(0.5, "lines"),
      legend.position  = "none",
      plot.tag         = element_text(face = "bold", size = 9)
    )
}

col_inc <- "#E64B35"   # NPG red  — increasing
col_dec <- "#4DBBD5"   # NPG teal — decreasing

# ── Summarise posterior slopes (order + color mapping) ─────────────────────
slope_summary <- temporal_results |>
  group_by(Parameter) |>
  summarise(
    median_slope = median(Year),
    direction    = if_else(median_slope > 0, "Increasing", "Decreasing"),
    .groups = "drop"
  ) |>
  arrange(median_slope)

param_levels <- slope_summary$Parameter

# Short names for panel a y-axis (strip units line from panel b labels)
short_name_map <- scale_params |>
  select(Parameter, label) |>
  mutate(short_name = str_remove(label, "\n.*"))

# ── Panel a: Posterior density ridges ──────────────────────────────────────
panel_a <- temporal_results |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(short_name_map, by = "Parameter") |>
  mutate(short_name = factor(short_name, levels = short_name_map$short_name[match(param_levels, short_name_map$Parameter)])) |>
  ggplot(aes(x = Year, y = short_name, fill = direction)) +
  geom_vline(
    xintercept = 0, linetype = "dashed",
    linewidth = 0.35, color = "grey55"
  ) +
  stat_density_ridges(
    quantile_lines = TRUE, quantiles = c(0.025, 0.975),
    alpha = 0.75, linewidth = 0.25, color = "white",
    scale = 0.88, rel_min_height = 0.01
  ) +
  scale_fill_manual(values = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  annotate("text", x =  0.06, y = 0.6, label = "Increasing \u2192",
           hjust = 0, size = 1.9, color = col_inc, fontface = "italic", family = "noto") +
  annotate("text", x = -0.06, y = 0.6, label = "\u2190 Decreasing",
           hjust = 1, size = 1.9, color = col_dec, fontface = "italic", family = "noto") +
  labs(x = "Standardized change per year", y = NULL, tag = "a") +
  lims(x = c(-0.35, 0.35)) +
  theme_nature() +
  theme(axis.text.y = element_text(size = 6.5))

# ── Back-transform fitted values and raw data to original scales ────────────
# Scale parameters (mean + SD) from unscaled Year_Averages
short_label_map <- tibble(
  var = c("mean_coral", "mean_fleshy", "Corallivore", "Herbivores",
          "N_percent", "Rd", "Pmax", "NP_mean", "Max_temp", "Nitrite_and_Nitrate"),
  Parameter = c("% Coral Cover", "% Macroalgae Cover", "Corallivore",
                "Herbivore/bioeroding fish", "%N Content", "Ecosystem Respiration",
                "Maximum Photosynthetic Capacity", "Net Ecosystem Production",
                "Max Temperature", "Nitrate + Nitrite"),
  label = c(
    "Coral cover\n(%)",
    "Macroalgae cover\n(%)",
    "Corallivore\n(g m\u207b\u00b2)",
    "Herbivores\n(g m\u207b\u00b2)",
    "N content\n(%)",
    "Respiration\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "Pmax\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "NEP\n(mmol O\u2082 m\u207b\u00b2 hr\u207b\u00b9)",
    "Max temperature\n(\u00b0C)",
    "Nitrate+nitrite\n(\u00b5mol L\u207b\u00b9)"
  )
)

scale_params <- tibble(
  var = short_label_map$var,
  mu  = map_dbl(short_label_map$var, ~mean(Year_Averages[[.x]], na.rm = TRUE)),
  sd  = map_dbl(short_label_map$var, ~sd(Year_Averages[[.x]],   na.rm = TRUE))
) |>
  left_join(short_label_map, by = "var")

label_levels <- scale_params$label[match(param_levels, scale_params$Parameter)]

# ── Extract fitted values + raw data from each Bayesian model ───────────────
get_time_series <- function(model_obj) {
  mod      <- model_obj$model
  param    <- model_obj$param
  resp_var <- as.character(formula(mod)$formula)[2]
  year_seq <- tibble(Year = seq(min(std_data$Year), max(std_data$Year), by = 0.5))
  fitted_df <- fitted(mod, newdata = year_seq) |>
    as_tibble() |>
    bind_cols(year_seq) |>
    mutate(Parameter = param)
  raw_df <- std_data |>
    select(Year, value = all_of(resp_var)) |>
    mutate(Parameter = param)
  list(fitted = fitted_df, raw = raw_df)
}

all_series <- map(temporal_models, get_time_series)

fitted_all <- map_df(all_series, ~.x$fitted) |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(scale_params |> select(Parameter, mu, sd, label), by = "Parameter") |>
  mutate(
    Estimate_orig = Estimate * sd + mu,
    Q2.5_orig     = Q2.5    * sd + mu,
    Q97.5_orig    = Q97.5   * sd + mu,
    label = factor(label, levels = label_levels)
  )

raw_all <- map_df(all_series, ~.x$raw) |>
  left_join(slope_summary |> select(Parameter, direction), by = "Parameter") |>
  left_join(scale_params |> select(Parameter, mu, sd, label), by = "Parameter") |>
  mutate(
    value_orig = value * sd + mu,
    label = factor(label, levels = label_levels)
  )

# ── Bayesian R² annotations ─────────────────────────────────────────────────
r2_df <- map_df(temporal_models, ~tibble(Parameter = .x$param, r2 = .x$r2)) |>
  left_join(scale_params |> select(Parameter, label), by = "Parameter") |>
  mutate(
    label    = factor(label, levels = label_levels),
    r2_label = paste0("R\u00b2 = ", round(r2, 2))
  )

# ── Panel b: Small-multiple time series on original scale ───────────────────
panel_b <- ggplot() +
  geom_ribbon(
    data = fitted_all,
    aes(x = Year, ymin = Q2.5_orig, ymax = Q97.5_orig, fill = direction),
    alpha = 0.18
  ) +
  geom_line(
    data = fitted_all,
    aes(x = Year, y = Estimate_orig, color = direction),
    linewidth = 0.55
  ) +
  geom_point(
    data = raw_all,
    aes(x = Year, y = value_orig),
    size = 0.85, alpha = 0.70, shape = 16, color = "grey25"
  ) +
  geom_text(
    data = r2_df,
    aes(label = r2_label),
    x = -Inf, y = Inf,
    hjust = -0.1, vjust = 1.4,
    size = 1.8, color = "grey45", fontface = "italic", family = "noto"
  ) +
  facet_wrap(~label, scales = "free_y", ncol = 5) +
  scale_fill_manual(values  = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  scale_color_manual(values = c("Increasing" = col_inc, "Decreasing" = col_dec)) +
  scale_x_continuous(breaks = c(2008, 2014, 2020)) +
  labs(x = "Year", y = NULL, tag = "b") +
  theme_nature() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 5.5),
    axis.text.y = element_text(size = 5.5)
  )

# ── Combine and render ───────────────────────────────────────────────────────
fig7 <- panel_a / panel_b +
  plot_layout(heights = c(1.05, 1.4)) &
  theme(plot.background = element_blank())

fig7

ggsave(
  here("Output", "fig7_temporal_trends.pdf"),
  plot   = fig7,
  width  = 183,
  height = 155,
  units  = "mm",
  device = cairo_pdf
)
```

**Model Performance Summary**: The Bayesian temporal trend models showed varying explanatory power across ecosystem variables. Temperature and coral cover trends had the highest R² values (>0.5), indicating highly predictable linear trajectories over time. Metabolic variables (Pmax, Rd, NP) showed moderate explanatory power (R² = 0.2-0.5), reflecting both directional trends and substantial year-to-year variability. Fish biomass trends had lower R² values, suggesting additional drivers beyond benthic habitat changes influence these populations.

## Percent Change Analysis

```{r percent-change}
#| label: fig-percent-change
#| fig-cap: "Percent change from start to end of time series"
#| fig-width: 10
#| fig-height: 6

# Helper functions for first/last non-NA values
get_first_non_na <- function(x) {
  x[which(!is.na(x))[1]]
}

get_last_non_na <- function(x) {
  non_na <- x[!is.na(x)]
  if (length(non_na) > 0) tail(non_na, 1) else NA
}

# Calculate percent changes
change_data <- Year_Averages |>
  select(
    NP_mean, mean_fleshy, mean_coral, Pmax, Rd,
    N_percent, Herbivores, Corallivore, Max_temp, Nitrite_and_Nitrate
  ) |>
  summarise(across(everything(), list(
    first = get_first_non_na,
    last = get_last_non_na
  ))) |>
  pivot_longer(everything()) %>%
  separate(col = name, into = c("Params", "stat"), sep = "_(?=first|last)", remove = TRUE) |>
  pivot_wider(names_from = stat, values_from = value) |>
  mutate(
    percent_change = (last - first) / first * 100,
    percent_change = if_else(Params == "NP_mean", -percent_change, percent_change)
  ) |>
  mutate(nicenames = case_when(
    Params == "mean_fleshy" ~ "% Macroalgae Cover",
    Params == "NP_mean" ~ "Net Ecosystem Production",
    Params == "Max_temp" ~ "Max Temperature",
    Params == "Herbivores" ~ "Herbivore/bioeroding fish",
    Params == "Corallivore" ~ "Corallivore",
    Params == "Pmax" ~ "Maximum Photosynthetic Capacity",
    Params == "N_percent" ~ "%N Content",
    Params == "Rd" ~ "Ecosystem Respiration",
    Params == "mean_coral" ~ "% Coral Cover",
    Params == "Nitrite_and_Nitrate" ~ "Nitrate + Nitrite"
  )) |>
  arrange(desc(percent_change))

# Calculate rates per year for annotation
meanyearmodels <- Year_Averages |>
  select(change_data$Params) |>
  summarise(across(everything(), mean, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "Params", values_to = "Mean")

sdyearmodels <- Year_Averages |>
  select(change_data$Params) |>
  summarise(across(everything(), sd, na.rm = TRUE)) |>
  pivot_longer(everything(), names_to = "Params", values_to = "SD")

Rates <- meanyearmodels |>
  left_join(sdyearmodels, by = "Params") |>
  mutate(nicenames = case_when(
    Params == "mean_coral" ~ "% Coral Cover",
    Params == "mean_fleshy" ~ "% Macroalgae Cover",
    Params == "N_percent" ~ "%N Content",
    Params == "Corallivore" ~ "Corallivore",
    Params == "Rd" ~ "Ecosystem Respiration",
    Params == "Herbivores" ~ "Herbivore/bioeroding fish",
    Params == "Max_temp" ~ "Max Temperature",
    Params == "Pmax" ~ "Maximum Photosynthetic Capacity",
    Params == "NP_mean" ~ "Net Ecosystem Production",
    Params == "Nitrite_and_Nitrate" ~ "Nitrate + Nitrite"
  )) |>
  drop_na() |>
  left_join(
    temporal_results |>
      group_by(Parameter) |>
      summarise(Rate = median(Year)),
    by = c("nicenames" = "Parameter")
  ) |>
  mutate(Rate_Year = round(Rate * SD, 2))

# Lollipop plot
change_data |>
  mutate(
    nicenames = factor(nicenames, levels = rev(unique(change_data$nicenames))),
    color = if_else(percent_change > 0, "pos", "neg")
  ) |>
  ggplot(aes(x = percent_change, y = nicenames)) +
  geom_segment(aes(x = 0, xend = percent_change, yend = nicenames), linewidth = 1.1) +
  geom_point(aes(color = color), size = 4) +
  geom_vline(xintercept = 0) +
  geom_text(data = Rates, aes(x = -120, y = nicenames, label = Rate_Year), size = 4) +
  scale_color_manual(values = c("#B22222", "#22B2B2")) +
  labs(
    x = "Percent change over time series (%)",
    y = "",
    title = "Ecosystem Change Magnitudes",
    subtitle = "Numbers on left show annual rate of change in original units"
  ) +
  lims(x = c(-130, 185)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 12),
    axis.title.x = element_text(size = 14)
  )
```

## Correlation Analysis

```{r correlations}
#| label: fig-correlations
#| fig-cap: "Correlations among ecosystem variables"
#| fig-width: 10
#| fig-height: 10

# Compute correlation matrix
ct <- corr.test(
  Year_Averages |>
    select(
      NP = NP_mean, Rd, Pmax, `%N` = N_percent,
      `N+N` = Nitrite_and_Nitrate, `% Coral` = mean_coral,
      `% Algae` = mean_fleshy, `Herbs/eroder` = Herbivores,
      `Corallivore` = Corallivore, `Max Temp` = Max_temp
    ),
  adjust = "none"
)

corr <- ct$r
p.mat <- ct$p

ggcorrplot.mixed(corr,
  upper = "ellipse",
  lower = "number",
  p.mat = p.mat,
  insig = "label_sig",
  sig.lvl = c(0.05, 0.01, 0.001)
) +
  labs(title = "Ecosystem Variable Correlations")
```

## Structural Equation Modeling

Fit a Bayesian SEM to understand causal pathways among ecosystem components.

```{r sem-prep}
#| label: sem-preparation

# Calculate year residuals (disturbances beyond temperature)
yearresid <- resid(lm(Year ~ Max_temp, data = Year_Averages, na.action = na.pass))

# Prepare scaled SEM data
semdata2 <- Year_Averages |>
  mutate(
    logfleshy = log(mean_fleshy),
    logrd = log(Rd),
    logpmax = log(Pmax),
    logproducers = log(mean_alive),
    N_percent = log(N_percent),
    yearresid = yearresid
  ) |>
  select(
    Year, meancoral = mean_coral, logproducers, meanfleshy = mean_fleshy,
    temperature = Max_temp, flow = Flow_mean, Npercent = N_percent,
    logrd, logpmax, logcoral = log_coral, logfleshy, Corallivore,
    herbs = Herbivores, yearresid
  ) |>
  drop_na(logcoral) |>
  mutate(across(everything(), ~ as.numeric(scale(.x))))
```

```{r sem-fit}
#| label: sem-model
#| cache: true

# Define SEM structure
bf_temp <- bf(temperature | mi() ~ 1 + Year)
bf_coraltemp <- bf(logcoral | trunc(ub = 3.39) ~ 1 + temperature #+ yearresid
                   )
bf_logpmax <- bf(logpmax | mi() ~ 1 + logcoral + mi(temperature) + mi(logrd)) ####
bf_logrd <- bf(logrd | mi() ~ 1 + logcoral + mi(temperature))
bf_herbs <- bf(herbs | mi() ~ 1 + logfleshy)
bf_corallivore <- bf(Corallivore | mi() ~ 1 + logcoral)
bf_fleshy <- bf(logfleshy | trunc(ub = 2.26) ~ logcoral)
bf_nutrients <- bf(Npercent | mi() ~ 1 + mi(logrd))

# Fit full SEM
brms_sem_full <- brm(
  bf_temp + bf_coraltemp + bf_logrd + bf_herbs + bf_corallivore +
    bf_nutrients + bf_fleshy + bf_logpmax + set_rescor(FALSE),
  data = semdata2,
  chains = 3,
  iter = 10000,
  warmup = 5000,
  seed = 11,
  sample_prior = "no",
  prior = c(
    set_prior("normal(0, 2)", class = "b"),
    set_prior("normal(0, 0.25)", resp = "temperature", class = "sigma", lb = 0),
    set_prior("normal(0, 2)", class = "Intercept")
  ),
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  family = gaussian(),
  silent = 2,
  refresh = 0
)
```

```{r sem-summary}
#| label: sem-results

summary(brms_sem_full)
```

```{r}
#| label: sem-r2-calculation

# Calculate Bayesian R² for each response variable in the SEM
# Note: Some responses with missing data imputation may have NaN - filter these out
sem_r2 <- bayes_R2(brms_sem_full, summary = FALSE)
sem_r2_summary <- apply(sem_r2, 2, median, na.rm = TRUE)

# Remove NaN values
sem_r2_valid <- sem_r2_summary[!is.na(sem_r2_summary)]

cat("\n=== Structural Equation Model: Bayesian R² by Response ===\n\n")
sem_r2_df <- tibble(
  Response = names(sem_r2_valid),
  R2 = sem_r2_valid
) |>
  mutate(Response = case_when(
    Response == "R2temperature" ~ "Temperature",
    Response == "R2logcoral" ~ "Coral Cover",
    Response == "R2logpmax" ~ "Ecosystem Pmax",
    Response == "R2logrd" ~ "Ecosystem Rd",
    Response == "R2herbs" ~ "Herbivorous Fish",
    Response == "R2Corallivore" ~ "Corallivore Fish",
    Response == "R2logfleshy" ~ "Fleshy Macroalgae",
    Response == "R2Npercent" ~ "N Content",
    TRUE ~ Response
  )) |>
  arrange(desc(R2))

print(sem_r2_df, n = Inf)

cat("\n")
cat(sprintf("Mean R² across computable responses: %.3f\n", mean(sem_r2_valid)))
cat(sprintf("The SEM explains between %.1f%% and %.1f%% of variance across responses.\n",
            min(sem_r2_valid) * 100, max(sem_r2_valid) * 100))
cat("Note: R² not computed for responses with extensive missing data imputation.\n")
```

## SEM Conceptual Model

```{r sem-dag}
#| label: fig-sem-dag
#| fig-cap: "Causal pathways in the Bayesian SEM. Arrow width is proportional to the standardized path coefficient (|β|); blue = positive effect, red = negative effect; gray = 95% credible interval crosses zero."
#| fig-width: 10
#| fig-height: 7

pos_col  <- "#1565C0"
neg_col  <- "#B71C1C"
nsig_col <- "#AAAAAA"

dag_nodes <- tribble(
  ~id,           ~label,                    ~x,     ~y,    ~fill,       ~border,
  "Year",        "Year",                     0,     3.0,   "white",     "#888888",
  "Temp",        "Temperature",              2.2,   3.0,   "white",     "#888888",
  "Coral",       "Coral\nCover",             4.5,   3.0,   "#E8F5E9",   "#2E7D32",
  "Algae",       "Fleshy\nAlgae",            7.2,   5.2,   "#E8F5E9",   "#2E7D32",
  "Rd",          "Ecosystem\nRespiration",   7.2,   3.0,   "#EDE7F6",   "#4527A0",
  "Pmax",        "Ecosystem\nGPP",           7.2,   0.8,   "#EDE7F6",   "#4527A0",
  "N",           "Algal N\nContent",        10.0,   3.0,   "#FFF8E1",   "#E65100",
  "Corallivore", "Corallivore\nFish",        9.8,   4.5,   "#E3F2FD",   "#0D47A1",
  "Herbs",       "Herbivore\nFish",         10.0,   5.2,   "#E3F2FD",   "#0D47A1"
)

dag_edges <- tribble(
  ~from,   ~to,           ~est,   ~sig,  ~loff_x, ~loff_y,
  "Year",  "Temp",        0.462,  TRUE,   0.0,     0.25,
  "Temp",  "Coral",      -0.558,  TRUE,   0.0,     0.25,
  "Temp",  "Rd",          0.214,  FALSE,  0.0,     0.0,
  "Temp",  "Pmax",       -0.158,  FALSE,  0.0,     0.0,
  "Coral", "Algae",      -0.914,  TRUE,  -0.3,     0.0,
  "Coral", "Rd",          1.080,  TRUE,   0.0,     0.30,
  "Coral", "Corallivore", 0.645,  TRUE,   0.2,     0.2,
  "Coral", "Pmax",       -0.353,  FALSE, -0.3,     0.0,
  "Rd",    "N",           0.623,  TRUE,   0.0,     0.30,
  "Rd",    "Pmax",        0.819,  TRUE,  -0.30,    0.0,
  "Algae", "Herbs",       0.711,  TRUE,   0.0,     0.30
) |>
  mutate(
    direction = if_else(est >= 0, "pos", "neg"),
    abs_est   = abs(est)
  ) |>
  left_join(dag_nodes |> select(id, x, y) |> rename(from = id, x0 = x, y0 = y), by = "from") |>
  left_join(dag_nodes |> select(id, x, y) |> rename(to   = id, x1 = x, y1 = y), by = "to") |>
  mutate(mx = (x0 + x1) / 2, my = (y0 + y1) / 2, lbl = sprintf("%.2f", est))

dag_straight    <- dag_edges |> filter(!(from == "Temp" & to %in% c("Rd", "Pmax")))
dag_curved_rd   <- dag_edges |> filter(from == "Temp", to == "Rd")
dag_curved_pmax <- dag_edges |> filter(from == "Temp", to == "Pmax")

dag_arr <- arrow(length = unit(3.5, "mm"), type = "closed")

ggplot() +
  geom_segment(data = filter(dag_straight,  sig),
    aes(x = x0, y = y0, xend = x1, yend = y1, color = direction, linewidth = abs_est),
    alpha = 0.82, arrow = dag_arr, lineend = "butt") +
  geom_segment(data = filter(dag_straight, !sig),
    aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = abs_est),
    color = nsig_col, alpha = 0.70, arrow = dag_arr, lineend = "butt") +
  geom_label(data = filter(dag_straight,  sig),
    aes(x = mx + loff_x, y = my + loff_y, label = lbl, color = direction),
    size = 3.0, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.10, "lines"), show.legend = FALSE) +
  geom_label(data = filter(dag_straight, !sig),
    aes(x = mx + loff_x, y = my + loff_y, label = lbl),
    color = nsig_col, size = 3.0, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.10, "lines"), show.legend = FALSE) +
  geom_curve(data = dag_curved_rd,
    aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = abs_est),
    color = nsig_col, alpha = 0.70, curvature = -0.28, arrow = dag_arr, lineend = "butt") +
  geom_label(data = dag_curved_rd,
    aes(x = mx + 0.3, y = my + 0.85, label = lbl),
    color = nsig_col, size = 3.0, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.10, "lines"), show.legend = FALSE) +
  geom_curve(data = dag_curved_pmax,
    aes(x = x0, y = y0, xend = x1, yend = y1, linewidth = abs_est),
    color = nsig_col, alpha = 0.70, curvature = 0.30, arrow = dag_arr, lineend = "butt") +
  geom_label(data = dag_curved_pmax,
    aes(x = mx - 0.1, y = my - 0.85, label = lbl),
    color = nsig_col, size = 3.0, fontface = "bold", fill = "white",
    label.size = 0, label.padding = unit(0.10, "lines"), show.legend = FALSE) +
  {lapply(seq_len(nrow(dag_nodes)), function(i) {
    nd <- dag_nodes[i, ]
    geom_label(data = nd, aes(x = x, y = y, label = label),
               fill = nd$fill, color = nd$border, size = 3.2,
               fontface = "bold", lineheight = 0.9, label.size = 0.75,
               label.padding = unit(0.45, "lines"), label.r = unit(5, "pt"),
               show.legend = FALSE)
  })} +
  scale_color_manual(values = c("pos" = pos_col, "neg" = neg_col),
                     labels = c("pos" = "Positive", "neg" = "Negative"),
                     name = "Effect direction") +
  scale_linewidth(range = c(0.8, 3.2), guide = "none") +
  labs(caption = "Standardized β on arrows  ·  gray = 95% CI crosses zero") +
  theme_void(base_size = 12) +
  theme(
    plot.caption    = element_text(hjust = 0.5, color = "grey55", size = 9, margin = margin(t = 8)),
    legend.position = "bottom",
    legend.title    = element_text(size = 10, face = "bold"),
    legend.text     = element_text(size = 9),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin     = margin(15, 20, 10, 20)
  ) +
  guides(color = guide_legend(override.aes = list(linewidth = 1.5))) +
  coord_cartesian(xlim = c(-0.5, 11.5), ylim = c(-0.3, 6.3))
```

## SEM Path Coefficients

```{r sem-paths}
#| label: fig-sem-paths
#| fig-cap: "Standardized effect sizes from structural equation model"
#| fig-width: 16
#| fig-height: 6

# Extract and format path coefficients
gp_edges <- as_tibble(fixef(brms_sem_full)) |>
  mutate(coef = rownames(fixef(brms_sem_full))) |>
  separate(col = coef, into = c("resp", "pred")) |>
  filter(pred != "Intercept") |>
  rename(est = Estimate, lo = `Q2.5`, hi = `Q97.5`) |>
  mutate(sig = if_else(sign(lo) == sign(hi), 1, 0.5)) |>
  mutate(resp = case_when(
    resp == "temperature" ~ "Temperature",
    resp == "logfleshy" ~ "Fleshy Macroalgae",
    resp == "logpmax" ~ "Ecosystem Pmax",
    resp == "logrd" ~ "Ecosystem Rd",
    resp == "herbs" ~ "Herbivorous Fish",
    resp == "Corallivore" ~ "Corallivore Fish",
    resp == "logcoral" ~ "Coral",
    resp == "Npercent" ~ "N stock"
  )) |>
  mutate(resp = factor(resp, c(
    "Temperature", "Coral", "Fleshy Macroalgae", "Herbivorous Fish",
    "Corallivore Fish", "Ecosystem Rd", "Ecosystem Pmax", "N stock"
  ))) |>
  mutate(pred = case_when(
    pred == "temperature" ~ "Temperature",
    pred == "mitemperature" ~ "Temperature",
    pred == "logproducers" ~ "Total Producer Cover",
    pred == "milogrd" ~ "Ecosystem Rd Rate",
    pred == "logcoral" ~ "Coral Cover",
    pred == "logfleshy" ~ "Fleshy Macroalgae Cover",
    pred == "Year" ~ "Year",
    pred == "yearresid" ~ "Non-Temperature Yearly Variability"
  )) |>
  mutate(pred = factor(pred, c(
    "Ecosystem Rd Rate", "Total Producer Cover",
    "Fleshy Macroalgae Cover", "Coral Cover", "Non-Temperature Yearly Variability",
    "Temperature", "Year"
  )))

gp_edges |>
  ggplot(aes(x = est, y = pred)) +
  geom_point(size = 3, color = "firebrick", aes(alpha = sig)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi, alpha = sig), height = 0, color = "firebrick") +
  geom_vline(xintercept = 0) +
  scale_alpha(range = c(0.25, 1)) +
  labs(
    x = "Standardized Effect Size",
    y = "Predictor Variables",
    title = "SEM Path Coefficients"
  ) +
  theme_minimal() +
  facet_wrap(~resp, nrow = 1) +
  theme(
    text = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.position = "none"
  )
```

## Temperature Effects: Direct vs Indirect

Calculate the total, direct, and indirect effects of temperature on all ecosystem responses.

```{r temp-effects}
#| label: fig-temp-effects
#| fig-cap: "Direct and indirect effects of temperature on ecosystem components"
#| fig-width: 10
#| fig-height: 8
#| cache: true

# Extract posterior samples for path coefficients
post_samples <- as_tibble(as.matrix(brms_sem_full))

# Calculate indirect effects of temperature through coral cover
# Temperature -> Coral -> [Response]

# Direct effects (already in the model)
# Temperature -> Coral
temp_to_coral <- post_samples$b_logcoral_temperature

# Temperature -> Pmax
temp_to_pmax <- post_samples$bsp_logpmax_mitemperature

# Temperature -> Rd
temp_to_rd <- post_samples$bsp_logrd_mitemperature

# Indirect effects through coral
# Coral -> Pmax
coral_to_pmax <- post_samples$b_logpmax_logcoral

# Coral -> Rd
coral_to_rd <- post_samples$b_logrd_logcoral

# Coral -> Fleshy Algae
coral_to_fleshy <- post_samples$b_logfleshy_logcoral

# Coral -> Corallivores
coral_to_corallivore <- post_samples$b_Corallivore_logcoral

# Fleshy Algae -> Herbivores
fleshy_to_herbs <- post_samples$b_herbs_logfleshy

# Rd -> N content
rd_to_n <- post_samples$bsp_Npercent_milogrd

# Calculate total and indirect effects
effects_data <- tibble(
  # Coral (direct only)
  response = c("Coral Cover", "Coral Cover",
               # Pmax (direct + indirect through coral)
               "Ecosystem Pmax", "Ecosystem Pmax",
               # Rd (direct + indirect through coral)
               "Ecosystem Rd", "Ecosystem Rd",
               # Fleshy Algae (indirect through coral)
               "Fleshy Macroalgae", "Fleshy Macroalgae",
               # Corallivores (indirect through coral)
               "Corallivore Fish", "Corallivore Fish",
               # Herbivores (indirect through coral -> algae)
               "Herbivore Fish", "Herbivore Fish",
               # N content (indirect through coral -> Rd)
               "N Content", "N Content"),
  
  effect_type = rep(c("Direct", "Indirect"), 7),
  
  samples = list(
    # Coral
    temp_to_coral,  # Direct
    rep(0, length(temp_to_coral)),  # No indirect (coral is first response)
    
    # Pmax
    temp_to_pmax,  # Direct
    temp_to_coral * coral_to_pmax,  # Indirect: Temp -> Coral -> Pmax
    
    # Rd
    temp_to_rd,  # Direct
    temp_to_coral * coral_to_rd,  # Indirect: Temp -> Coral -> Rd
    
    # Fleshy Algae
    rep(0, length(temp_to_coral)),  # No direct effect
    temp_to_coral * coral_to_fleshy,  # Indirect: Temp -> Coral -> Algae
    
    # Corallivores
    rep(0, length(temp_to_coral)),  # No direct effect
    temp_to_coral * coral_to_corallivore,  # Indirect: Temp -> Coral -> Corallivore
    
    # Herbivores
    rep(0, length(temp_to_coral)),  # No direct effect
    temp_to_coral * coral_to_fleshy * fleshy_to_herbs,  # Indirect: Temp -> Coral -> Algae -> Herbs
    
    # N content
    rep(0, length(temp_to_coral)),  # No direct effect
    temp_to_coral * coral_to_rd * rd_to_n  # Indirect: Temp -> Coral -> Rd -> N
  )
)

# Calculate summary statistics for each effect
effects_summary <- effects_data |>
  mutate(
    median = map_dbl(samples, median),
    lower = map_dbl(samples, ~quantile(.x, 0.025)),
    upper = map_dbl(samples, ~quantile(.x, 0.975)),
    significant = map_lgl(samples, ~sign(quantile(.x, 0.025)) == sign(quantile(.x, 0.975)))
  ) |>
  mutate(response = factor(response, levels = c(
    "Coral Cover", "Fleshy Macroalgae", "Herbivore Fish",
    "Corallivore Fish", "Ecosystem Rd", "Ecosystem Pmax", "N Content"
  )))

# Calculate total effects
total_effects <- effects_summary |>
  select(response, effect_type, samples) |>
  pivot_wider(names_from = effect_type, values_from = samples) |>
  mutate(Total = map2(Direct, Indirect, ~.x + .y)) |>
  select(response, Total) |>
  mutate(
    effect_type = "Total",
    median = map_dbl(Total, median),
    lower = map_dbl(Total, ~quantile(.x, 0.025)),
    upper = map_dbl(Total, ~quantile(.x, 0.975)),
    significant = map_lgl(Total, ~sign(quantile(.x, 0.025)) == sign(quantile(.x, 0.975)))
  ) |>
  rename(samples = Total)

# Combine all effects
all_effects <- bind_rows(effects_summary, total_effects) |>
  mutate(effect_type = factor(effect_type, levels = c("Direct", "Indirect", "Total")))

# Create publication-quality plot
all_effects |>
  filter(effect_type != "Total") |>  # Show only direct and indirect for clarity
  ggplot(aes(x = median, y = response, color = effect_type, shape = effect_type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(
    aes(xmin = lower, xmax = upper, alpha = significant),
    position = position_dodge(width = 0.5),
    size = 0.8,
    linewidth = 1
  ) +
  scale_color_manual(
    values = c("Direct" = "#E64B35", "Indirect" = "#4DBBD5"),
    name = "Effect Type"
  ) +
  scale_shape_manual(
    values = c("Direct" = 16, "Indirect" = 17),
    name = "Effect Type"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 1, "FALSE" = 0.4),
    guide = "none"
  ) +
  labs(
    x = "Standardized Effect Size",
    y = "",
    title = "Temperature Effects on Ecosystem Components",
    subtitle = "Direct effects and indirect effects mediated through coral decline"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.y = element_text(size = 13),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14, margin = margin(t = 10)),
    legend.position = "bottom",
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray40")
  )

# Print summary table
cat("\n=== Temperature Effects Summary ===\n\n")
all_effects |>
  select(response, effect_type, median, lower, upper, significant) |>
  mutate(across(median:upper, ~round(.x, 3))) |>
  arrange(response, effect_type) |>
  print(n = Inf)
```

```{r temp-indirect-ranking}
#| label: fig-temp-indirect
#| fig-cap: "Ranking of indirect temperature effects on ecosystem components"
#| fig-width: 8
#| fig-height: 6

# Extract only indirect effects and rank them
indirect_effects <- all_effects |>
  filter(effect_type == "Indirect") |>
  # Remove coral (no indirect effect)
  filter(response != "Coral Cover") |>
  arrange(median) |>
  mutate(
    response = fct_reorder(response, median),
    direction = if_else(median < 0, "Negative", "Positive")
  )

# Create ranked bar plot
indirect_effects |>
  ggplot(aes(x = response, y = median, fill = direction, alpha = significant)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30", linewidth = 0.5) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.1,
    linewidth = 0.8,
    color = "gray20",
    alpha = 0.6
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Negative" = "#E64B35", "Positive" = "#4DBBD5"),
    name = "Effect Direction"
  ) +
  scale_alpha_manual(
    values = c("TRUE" = 0.9, "FALSE" = 0.2),
    guide = "none"
  ) +
  labs(
    x = "",
    y = "Standardized Indirect Effect Size",
    title = "Indirect Effects of Temperature via Coral Decline",
  #  subtitle = "Ranked from negative to positive effects with 95% credible intervals"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.y = element_text(size = 16, hjust = 1),
    axis.text.x = element_text(size = 15),
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
   # legend.position = "bottom",
    legend.position = "none",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 19, face = "bold"),
    plot.subtitle = element_text(size = 15, color = "gray40")
  )

# Print ranked summary
cat("\n=== Ranked Indirect Effects (negative to positive) ===\n\n")
indirect_effects |>
  select(response, median, lower, upper, significant) |>
  mutate(across(median:upper, ~round(.x, 3))) |>
  arrange(median) |>
  print(n = Inf)
```

## SEM Conditional Effects

```{r sem-conditional}
#| label: fig-sem-conditional
#| fig-cap: "Key relationships from structural equation model"
#| fig-width: 10
#| fig-height: 18

# Get unscaling parameters
Yearmodelvalues <- Year_Averages |>
  mutate(
    logfleshy = log(mean_fleshy),
    logproducers = log(mean_alive),
    logrd = log(Rd),
    logpmax = log(Pmax),
    N_percent = log(N_percent),
    yearresid = yearresid
  ) |>
  select(
    Year, meancoral = mean_coral, meanfleshy = mean_fleshy,
    temperature = Max_temp, flow = Flow_mean, Npercent = N_percent,
    logrd, logpmax, logcoral = log_coral, logproducers,
    logfleshy, Corallivore, herbs = Herbivores, yearresid
  ) |>
  drop_na(logcoral)

modelmean <- Yearmodelvalues |>
  summarise(across(everything(), mean, na.rm = TRUE))

modelsd <- Yearmodelvalues |>
  summarise(across(everything(), sd, na.rm = TRUE))

# Year -> Temperature
temp_year <- conditional_effects(brms_sem_full, resp = "temperature", effects = "Year")
ty_plot <- as_tibble(temp_year$temperature.temperature_Year) |>
  mutate(
    Year = Year * modelsd$Year + modelmean$Year,
    estimate__ = estimate__ * modelsd$temperature + modelmean$temperature,
    lower__ = lower__ * modelsd$temperature + modelmean$temperature,
    upper__ = upper__ * modelsd$temperature + modelmean$temperature
  ) |>
  ggplot(aes(x = Year, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = Year, y = Max_temp), alpha = 0.5) +
  labs(
    x = "Year",
    y = "Temperature (°C)",
    title = "G. Year → Temperature"
  )

# Temperature -> Coral
coral_temp <- conditional_effects(brms_sem_full, resp = "logcoral", effects = "temperature")
ct_plot <- as_tibble(coral_temp$logcoral.logcoral_temperature) |>
  mutate(
    temperature = temperature * modelsd$temperature + modelmean$temperature,
    estimate__ = exp(estimate__ * modelsd$logcoral + modelmean$logcoral),
    lower__ = exp(lower__ * modelsd$logcoral + modelmean$logcoral),
    upper__ = exp(upper__ * modelsd$logcoral + modelmean$logcoral)
  ) |>
  ggplot(aes(x = temperature, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = Max_temp, y = exp(log_coral)), alpha = 0.5) +
  coord_trans(y = "log") +
  scale_y_continuous(breaks = c(1, 5, 10, 20, 40)) +
  labs(
    x = "Temperature (°C)",
    y = "Coral Cover (%)",
    title = "E. Temperature → Coral"
  )

# Coral -> Respiration
rd_coral <- conditional_effects(brms_sem_full, resp = "logrd", effects = "logcoral")
rc_plot <- as_tibble(rd_coral$logrd.logrd_logcoral) |>
  mutate(
    logcoral = exp(logcoral * modelsd$logcoral + modelmean$logcoral),
    estimate__ = exp(estimate__ * modelsd$logrd + modelmean$logrd),
    lower__ = exp(lower__ * modelsd$logrd + modelmean$logrd),
    upper__ = exp(upper__ * modelsd$logrd + modelmean$logrd)
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = Rd), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = "Coral Cover (%)",
    y = expression(atop("Ecosystem Respiration", "(mmol O"[2]*" m"^-2*" hr"^-1*")")),
    title = "B. Coral → Respiration"
  )

# Coral -> Fleshy Algae
fleshy_coral <- conditional_effects(brms_sem_full, resp = "logfleshy", effects = "logcoral")
fc_plot <- as_tibble(fleshy_coral$logfleshy.logfleshy_logcoral) |>
  mutate(
    logcoral = exp(logcoral * modelsd$logcoral + modelmean$logcoral),
    estimate__ = exp(estimate__ * modelsd$logfleshy + modelmean$logfleshy),
    lower__ = exp(lower__ * modelsd$logfleshy + modelmean$logfleshy),
    upper__ = exp(upper__ * modelsd$logfleshy + modelmean$logfleshy)
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = mean_fleshy), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = "Coral Cover (%)",
    y = "Fleshy Macroalgae Cover (%)",
    title = "F. Coral → Macroalgae"
  )

# Coral -> Corallivores
corallivore_coral <- conditional_effects(brms_sem_full, resp = "Corallivore", effects = "logcoral")
cc_plot <- as_tibble(corallivore_coral$Corallivore.Corallivore_logcoral) |>
  mutate(
    logcoral = exp(logcoral * modelsd$logcoral + modelmean$logcoral),
    estimate__ = estimate__ * modelsd$Corallivore + modelmean$Corallivore,
    lower__ = lower__ * modelsd$Corallivore + modelmean$Corallivore,
    upper__ = upper__ * modelsd$Corallivore + modelmean$Corallivore
  ) |>
  ggplot(aes(x = logcoral, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = exp(log_coral), y = Corallivore), alpha = 0.5) +
  coord_trans(x = "log") +
  labs(
    x = "Coral Cover (%)",
    y = expression(atop("Corallivore Fish Biomass", "(g m"^-2*")")),
    title = "C. Coral → Corallivores"
  )

# Fleshy Algae -> Herbivores
herbs_fleshy <- conditional_effects(brms_sem_full, resp = "herbs", effects = "logfleshy")
hf_plot <- as_tibble(herbs_fleshy$herbs.herbs_logfleshy) |>
  mutate(
    logfleshy = exp(logfleshy * modelsd$logfleshy + modelmean$logfleshy),
    estimate__ = estimate__ * modelsd$herbs + modelmean$herbs,
    lower__ = lower__ * modelsd$herbs + modelmean$herbs,
    upper__ = upper__ * modelsd$herbs + modelmean$herbs
  ) |>
  ggplot(aes(x = logfleshy, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = mean_fleshy, y = Herbivores), alpha = 0.5) +
  coord_trans(x = "log") +
  labs(
    x = "Fleshy Macroalgae Cover (%)",
    y = expression(atop("Herbivore Fish Biomass", "(g m"^-2*")")),
    title = "D. Macroalgae → Herbivores"
  )

# Respiration -> N Content
N_Rd <- conditional_effects(brms_sem_full, resp = "Npercent", effects = "logrd")
nr_plot <- as_tibble(N_Rd$Npercent.Npercent_logrd) |>
  mutate(
    logrd = exp(logrd * modelsd$logrd + modelmean$logrd),
    estimate__ = exp(estimate__ * modelsd$Npercent + modelmean$Npercent),
    lower__ = exp(lower__ * modelsd$Npercent + modelmean$Npercent),
    upper__ = exp(upper__ * modelsd$Npercent + modelmean$Npercent)
  ) |>
  ggplot(aes(x = logrd, y = estimate__)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.25, fill = "#67bed9") +
  geom_point(data = Year_Averages, aes(x = Rd, y = N_percent), alpha = 0.5) +
  coord_trans(x = "log", y = "log") +
  labs(
    x = expression("Ecosystem Respiration (mmol O"[2]*" m"^-2*" hr"^-1*")"),
    y = "Nitrogen Content (%)",
    title = "A. Respiration → N Stock"
  )

# Combine plots with centering for single-plot rows
# Using plot_layout(widths = c(1, 2, 1)) to ensure centered plots are same size as others
((plot_spacer() | nr_plot | plot_spacer()) + plot_layout(widths = c(1, 2, 1))) /
  ((plot_spacer() | rc_plot | plot_spacer()) + plot_layout(widths = c(1, 2, 1))) /
  (cc_plot + hf_plot) /
  (ct_plot + fc_plot) /
  ((plot_spacer() | ty_plot | plot_spacer()) + plot_layout(widths = c(1, 2, 1))) +
  plot_layout(heights = c(1, 1, 1, 1, 1))
```

## Future Predictions

Project ecosystem state to 2050 using the SEM.

```{r predictions}
#| label: fig-predictions
#| fig-cap: "Ecosystem projections to 2050"
#| fig-width: 8
#| fig-height: 10
#| cache: true

# Define forecast years
test_years <- c(2010, 2020, 2030, 2040, 2050)
test_years_scale <- (test_years - modelmean$Year) / modelsd$Year

# Hindcast data for validation
hind_values <- Year_Averages |>
  filter(Year %in% c(2010, 2020))

# Predict temperature
new_data <- data.frame(Year = test_years_scale)
pred_new <- posterior_predict(brms_sem_full, newdata = new_data, resp = "temperature")

Temp_pred_plot <- as_tibble(pred_new * modelsd$temperature + modelmean$temperature) |>
  pivot_longer(cols = everything()) |>
  mutate(year = case_when(
    name == "V1" ~ "2010", name == "V2" ~ "2020",
    name == "V3" ~ "2030", name == "V4" ~ "2040", name == "V5" ~ "2050"
  )) |>
  ggplot(aes(y = value, x = year)) +
  stat_halfeye(
    point_interval = mean_hdi, .width = c(.95, .75),
    fatten_point = 2, slab_alpha = 0.6, fill = scales::alpha("#009E73", 0.6)
  ) +
  geom_point(
    data = hind_values, aes(x = as.character(Year), y = Max_temp),
    size = 4, color = "firebrick"
  ) +
  labs(x = "Year", y = "Temperature (°C)", title = "A. Temperature") +
  lims(y = c(28, 33))

# Predict coral cover
modyear <- lm(yearresid ~ Year, data = semdata2)
test_years_resid <- predict(modyear, newdata = data.frame(Year = test_years_scale))
newdata_temp <- data.frame(temperature = colMeans(pred_new), yearresid = test_years_resid)
pred_new_coral <- posterior_predict(brms_sem_full, newdata = newdata_temp, resp = "logcoral")

Coral_pred_plot <- exp(as_tibble(pred_new_coral * modelsd$logcoral + modelmean$logcoral)) |>
  pivot_longer(cols = everything()) |>
  mutate(year = case_when(
    name == "V1" ~ "2010", name == "V2" ~ "2020",
    name == "V3" ~ "2030", name == "V4" ~ "2040", name == "V5" ~ "2050"
  )) |>
  ggplot(aes(y = value, x = year)) +
  stat_halfeye(
    point_interval = mean_hdi, .width = c(.95, .75),
    fatten_point = 2, slab_alpha = 0.6, fill = scales::alpha("#009E73", 0.6)
  ) +
  geom_point(
    data = hind_values, aes(x = as.character(Year), y = mean_coral),
    size = 4, color = "firebrick"
  ) +
  labs(x = "Year", y = "Coral Cover (%)", title = "B. Coral Cover") +
  lims(y = c(0, 25))

# Predict Rd
newdata_logcoral <- data.frame(
  logcoral = colMeans(pred_new_coral),
  temperature = newdata_temp$temperature
)
pred_new_Rd <- posterior_predict(brms_sem_full, newdata = newdata_logcoral, resp = "logrd")

Rd_pred_plot <- exp(as_tibble(pred_new_Rd * modelsd$logrd + modelmean$logrd)) |>
  pivot_longer(cols = everything()) |>
  mutate(year = case_when(
    name == "V1" ~ "2010", name == "V2" ~ "2020",
    name == "V3" ~ "2030", name == "V4" ~ "2040", name == "V5" ~ "2050"
  )) |>
  ggplot(aes(y = value, x = year)) +
  stat_halfeye(
    point_interval = mean_hdi, .width = c(.95, .75),
    fatten_point = 2, slab_alpha = 0.6, fill = scales::alpha("#009E73", 0.6)
  ) +
  geom_point(
    data = hind_values, aes(x = as.character(Year), y = Rd),
    size = 4, color = "firebrick"
  ) +
  labs(
    x = "Year",
    y = expression(atop("Ecosystem Respiration", "(mmol O"[2]*" m"^-2*" hr"^-1*")")),
    title = "C. Respiration"
  ) +
  lims(y = c(0, 80))

# Predict N content
newdata_N <- data.frame(logrd = colMeans(pred_new_Rd))
pred_new_N <- posterior_epred(brms_sem_full, newdata = newdata_N, resp = "Npercent")

N_pred_plot <- exp(as_tibble(pred_new_N * modelsd$Npercent + modelmean$Npercent)) |>
  pivot_longer(cols = everything()) |>
  mutate(year = case_when(
    name == "V1" ~ "2010", name == "V2" ~ "2020",
    name == "V3" ~ "2030", name == "V4" ~ "2040", name == "V5" ~ "2050"
  )) |>
  ggplot(aes(y = value, x = year)) +
  stat_halfeye(
    point_interval = mean_hdi, .width = c(.95, .75),
    fatten_point = 2, slab_alpha = 0.6, fill = scales::alpha("#009E73", 0.6)
  ) +
  geom_point(
    data = hind_values, aes(x = as.character(Year), y = N_percent),
    size = 4, color = "firebrick"
  ) +
  labs(x = "Year", y = "N Content (%)", title = "D. Nitrogen Stock")

# Combine predictions
(Temp_pred_plot / Coral_pred_plot / Rd_pred_plot / N_pred_plot)
```

# Results & Interpretation

## Key Findings

### 1. Community Shifts

The backreef at LTER Site 1 has undergone dramatic community reorganization:

-   **Coral cover declined by `r round((1 - (Benthic_summary_Algae |> filter(Year == 2025, name == "Coral", Site == "LTER 1") |> pull(mean_cover)) / (Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover))) * 100, 1)`%** from 2006 to 2025
-   Fleshy macroalgae increased substantially
-   Fish community composition shifted with changes in habitat structure

### 2. Environmental Drivers

Temperature has increased significantly and appears to be the primary driver of coral decline through both direct thermal stress and indirect effects mediated through disturbance events. **Effect size interpretation**: The observed temperature increase of ~1°C in maximum daily temperature over 20 years translates to approximately 0.05°C per year. Based on the SEM standardized coefficients, each 1°C increase in temperature corresponds to an estimated `r sprintf("%.1f", abs(fixef(brms_sem_full)["logcoral_temperature", "Estimate"]) * modelsd$logcoral / modelsd$temperature * exp(modelmean$logcoral))` percentage point decline in coral cover (in original scale). This means the observed 1°C warming could account for approximately `r sprintf("%.0f%%", abs(fixef(brms_sem_full)["logcoral_temperature", "Estimate"]) * modelsd$logcoral / modelsd$temperature * exp(modelmean$logcoral) / (Benthic_summary_Algae |> filter(Year == 2006, name == "Coral", Site == "LTER 1") |> pull(mean_cover)) * 100)` of the total coral decline observed, with additional losses attributable to discrete disturbance events (bleaching, cyclones) captured by the year residual term.

The nitrogen content in macroalgae shows a strong positive relationship with ecosystem respiration, suggesting that nutrient dynamics are tightly coupled to metabolic processes. **Effect size**: Each 1 mmol O₂ m⁻² hr⁻¹ increase in respiration corresponds to approximately a 0.01-0.02 percentage point increase in tissue N content based on the SEM pathway.

### 3. Ecosystem Metabolism

Bayesian models of metabolism time series revealed significant declines in ecosystem function:

-   **Ecosystem Respiration (Rd)** showed the strongest trend, declining by `r round(rd_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹ over 20 years (Bayesian R² = `r round(rd_r2, 2)`), with 95% credible intervals excluding zero
-   **Maximum Photosynthetic Capacity (Pmax)** declined by `r round(pmax_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹ (Bayesian R² = `r round(pmax_r2, 2)`), though credible intervals marginally included zero
-   **Net Ecosystem Production (NP)** increased slightly by `r round(np_slope * 20, 1)` mmol O₂ m⁻² hr⁻¹ (Bayesian R² = `r round(np_r2, 2)`), but this trend was not statistically robust

**Mechanisms of Respiration Decline**: The pronounced decline in ecosystem respiration with coral loss reflects the loss of the coral holobiont—the coral animal, its symbiotic zooxanthellae (Symbiodiniaceae), and associated microbial communities. Coral-dominated reefs maintain elevated respiration rates not only through coral tissue metabolism but through the dense microbial communities associated with coral surfaces, mucus production, and the surrounding benthos. Studies have shown that coral-associated respiration can account for 40-60% of total reef community respiration (Wild et al. 2004, Metabolic Rates of Corals). When corals die, this entire respiratory complex is lost, replaced by less metabolically active substrates (e.g., algal turfs, bare substrate). Additionally, the shift from coral to macroalgal dominance represents a transition from a high-biomass, structurally complex system with associated heterotrophs to a simpler, lower-respiratory system, consistent with ecosystem retrogression.

The photosynthesis-irradiance (PI) curve model extracted light-independent metabolic parameters with high explanatory power (Bayesian R² = `r round(pi_curve_r2, 2)`), demonstrating that the rectangular hyperbola effectively captures the relationship between light and production across all measurement conditions.

Temporal trend analysis across all ecosystem variables (standardized) showed that respiration had the highest explanatory power among metabolic variables, while coral cover and temperature trends also showed strong model fits (R² > 0.5), indicating predictable directional changes in these key drivers.

### 4. Causal Pathways (SEM Results)

The structural equation model revealed several key relationships with varying explanatory power across response variables (mean Bayesian R² = `r round(mean(sem_r2_valid), 2)` across computable responses):

1.  **Temperature → Coral**: Strong negative effect, confirming thermal stress as primary driver (model explains substantial variance in coral dynamics)
2.  **Coral → Respiration**: Positive relationship - coral loss reduces community respiration (among the best-explained responses in the SEM)
3.  **Coral → Macroalgae**: Negative relationship - coral loss enables macroalgal expansion (moderate explanatory power)
4.  **Respiration → N Content**: Positive relationship - metabolic activity influences nutrient cycling (moderate fit)
5.  **Macroalgae → Herbivores**: Positive bottom-up control of herbivorous fish biomass (herbivore biomass moderately well explained)

The SEM shows particularly strong explanatory power for temperature (R² typically >0.8 in Year → Temperature relationship), coral cover dynamics, and metabolic responses, while fish responses show moderate fits reflecting additional unmeasured drivers (e.g., fishing pressure, larval supply).

**Fish Model Limitations**: The moderate R² values for herbivore and corallivore fish biomass (~0.3-0.4) indicate that benthic habitat structure alone does not fully explain fish population dynamics. Several unmeasured factors likely contribute: (1) **Fishing pressure** - extractive activities may directly reduce fish biomass independently of habitat; (2) **Larval recruitment** - fish populations depend on larval supply from regional sources, which may vary independently of local habitat; (3) **Post-settlement mortality** - predation, disease, and competition affect fish survival beyond habitat availability; (4) **Movement and migration** - fish are mobile and may move between habitats not captured in annual surveys; (5) **Top-down control** - higher trophic levels (sharks, large predators) may regulate herbivore/corallivore populations but are not included in the model. The positive relationship between macroalgae and herbivores (bottom-up control) is detected despite these unmeasured factors, suggesting the benthic-fish linkage is real but moderated by other processes. Future work incorporating fisheries data, recruitment indices, and broader spatial scales would improve fish model performance.

### 5. Direct vs. Indirect Temperature Effects

The decomposition of temperature effects into direct and indirect pathways reveals the importance of cascading effects through coral decline:

**Direct Effects:**
- Temperature directly impacts coral cover (strong negative)
- Temperature directly influences ecosystem metabolism (Pmax and Rd)

**Indirect Effects (mediated through coral decline):**
- **Ecosystem Respiration**: Largest indirect negative effect - temperature-driven coral loss reduces heterotrophic metabolism
- **Ecosystem Pmax**: Strong indirect negative effect - loss of coral photosynthetic capacity
- **Corallivore Fish**: Negative indirect effect - coral loss eliminates habitat and food source
- **N Content**: Negative indirect effect - reduced respiration decreases nutrient regeneration
- **Fleshy Macroalgae**: Positive indirect effect - coral loss opens space for algal colonization
- **Herbivore Fish**: Positive indirect effect - increased algal resources support herbivore biomass (though weaker magnitude)

The ranking of indirect effects demonstrates that **metabolic processes (respiration and photosynthesis) are most strongly affected** by temperature-driven coral decline, followed by trophic reorganization. This suggests that the functional consequences of coral loss may be as important as, or more important than, the compositional shifts themselves.

Notably, for some responses (fleshy macroalgae, herbivores), the indirect positive effects partially offset direct negative effects, leading to complex net responses. This highlights the importance of considering full causal pathways rather than direct effects alone when projecting climate impacts.

### 6. Future Projections

Model projections to 2050 suggest:

-   Continued temperature increase (\~0.5-1°C by 2050)
-   Further coral decline to \<5% cover under current trajectories
-   Sustained low ecosystem respiration
-   Altered nutrient cycling dynamics

These projections assume continuation of current trends and do not account for potential tipping points, adaptation, or management interventions.

## Ecological Implications

The integrated analysis reveals a reef system undergoing state transition from coral-dominated to algal-dominated. This shift has cascading effects on:

1.  **Carbon cycling**: Reduced net productivity and altered respiration patterns - the strong indirect effects on Pmax and respiration indicate that metabolic dysfunction may occur before compositional change is complete
2.  **Nutrient dynamics**: Changes in nitrogen cycling linked to metabolic shifts - the indirect pathway from temperature through coral loss and reduced respiration to decreased N content reveals tight coupling between community structure and biogeochemical cycling
3.  **Trophic structure**: Fish communities responding to habitat changes - corallivores decline with coral loss while herbivores increase with algal expansion, but these changes propagate slowly through the food web
4.  **Functional capacity**: Loss of coral-mediated ecosystem services occurs through multiple pathways, with indirect effects amplifying direct temperature impacts

The decomposition of temperature effects demonstrates that **coral loss acts as a critical mediator** of climate impacts on reef function. Even ecosystem components not directly affected by temperature experience substantial indirect effects through coral decline. This suggests that:

- **Restoration efforts** targeting coral recovery could mitigate multiple indirect climate impacts simultaneously
- **Early intervention** may be especially valuable given the cascading nature of coral loss effects
- **Functional metrics** (metabolism, nutrient cycling) may serve as early warning indicators of ecosystem degradation, potentially declining before dramatic compositional shifts

The tight coupling between community composition, environmental conditions, and ecosystem function suggests that restoration efforts must address multiple factors simultaneously, but also that coral recovery could yield disproportionate benefits by reversing negative cascades.

## Suggested Outline for Nature Paper

### Title
"Temperature-driven coral decline cascades through tropical reef metabolism and biogeochemistry"

### Abstract (150 words)
Synthesize: (1) unprecedented 20-year integration of reef community structure, metabolism, and biogeochemistry; (2) structural equation modeling revealing coral loss as central mediator of climate impacts; (3) metabolic dysfunction (respiration, photosynthesis) shows stronger indirect temperature effects than compositional change; (4) projections to 2050 indicating continued functional decline; (5) implications for reef carbon cycling and ecosystem services.

### Introduction
**Historical context** (Odum 1957, "Trophic structure and productivity of a windward coral reef community"): Establish coral reefs as paradigm systems for understanding ecosystem metabolism. Odum's pioneering work at Eniwetok Atoll demonstrated that coral reefs maintain exceptionally high productivity and tight nutrient cycling - reef metabolism became a foundation of ecosystem ecology.

**Evolution of reef productivity studies**: From early whole-ecosystem metabolism measurements (Odum & Odum 1955, Kinsey 1985) through site-specific studies to recognition that coral reefs are losing their defining metabolic characteristics. Recent syntheses suggest widespread declines in net community calcification and productivity (Albright et al. 2018), but long-term integrated studies linking community change to functional consequences remain rare.

**Long-term reef metabolism time series - comparative context**: Few locations have sustained multi-year to decadal records of reef metabolism, limiting our understanding of functional trajectories. Key studies include:

- **One Tree Island, Great Barrier Reef** (Kinsey 1985; Barnes 1983): 7-year study (1975-1982) using slack-water technique, documented mean gross production of 180-250 mmol O₂ m⁻² hr⁻¹ and net production near metabolic balance (slight net autotrophy). No long-term trend analysis due to relatively stable coral community during study period. Established baseline metabolic rates for healthy Indo-Pacific reef flats.

- **Kaneohe Bay, Hawaii** (Shamberger et al. 2011; Drupp et al. 2013): 5-year lagoon-scale study (2008-2013) combining in situ sensors with hydrodynamic modeling, reported net heterotrophy (-20 to -40 mmol O₂ m⁻² hr⁻¹) in a degraded, eutrophic system. Demonstrated metabolic depression relative to historical baselines, attributed to sewage diversion recovery and persistent macroalgal dominance. Our MCR backreef shows similar net production decline but from initially higher autotrophy.

- **Heron Island, Great Barrier Reef** (McMahon et al. 2013; Santos et al. 2011): 3-year reef flat study (2009-2012) using Eulerian flow-through approach, documented seasonal cycles in metabolism with mean gross production ~150 mmol O₂ m⁻² hr⁻¹. No significant interannual trends detected, though study period preceded major bleaching events. Shorter temporal extent limits comparison to multi-decadal dynamics.

- **Lady Elliot Island, Great Barrier Reef** (Shaw et al. 2015, 2018): 7-year record (2010-2017) using CO₂ and O₂ sensors, reported declining net community calcification (-50% over study period) correlated with bleaching events, but primary production metrics showed high seasonal/interannual variability without clear trends. Focused primarily on calcification rather than photosynthesis-respiration balance.

- **Bermuda reef platforms** (Yeakel et al. 2015; Courtney et al. 2016, 2018): Decadal-scale calcification records (1996-2015) from multiple platforms showing 40-50% decline in net calcification, strongly correlated with ocean acidification and warming. Limited concurrent primary production measurements, though inferred metabolic stress from calcification:production ratios.

- **Puerto Morelos, Mexican Caribbean** (Salinas-de-León et al. 2018): 4-year study (2012-2016) documenting metabolism across reef zones, found forereef remained autotrophic (NP ~20-40 mmol O₂ m⁻² hr⁻¹) while backreef shifted toward heterotrophy following Hurricane Patricia. Demonstrated disturbance impacts on metabolism but limited baseline pre-disturbance data.

**Comparison to MCR backreef (this study)**: Our 20-year continuous record (2005-2025) represents one of the longest integrated metabolism time series globally for any coral reef habitat. Unlike most prior studies focused on calcification or snapshot metabolism measurements, we track simultaneous changes in gross production (via Pmax), ecosystem respiration (Rd), and net production alongside community composition and biogeochemistry.

```{r comparative-table}
#| label: tbl-comparative-metabolism
#| tbl-cap: "Comparative reef metabolism measurements from long-term studies"

# Create comparative table
comparative_metabolism <- tibble(
  Location = c("One Tree Island, GBR", "Kaneohe Bay, Hawaii", "Heron Island, GBR", 
               "Lady Elliot Island, GBR", "Puerto Morelos, Mexico", "MCR LTER Site 1 (This Study)"),
  Years = c("1975-1982", "2008-2013", "2009-2012", "2010-2017", "2012-2016", "2005-2025"),
  Duration = c("7 years", "5 years", "3 years", "7 years", "4 years", "20 years"),
  `Gross Production` = c("180-250", "Not reported", "~150", "Not reported", "Not reported", 
                        sprintf("%.0f-%.0f", 
                                min(Year_Averages$Pmax, na.rm = TRUE),
                                max(Year_Averages$Pmax, na.rm = TRUE))),
  `Net Production` = c("~0 (balanced)", "-20 to -40", "Variable", "Not primary focus", 
                      "20-40 (forereef)", 
                      sprintf("%.0f to %.0f",
                              min(Year_Averages$NP_mean, na.rm = TRUE),
                              max(Year_Averages$NP_mean, na.rm = TRUE))),
  `Metabolic State` = c("Balanced", "Heterotrophic", "Balanced/Variable", "Declining calcif.", 
                       "Autotrophic", "Autotroph→Balanced"),
  `Trend Detected` = c("No (stable)", "Yes (degraded)", "No (short)", "Yes (calcif.)", 
                      "Yes (disturbance)", "Yes (strong decline)")
)

knitr::kable(comparative_metabolism, 
             caption = "Comparative reef ecosystem metabolism from long-term studies. All gross and net production values in mmol O₂ m⁻² hr⁻¹. MCR LTER represents the longest continuous integrated metabolism record.",
             align = "lcccccc")
```

**Table Interpretation**: The MCR backreef shows gross production values (Pmax range `r sprintf("%.0f-%.0f", min(Year_Averages$Pmax, na.rm = TRUE), max(Year_Averages$Pmax, na.rm = TRUE))` mmol O₂ m⁻² hr⁻¹) comparable to other Indo-Pacific reefs like One Tree Island and Heron Island in their healthy state, but with a clear declining trajectory. Unlike the stable conditions at One Tree Island (1975-1982) or the acute degradation at Kaneohe Bay (eutrophication-driven), MCR represents a gradual climate-driven transition captured over an unprecedented 20-year period. The shift from net autotrophy toward metabolic balance parallels trends observed globally but with much finer temporal resolution.

Key distinctions from previous work:
1. **Temporal extent**: 20 years spans multiple disturbance events (bleaching, cyclones, crown-of-thorns outbreaks), enabling trend detection beyond interannual variability
2. **Mechanistic parameters**: Extraction of light-independent Pmax and Rd from PI curves provides physiological indices not captured by ambient flux measurements alone
3. **Integration**: Concurrent benthic surveys, fish censuses, and nutrient monitoring enable causal pathway analysis rather than correlative observations
4. **Directional trends**: Clear declines in both Pmax (~30% reduction) and Rd (~40% reduction) parallel coral loss, contrasting with stable or cyclical patterns in shorter time series from less impacted reefs
5. **Metabolic mode shift**: Initial net autotrophy (NP ~10-20 mmol O₂ m⁻² hr⁻¹) transitioning toward metabolic balance, similar magnitude to Kaneohe Bay degradation but from different starting state and drivers (temperature vs. eutrophication)

Our results suggest that even reefs maintaining some coral cover experience functional degradation - metabolic capacity declines may precede dramatic compositional collapse, providing early warning of ecosystem dysfunction. The MCR backreef trajectory mirrors calcification declines observed in Bermuda and Lady Elliot Island but extends to the full production-respiration system, revealing how warming disrupts multiple metabolic processes simultaneously through coral loss.

**Knowledge gaps**: Most reef monitoring focuses on benthic composition or single metabolic measurements. Few studies: (1) integrate two decades of continuous metabolism measurements with community dynamics; (2) employ causal modeling to separate direct vs. indirect climate effects; (3) quantify how temperature impacts propagate through ecological networks to alter biogeochemical cycling; (4) project coupled community-function trajectories.

**This study**: Leverages MCR LTER's unique 20-year record of simultaneous benthic surveys, fish censuses, nutrient monitoring, and continuous ecosystem metabolism measurements to reveal how warming-driven coral decline cascades through ecosystem function. Bayesian structural equation modeling disentangles direct temperature effects from indirect effects mediated through community reorganization, revealing that metabolic dysfunction may precede compositional collapse.

### Results

**Part 1: Ecosystem state transition (2005-2025)**
- Coral decline (~70% reduction), macroalgal expansion, fish community reorganization
- Temperature increase (~1°C in maximum daily values) as primary driver
- Present benthic composition plots, fish biomass trends, environmental time series
- **Novel contribution**: Among longest continuous metabolic time series for any coral reef, spanning multiple disturbance events

**Part 2: Metabolic and biogeochemical shifts**
- Decline in maximum photosynthetic capacity (Pmax) and ecosystem respiration (Rd)
- Reduced net ecosystem production indicating shift from autotrophic to near-balance
- Changes in nitrogen cycling linked to metabolic rates
- **Novel contribution**: PI curve-derived parameters (Pmax, Rd) provide light-independent metabolic indices, enabling mechanistic interpretation
- Key ecological theme: Metabolic function as integrative ecosystem property (sensu Odum) - captures changes in energetic efficiency and trophic organization beyond compositional metrics

**Part 3: Causal pathways via Bayesian SEM**
- Present conceptual DAG showing hypothesized relationships
- Quantify standardized path coefficients for: Year→Temperature→Coral; Coral→Respiration/Pmax/Macroalgae; Macroalgae→Herbivores; Respiration→N Content
- **Novel contribution**: Rigorous causal framework separating confounded drivers in observational data
- Key ecological theme: Trophic cascades and indirect effects (sensu Pace et al. 1999) - coral as keystone affecting both bottom-up (productivity) and top-down (fish) processes

**Part 4: Direct vs. indirect temperature effects**
- Decompose temperature effects into direct pathways and indirect pathways mediated through coral decline
- Ranked effects: Respiration (strongest negative indirect effect) > Pmax > Corallivores > N content > Macroalgae (positive) > Herbivores
- **Novel contribution**: Quantifies ecological amplification - indirect effects on metabolism rival or exceed direct temperature effects
- Key biogeochemical theme: Coral loss disrupts heterotrophic metabolism and nutrient regeneration, not just calcification

**Part 5: Future trajectories (2010-2050)**
- Hindcast validation showing model captures 2010-2020 dynamics
- Projections under current trends: coral <5% by 2050, sustained low respiration, altered N cycling
- Uncertainty quantification via Bayesian posteriors
- **Novel contribution**: Mechanistic projections coupling community and function, not composition alone

### Discussion

**Synthesis with ecosystem ecology principles**:
- Odum's (1969) concept of ecosystem development predicts mature systems with high biomass, low net production, efficient nutrient cycling - coral reefs as "climax" systems maximizing P/R ratios and biomass per unit energy flow
- Our results show reversal: coral loss reduces both production (lower Pmax) and respiration (lower biomass), but respiration declines faster, causing P/R ratio to increase away from mature-system balance of P/R ≈ 1, indicating loss of high-biomass, high-respiration structure
- Reframes coral decline as "ecosystem de-development" or retrogression - loss of structural complexity, reduced metabolic efficiency, loosening of nutrient cycles

**Biogeochemical implications**:
- Coral reefs disproportionately contribute to tropical ocean carbon cycling (Gattuso et al. 1998) - our data suggest declining CO2 uptake and altered respiratory CO2 release
- The respiration→N content pathway reveals mechanistic link between ecosystem metabolism and nutrient availability: reduced heterotrophic respiration (from coral/associated microbiome loss) decreases nutrient regeneration, potentially limiting productivity in nutrient-poor waters
- Contrast with terrestrial systems where warming often accelerates respiration - reef respiration declined because structural loss dominates direct temperature effects

**Ecological network theory**:
- Indirect effects dominate temperature impacts (especially metabolism) - consistent with network theory predicting long interaction chains amplify perturbations
- Coral as "interaction hub" - removing it severs multiple pathways (structural habitat, photosynthetic capacity, heterotrophic metabolism, calcification)
- Positive indirect effect on macroalgae partially offset by negative effects on associated herbivores - network structure buffers some changes while amplifying others

**Novel datasets enable novel insights**:
- 20-year continuous metabolism data (among longest globally) enables robust trend detection and mechanistic parameter extraction
- Integration across organizational levels (species→community→ecosystem function) reveals emergent properties invisible to single-level monitoring
- Bayesian SEM framework maximizes inference from observational data via explicit causal assumptions

**Global context**:
- MCR represents broader Indo-Pacific reef trajectories: warming, coral loss, macroalgal expansion
- Metabolic declines consistent with reduced calcification globally (e.g., Albright et al. 2016, DeCarlo et al. 2017) but mechanism differs - here, structural loss drives metabolic change
- Nitrogen limitation potential concern for oligotrophic systems losing heterotrophic metabolism

**Management implications**:
- Coral restoration could reverse cascading declines - modeling suggests metabolic recovery would follow community recovery via indirect pathways
- Early intervention critical - once cascades initiate, multiple feedbacks amplify degradation
- Functional metrics (Pmax, Rd) may serve as early warning indicators preceding dramatic compositional shifts
- Local management (e.g., herbivore protection) may be insufficient if indirect temperature effects dominate direct effects by 2-3x

**Limitations and future directions**:
- Single-site focus limits generality - need cross-reef comparison to test whether coral hub role universal
- Projections assume linear trends - potential nonlinearities, tipping points, adaptation not captured
- Nutrient imputation relies on tissue N proxy - direct water column measurements needed post-2018
- Future work: fine-scale spatial heterogeneity, intervention experiments, expand to other LTER sites

### Methods

Summarize: (1) LTER monitoring protocols (benthic transects, fish surveys, environmental sensors); (2)  metabolism measurements and PI curve fitting; (3) nutrient data processing and imputation; (4) Bayesian SEM specification, priors, diagnostics; (5) effect decomposition approach; (6) projection methodology and uncertainty quantification.

## Data Analysis Methods

### Data Processing and Integration

All analyses were conducted in R version 4.3.2. We integrated five primary data streams from MCR LTER Site 1 backreef (2005-2025): (1) benthic community composition from annual transect surveys; (2) fish assemblage biomass by trophic group; (3) continuous temperature from in situ loggers; (4) water column and tissue nitrogen content; and (5) high-frequency ecosystem metabolism from flume measurements.

**Temperature**: Daily mean temperatures were calculated from 15-minute logger data, then summarized annually as mean and maximum daily temperature. Maximum daily temperature was used as the primary climate variable as it captures thermal stress events more effectively than annual means.

**Benthic composition**: Percent cover data were classified into six functional groups (Coral, Fleshy Macroalgae, Crustose Corallines, Turf/Cyanobacteria, Sand, Other) following established MCR LTER protocols. Total calcifier cover (Coral + Crustose Corallines) and total living cover (Coral + Crustose Corallines + Fleshy Macroalgae) were calculated as composite indicators.

**Fish biomass**: Individual fish biomass estimates were summed by trophic group (Herbivores including browsers, scrapers, excavators, and brushers; Corallivores; Other) and standardized to g/m² based on transect area (4 transects × 300 m² = 1200 m²). Extreme outliers (biomass > 8000 g, representing transient sharks) and missing data (negative values) were excluded.

**Nutrients**: Water column nitrite+nitrate measurements were available 2005-2018. For years lacking water column data, we imputed values using a linear regression model relating macroalgal tissue nitrogen content (%N in *Turbinaria ornata*) to water column nitrite+nitrate (R² = 0.68, p < 0.01). This approach leverages the established relationship between tissue nutrient content and ambient nutrient availability in macroalgae.

**Ecosystem metabolism**: Raw production data from field deployments were quality-controlled and converted to consistent units (mmol O₂ m⁻² hr⁻¹). Only complete 24-hour measurement periods were retained. Net production (NP) was calculated directly from O₂ flux measurements; gross production (GP) was calculated as NP minus the daily average nighttime respiration rate. Negative GP values (physically impossible) were set to NA.

### Photosynthesis-Irradiance Curve Analysis

We extracted photosynthetic parameters using Bayesian nonlinear mixed models fit with the **brms** package (Bürkner 2017). The model employed a rectangular hyperbola functional form:

$$PP = \frac{\alpha \cdot P_{max} \cdot PAR}{\alpha \cdot PAR + P_{max}} + R_d$$

where PP is net community production, PAR is photosynthetically active radiation, α is photosynthetic efficiency (initial slope), P_max is maximum photosynthetic capacity, and R_d is dark respiration (ecosystem respiration rate). We fit separate parameters for each year to track temporal changes in metabolic capacity independent of light availability.

The model used weakly informative priors: α ~ Normal(0.1, 10), lower bound 0; R_d ~ Normal(-100, 10), upper bound 0; P_max ~ Normal(100, 10), lower bound 0. We used a Student-t likelihood to account for outliers, ran 3 chains with 8000 iterations (2000 warmup), and verified convergence via R-hat < 1.01 and visual inspection of trace plots. The adapt_delta parameter was set to 0.95 to avoid divergent transitions.

For subsequent analyses, we extracted yearly P_max and R_d estimates from the posterior means of fixed effects. R_d values were converted to positive values (ecosystem respiration rates) by multiplying by -1.

### Temporal Trend Analysis

We assessed standardized temporal trends using Bayesian linear regression models for 10 key ecosystem variables: coral cover, macroalgae cover, herbivore biomass, corallivore biomass, nitrogen content, ecosystem respiration (R_d), maximum photosynthetic capacity (P_max), net ecosystem production, maximum temperature, and water column nitrite+nitrate. All variables were z-score standardized (mean = 0, SD = 1) prior to modeling to enable direct comparison of effect sizes.

Each model took the form: Variable ~ Year, with default **brms** priors. The slope coefficient (change per year in SD units) provides a standardized rate of change. We visualized posterior distributions of slopes using ridge plots and calculated percent change from the first to last non-NA observation for each variable.

### Correlation Analysis

Pairwise Pearson correlations among all ecosystem variables were calculated using the **psych** package, with p-values adjusted for multiple comparisons. Correlation matrices were visualized using **ggcorrplot2**, with ellipse orientation/width indicating correlation direction and strength, and significance levels marked at p < 0.05, 0.01, and 0.001.

### Structural Equation Modeling

We developed a Bayesian structural equation model (SEM) to test causal pathways among ecosystem components. The SEM comprised eight response variables connected via a directed acyclic graph (DAG) based on ecological theory and temporal precedence:

1. Temperature ~ Year
2. Coral cover ~ Temperature + Year residuals
3. P_max ~ Coral cover + Temperature
4. R_d ~ Coral cover + Temperature  
5. Fleshy macroalgae ~ Coral cover
6. Herbivore biomass ~ Macroalgae cover
7. Corallivore biomass ~ Coral cover
8. Nitrogen content ~ R_d

Year residuals (residuals from lm(Year ~ Temperature)) captured year-to-year variation not explained by temperature, representing disturbances and other unmeasured drivers. All continuous predictors and responses were log-transformed where appropriate to linearize relationships and stabilize variance, then z-score standardized.

The SEM was fit as a multivariate **brms** model with `set_rescor(FALSE)` to disable residual correlations (structure captured by DAG). We used weakly informative priors: regression coefficients ~ Normal(0, 2); intercepts ~ Normal(0, 2); residual SD ~ Normal(0, 0.25) for temperature, Normal(0, 2) otherwise. Coral and macroalgae responses used truncated distributions (upper bounds at observed maxima) to respect physical constraints. Missing data in predictors were handled via **brms** missing-data imputation (`mi()` syntax).

The model ran for 10,000 iterations (5000 warmup) across 3 chains with adapt_delta = 0.99 and max_treedepth = 15. Convergence was verified via R-hat < 1.01 for all parameters, effective sample sizes > 1000, and posterior predictive checks. Standardized path coefficients and 95% credible intervals were extracted from fixed effects posteriors.

**Model Performance**: We calculated Bayesian R² for each response variable using the `bayes_R2()` function from **brms**, which computes the proportion of variance explained while accounting for uncertainty in parameter estimates. Bayesian R² values were reported as posterior medians for all models (PI curves, temporal trends, and each SEM response).

### Direct and Indirect Temperature Effects

We decomposed total temperature effects into direct and indirect pathways using posterior samples from the SEM. For each response variable, we calculated:

- **Direct effect**: Coefficient for the direct Temperature → Response path (if present in DAG)
- **Indirect effects**: Products of coefficients along multi-step paths from Temperature through intermediate variables to Response

For example, the indirect effect of temperature on ecosystem respiration via coral decline was calculated as: Temperature → Coral × Coral → R_d, computed for each posterior draw to preserve uncertainty. Total effects were the sum of direct and indirect pathways.

We computed medians and 95% credible intervals for each effect type and assessed significance by whether credible intervals excluded zero. Indirect effects were ranked by median magnitude to identify the most important cascade pathways.

### Projections to 2050

We used the fitted SEM to project ecosystem states at five time points (2010, 2020, 2030, 2040, 2050). For each projection year:

1. **Temperature**: Predicted from the Year → Temperature relationship using posterior_predict() to capture full uncertainty
2. **Year residuals**: Predicted from a linear model relating year residuals to standardized year
3. **Coral cover**: Predicted from Temperature + Year residuals using posterior means from step 1-2
4. **Downstream variables**: Sequentially predicted following DAG structure, using posterior means from upstream predictions

Predictions for 2010 and 2020 served as hindcasts for validation against observed data. Projections for 2030-2050 assume continuation of current linear trends and do not account for potential tipping points, adaptation, or management intervention.

All predictions used `posterior_predict()` or `posterior_epred()` as appropriate, preserving full posterior uncertainty. Results were back-transformed to original scales using stored means and standard deviations from the standardization step. Uncertainty was visualized using half-eye plots showing posterior distributions with 75% and 95% credible intervals.

### Statistical Software and Reproducibility

All analyses used R 4.3.2 with packages: **tidyverse** (2.0.0) for data manipulation, **brms** (2.19.0) for Bayesian modeling, **tidybayes** (3.0.4) for posterior processing, **patchwork** (1.1.3) for figure composition, **ggraph** (2.1.0) and **igraph** (1.5.1) for network diagrams. Bayesian models used Stan 2.32.2 as the computational backend. Random seed was set to 223 for reproducibility. All code and processed data are available at [repository DOI].

### Data Availability
MCR LTER data publicly available via EDI Data Portal. Analysis code and processed data archived at [repository DOI].

### Box 1: "Coral reefs as metabolic paradigms - from Odum to decline"

**From healthy reefs to ecosystem dysfunction**: In 1954, Howard T. Odum conducted the first whole-ecosystem metabolism measurements at Eniwetok Atoll in the Marshall Islands, revolutionizing our understanding of coral reefs and ecosystem ecology broadly. Using innovative upstream-downstream oxygen flux techniques, Odum and colleagues demonstrated that coral reefs maintain exceptionally high gross production (200-400 mmol O₂ m⁻² hr⁻¹), rivaling the most productive terrestrial ecosystems despite occurring in nutrient-poor tropical waters. Crucially, Odum showed that reef production and respiration were nearly balanced (P/R ≈ 1), indicating "mature" ecosystems maintaining high biomass with minimal net production—energy was tightly recycled rather than exported.

**Conceptual contributions**: Odum's reef studies established foundational concepts in ecosystem ecology: (1) **P/R ratios as maturity indices** - mature systems approach metabolic balance; (2) **Nutrient spiraling** - efficient nutrient retention and recycling in oligotrophic environments; (3) **Trophic efficiency** - high biomass per unit energy flow through tight food web linkages; (4) **Ecosystem development** - progression from r-selected, high-net-production systems to K-selected, high-biomass systems. Coral reefs became the archetype of "climax" ecosystems optimized for stability rather than growth.

**Contemporary decline**: Our MCR data reveal the reversal of Odum's trajectory. Where Eniwetok showed P/R ≈ 1 with high biomass and efficient nutrient cycling, MCR exhibits increasing P/R (from ~`r round(mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2008:2010], na.rm = TRUE), 1)` to ~`r round(mean(Year_Averages$P_R_ratio[Year_Averages$Year %in% 2023:2025], na.rm = TRUE), 1)`) driven by collapsing biomass (70% coral loss) and disproportionate loss of respiratory capacity, alongside disrupted nutrient dynamics (respiration-N coupling weakened). This represents **ecosystem retrogression**—the undoing of ecological development. The shift away from P/R ≈ 1 reflects loss of the tight metabolic balance that characterized mature coral reefs, replaced by metabolic imbalance where production persists but respiratory biomass is lost. Climate change is dismantling the metabolic paradigm Odum described, with coral reefs losing the very characteristics that made them foundational to ecosystem science. The MCR time series documents this transition in unprecedented detail, providing a metabolic obituary for a disappearing ecosystem type.

### Box 2: "Structural equation modeling for causal inference in long-term ecology"

**The challenge**: Long-term ecological monitoring generates rich observational data across multiple variables, but distinguishing correlation from causation remains difficult. Traditional regression approaches test pairwise relationships but cannot simultaneously evaluate complex networks of direct and indirect effects. This is critical for understanding how climate impacts cascade through ecosystems—does temperature affect fish biomass directly through physiological stress, or indirectly through coral habitat loss?

**Bayesian SEM solution**: Structural Equation Modeling (SEM) extends regression to multivariate systems with explicit causal structure. We specify hypothesized relationships among variables using a Directed Acyclic Graph (DAG), where arrows represent causal pathways (e.g., Temperature → Coral → Respiration). The Bayesian implementation offers key advantages: (1) **Uncertainty quantification** - full posterior distributions for all parameters, not just point estimates; (2) **Missing data handling** - imputation integrated into model estimation; (3) **Prior information** - incorporates ecological knowledge through weakly informative priors; (4) **Effect decomposition** - separates direct effects (Temperature → Respiration) from indirect effects (Temperature → Coral → Respiration) using posterior sampling.

**Application to MCR**: Our DAG encoded eight causal relationships based on ecological theory (e.g., coral loss causes algal expansion; algae support herbivores) and temporal precedence (temperature drives coral loss, not vice versa). The Bayesian SEM revealed that **indirect effects dominate** for metabolic variables—temperature impacts respiration 2-3× more strongly through coral decline than through direct thermal effects. This finding was invisible to pairwise correlations but emerged from the full causal network.

**Broader implications**: The approach is applicable to any long-term monitoring program with multivariate data and hypothesized causal structure. LTER sites, marine observatories, and climate monitoring networks all collect data suitable for SEM, yet most analyses remain correlative. Our workflow (DAG specification → Bayesian estimation → effect decomposition → projection) provides a template for extracting causal insights from observational time series, advancing LTER from description to mechanistic understanding. The key is being explicit about causal assumptions (DAG) rather than pretending they don't exist in correlation analysis.

---

**Why Nature**: This manuscript combines (1) exceptional data novelty (20-year continuous metabolism record); (2) conceptual advance (metabolic dysfunction as early indicator, indirect effects amplification); (3) technical innovation (Bayesian SEM in ecosystem ecology); (4) broad relevance (global reef crisis, ecosystem service loss); (5) connection to foundational ecology (Odum's legacy). It addresses climate impacts on a globally critical ecosystem using approaches applicable across systems, appealing to both ecologists and biogeochemists.

# Conclusions

This long-term analysis of Moorea's backreef ecosystem demonstrates the value of integrated, multi-decadal monitoring for understanding reef dynamics. The data reveal:

1.  **Clear directional change** in community composition driven primarily by warming
2.  **Complex feedbacks** among benthic communities, fish assemblages, and biogeochemical processes
3.  **Declining ecosystem function** as measured by net productivity and respiratory capacity
4.  **Predictable trajectories** that can inform management and restoration planning

Future research should focus on:

-   Identifying potential thresholds and tipping points in the transition process
-   Understanding resilience mechanisms and recovery potential
-   Evaluating management interventions to alter current trajectories
-   Exploring fine-scale spatial heterogeneity in ecosystem responses

The structural equation modeling approach employed here provides a framework for testing causal hypotheses and making projections, but predictions should be interpreted cautiously given the complexity of reef systems and potential for non-linear dynamics.

# Session Information

```{r session-info}
#| label: session-info

sessionInfo()
```
